### 2022-08-20 surv.Rmd
initial

---
title: "Untitled"
author: "Dean Marchiori"
date: "8/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(survival)
library(survminer)
library(SMPD4)
library(dplyr)
```

```{r}
nedmaba <- smpd4_phenotype |> 
  select(id, survival_time, deceased, gender) |> 
  mutate(deceased = as.numeric(deceased)) 
```

```{r}
km1 <- survfit(Surv(survival_time, deceased) ~ 1, data = nedmaba)
summary(km1, times= (0:13)*365, scale=365)
ggsurvplot(km1,
          conf.int = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))
```

```{r}
km2 <- survfit(Surv(survival_time, deceased) ~ gender, data = nedmaba)
km2
summary(km2, times= (0:13)*365, scale=365)

ggsurvplot(km2,
          pval = TRUE, conf.int = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))

survdiff(Surv(survival_time, deceased) ~ gender, data = nedmaba)

fit <- coxph(Surv(survival_time, deceased) ~ gender, data = nedmaba)
summary(fit)
```

### 2022-08-27 paper_draft.qmd
push targets code

---
title: "paper_draft"
author: "DM"
format: pdf
editor: visual
bibliography: refs.bib
execute: 
  echo: false 
---

## Title

-   the name of the organism studied

-   the particular aspect or system studied

-   the variable(s) manipulated.

## Abstract

passive voice

(typically 150-300 words)

-   A brief introduction to the topic that you're investigating.

-   Explanation of why the topic is important in your field/s.

-   Statement about what the gap is in the research.

-   Your research question/s / aim/s.

-   An indication of your research methods and approach.

-   Your key message.

-   A summary of your key findings.

-   An explanation of why your findings and key message contribute to the field/s

## Introduction

Neurodevelopment disorders are a diverse and heterogeneous group of conditions impacting the development of the nervous system, brain function, physical development, emotional development and learning ability. A recently identified condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) (MIM:[618622](https://omim.org/entry/618622)) has been described in children with biallelic loss-of-function variants in *SMPD4* [@magini2019loss].

Sphingomyelinases such as *SMPD4* hydrolyze sphingomyelin into ceramide and phosphorylcholine. *SMPD4* specifically encodes one of 4 neutral sphinogmyelinases, nSMase3 (MIM: [610457](http://omim.org/entry/610457)).

So far, only approximately 30 cases have been reported across over 12 families. Clinical phenotype data of these cases is largely heterogeneous with severe neurological complications and early demise a key feature. This makes the identification, diagnosis and ongoing management of cases a challenge for patients, their families and medical practitioners.

@ravenscroft2021neurogenetic performed a study over 190 probands with a diagnosis of arthrogryposis multiplex congenita, distal arthrogryposis, fetal akinesia deformation sequence or multiple pterygium syndrome. This study identified a novel missense variation in SMPD4 which at the time of whole exome sequencing was not well described in the literature. The impacted family from Melbourne exhibited features involving arthrogryposis multiplex congenita, complex brain malformations, small for gestation age and hypoplasia of the corpus callosum. In two of the three related cases were additional features of microcephaly, congenital encephalopathy, cerebellar malformation and hypoplasia and hypomyelination.

@monies2019lessons described two families with different homozygous truncating variants in SMPD4 as a syndrome of skeletal dysplasia with cerebella atrophy. The phenotypes described involved bilateral clenched hands, talipes, IUGR, partial absence of corpus callosum with family history of three neonatal deaths with similar features.

A detailed study by @magini2019loss involved 12 unrelated familes with 32 individuals (21 with detailed clinical information). Hallmark presentations were of microcephaly, simplified gyration, hypomyelination, thin corpus callosum, mild cerebellar hypoplasia, brainstem hypoplasia, congenital arthrogryposis, diabetes mellitus, heart disease, severe encephalopathy and respiratory problems often leading to early demise. Despite this being the largest cohort studied, the clinical features and survival times among participants varied greatly. Three missense changes were noted in the study with affected children in these families often showing a milder presentation suggestive of possible residual function. In these cases individuals were able to develop independent motor skills, have mild intellectual disability and arthrogryposis without evidence of simplified gyral patterns on brain MRI. Other patients with truncating variants are shown to have more severe presentations, while a range of additional significant phenotypes were reported involving dysmorphic facial features, seizure, vocal cord paralysis and hearing impairment.

A recent case study from China is described in @ji2022case involving a girl presenting in infancy with intrauterine growth restriction, microcephaly, postnatal developmental delay, arthrogryposis, hypertonicity, seizure, and hypomyelination on brain magnetic resonance imaging. @ji2022case argues the actual prevalence of this condition, based on the gene carrier rate is understated and may be attributable to typical symptoms of NEDMABA being non-specific, providing a diagnostic challenge in a clinical setting.

A recent study by @bijarnia2022growth presents the case of a 22-month old girl presenting with the typical phenotype of neurodevelopmental delay, prenatal onset growth failure, arthrogryposis, microcephaly and brain anomalies including severe hypomyelination, simplified gyral pattern and hypoplasia of corpus callosum and brainstem. Notably, there is also additional non-typical clinical findings of nystagmus and visual impairment secondary to macular dystrophy and retinal pigment epithelial stippling at posterior pole.

@ji2022case reports parallels with two cases showing the same homozygous null variant. The individual reported in @monies2019lessons presented with distinct symptoms of brain atrophy and skeletal dysplasia whereas the case in @magini2019loss with the same variant exhibited more typical clinical features.

Further work to collate and analyse the current body of research is challenging. While @magini2019loss cataloged a detailed clinical phenotype data set as a supplimentary to their study, these data are not suitable for statisical analysis. Many of the clinical features are represented as free text descriptions and a variety of non-standard terminology is applied between cases, making machine interpretation of the data difficult. Other studies [@ji2022case; @bijarnia2022growth] present only written case reports with some tabulated summaries. While larger studies [@ravenscroft2021neurogenetic; @monies2019lessons] offer data genotype analysis, there is less detail included on the clinical phenotype of indivuduals in the cohort.

Given the severe nature of this condition, often leading to early demise, a gap exists in understanding the expected survival time for individuals with bi-allelic null-variants in SMPD4. While a connection has been highlighted between some missense variants and a milder presentation with longer survival, this hypothesis has not been analysed further. Furthermore, the associations between diverse features of NEDMADA have not been analysed in detail, in-part due to the sparse and inconsistent collection of clinical and case report data for this new condition.

An open research question exists around the clinical pathway and survival time of children exhibiting bi-allelic loss-of-function variants in SMPD4. With current research highlighting a diverse and heterogeneous phenotype, the relationship and correlations between these diverse features is not well understood.

This research has three key aims. First, to collate and transform the variety of early case reports and studies on clinical phenotype data for this novel variant into an analysis-ready *tidy* dataset [@JSSv059i10]. Secondly, to statistically quantify the expected survival time of children with NEDMABA based on the current case reports. Finally, detailed analysis will be conducted to further understand the associations between both patients and their clinical features.

While current research in this area has highlighted a diverse range of presentations, this will be the first in-depth analysis of early studies of this novel variant, which aims to produce statistical findings to better describe and understand this condition. This research is significant as it will assist clinicians understand the typical and non-typical presentations of this challenging new condition. In addition, genetic counselling of families with affected children will benefit from enhanced analysis on outcomes of existing reported cases.

## Methods

### SMPD4 Data Package

The data contained in @magini2019loss "Summary of SMPD4-Related Clinical Phenotype" is an excel spreadsheet tabulating each of the 21 individuals from the study
with clinical details recorded. The file has 21 columns (one for each individual), and 64 rows (one for each phenotype or clinical remark).

The data were read into R [@rbase] without any changes, so as to preserve the reproducibility of the data transformation steps.

The data were transposed to ensure it could be presented as *tidy* formatted data [@JSSv059i10]. This requires:

1. Each variable (clinical phenotype) forms a column.
2. Each observation (individual) forms a row.
3. Each type of observational unit forms a table  (every cell has just one item)

In many cases, key clinical information was entered as free-text descriptions, which rendered any attempt of meaningful analysis impractical (@tbl-nontidy). In these cases, the text was tokenised by separating the list of clinical observations at each comma and forming a binary indicator column noting its presence '1' or absence '0' (@tbl-tidy).

|         | Family 1- Individual 1 | Family 1- Individual 4 |
|---------| -----------------------|------------------------|  
Facial dysmorphisms | short palpebral fissures, large ears, simple helices, smooth philtrum, thin lips, bilateral simian creases | short palpebral fissures, receding forehead, thin upper lip |

: Example of non-tidy free-text descriptions of features {#tbl-nontidy}

| id          | short palpebral fissures | large ears |simple helices | receding forehead | ... | 
| ------------|------|------|------|------|-----|
| Family 1- Individual 1  | 1 |  1 | 0 | ... |
| Family 1- Individual 4  | 1 | 0 | 0 | 1 | ... |

: Example of tidy formatted data where individuals are transposed into rows and text into binary indicators {#tbl-tidy}

In the case where two variables were formed from phenotypes that are considered to be synonymous, these
were merged into one indicator column to prevent duplication e.g. {bilateral_cleft_lips, bilateral_cleft_lip, cleft_lip_b_l}.

Some data type conversion and categorical level standardisation was performed to ensure the data were in consistent and appropriate data types. For example, 'Gender' was not consistently coded, and 'Birth Weight' was encoded as a text string rather than a more useful numeric format. (@tbl-data-types).

| Gender | Birth Weight |
|--------|--------------|
| male   | 2175 grams (- 2.5 SD) |
| female | 2045 g (-3 SD) |
| Female | 2300 gram (-2 SD) |
| Female fetus | n.a. |

: Example of inconsistent coding or sub-optimal data types {#tbl-data-types}

The final dataset consisted of 21 observations (one per individual) and 152 variables (one per clinical feature).

This format was preserved and other case studies identified in the literature [@ravenscroft2021neurogenetic; @monies2019lessons;
@bijarnia2022growth; @ji2022case] were manually entered to conform to this template to allow the data to be combined for further
analysis.

These data sets were packaged into an R Package called SMPD4 [@smpd4-data] in order to allow for reproducibility and sharing. This can 
be downloaded and installed from github at [https://github.com/deanmarchiori/SMPD4](https://github.com/deanmarchiori/SMPD4).

### Multiple Correspondence Analysis

### Cluster Analysis

### Survival Analysis

-  Explain why each procedure was done

-   Experimental procedures and results are narrated in the past tense

-   Mathematical equations and statistical tests are considered mathematical methods and should be described in this section along with the actual experimental work.

## Results

The function of this section is to summarize general trends in the data without comment, bias, or interpretation. The results of statistical tests applied to your data are reported in this section although conclusions about your original hypotheses are saved for the Discussion section.

## Discussion

-   the relationship between the results and the original hypothesis, i.e., whether they support the hypothesis, or cause it to be rejected or modified

-   an integration of your results with those of previous studies in order to arrive at explanations for the observed phenomena

-   possible explanations for unexpected results and observations, phrased as hypotheses that can be tested by realistic experimental procedures, which you should describe

-   End the Discussion with a summary of the principal points you want the reader to remember

@bijarnia2022growth

## References

```{r eval = FALSE}
source('_targets_packages.R')
library(targets)
library(knitr)

```

```{r fig.height=8, fig.width=8, eval=FALSE}
tar_load(subjects)
tar_load(cluster_data)
tar_load(clusters)
tar_load(cluster_assign)

cm <- as.matrix(cluster_data)
# Check the most similar
subjects[which(cm == min(cm[cm != min(cm)]), arr.ind = TRUE)[1, ], ]

# Check the most dissimilar
subjects[which(cm == max(cm[cm != max(cm)]), arr.ind = TRUE)[1, ], ]

summary(clusters)
plot(clusters)
rect.hclust(clusters, k=4, border="red")

tar_load(pca_data)

pca_data |> 
  bind_cols(cluster = cluster_assign) |> 
  pivot_longer(cols = -c(cluster), names_to = 'a', values_to = 'v') |> 
  group_by(cluster, a) |> 
  summarise(n = sum(v)) |>  
  arrange(n) |> 
  ggplot(aes(n, a)) +
  geom_col() +
  facet_wrap(~cluster, scales = "free_x")

tibble(n = 1:nrow(subjects), id = subjects$id, cluster  = cluster_assign) |> kable()

tibble(n = 1:nrow(subjects), 
       d = subjects$id, 
       dec = subjects$deceased,
       st = subjects$survival_time, 
       cluster  = cluster_assign) |> 
  group_by(cluster) |> 
  summarise(ast = mean(st),
            mst = max(st),
            minst = min(st),
            n = n(),
            sd = sum(dec))

```

```{r fig.height=8, fig.width=8, eval=FALSE}
tar_load(survival_data)
tar_load(kapm)
tar_load(cox_ph)
kapm <- survfit(Surv(survival_time, deceased) ~ cluster, data = survival_data)
summary(kapm, times= (0:60)*30, scale=30)
x <- survminer::surv_summary(kapm, data = survival_data)
plot(kapm)
survminer::ggsurvplot_df(x,
                              conf.int = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw())

survival::survdiff(Surv(survival_time, deceased) ~ cluster, data = survival_data)
coxph(Surv(survival_time, deceased) ~ cluster, data = survival_data)
summary(cox_ph)

```

```{r eval=FALSE}
library(FactoMineR)
library(factoextra)

plot.MCA(mca,invisible=c("quali.sup","quanti.sup"),cex=0.7,  autoLab = 'yes', select  = "contrib 5", selectMod = "cos2 5") +
  theme(panel.grid.major = element_blank(),
        plot.title=element_text(size=14, color="blue"),
        axis.title = element_text(size=12, color="red"))

plotellipses(res.mca,keepvar=1:20)

# eigenvalues
fviz_screeplot(res.mca, addlabels = TRUE, ylim = c(0, 45))

# biplot 
fviz_mca_biplot(res.mca, 
                repel = TRUE, # Avoid text overlapping (slow if many point)
                ggtheme = theme_minimal())

# variable plots

fviz_mca_var(res.mca, choice = "var",repel = T,
             ggtheme = theme_minimal())

fviz_mca_var(res.mca, choice = "var.cat", repel = T,
             ggtheme = theme_minimal())

### quality 
fviz_mca_var(res.mca, col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE, # Avoid text overlapping
             ggtheme = theme_minimal())

fviz_mca_var(res.mca, alpha.var="cos2",
             repel = TRUE,
             ggtheme = theme_minimal())

fviz_cos2(res.mca, choice = "var", axes = 1:2)

# indiv
fviz_mca_ind(res.mca, 
             label = "none", # hide individual labels
             habillage = "deceased", # color by groups 
             palette = c("#00AFBB", "#E7B800"),
             addEllipses = TRUE, ellipse.type = "confidence",
             ggtheme = theme_minimal())

fviz_mca_ind(res.mca, col.ind = "cos2", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, # Avoid text overlapping (slow if many points)
             ggtheme = theme_minimal())

res.desc <- dimdesc(res.mca, axes = c(1, 2))

res.desc[[2]]

```

### 2022-08-27 R/prep_cluster_data.R

@@ -1,0 +1,11 @@
+prep_cluster_data <-  function(data,
+                               keep_cols) {
+  clust <- data |>
+    dplyr::select(all_of(keep_cols))  |> 
+    select_if(function(x) (length(unique(x)) == 2 & min(table(x))>1)) |> 
+    dplyr::mutate(across(everything(), as.factor)) 
+  
+  rownames(clust) <- data$id
+  
+  return(clust)
+  }

### 2022-08-27 R/prep_mca_data.R

@@ -1,0 +1,8 @@
+prep_mca_data <- function(data,
+         keep_cols) {
+  data |>
+    dplyr::select(all_of(keep_cols))  |> 
+    select_if(function(x) (length(unique(x)) == 2 & min(table(x))>1)) |> 
+    dplyr::mutate(across(everything(), as.factor)) 
+  
+}

### 2022-08-27 R/prep_subjects_data.R

@@ -1,0 +1,19 @@
+prep_subjects_data <- function() {
+  comb <- dplyr::bind_rows(
+   SMPD4::bijarnia_mahay_2022,
+    SMPD4::ji_2022,
+    SMPD4::magini_2019,
+    SMPD4::monies_2019,
+    SMPD4::ravenscroft_2020
+  ) |>
+    dplyr::mutate(
+      feeding_swallow_dysfxn_ind = tidyr::replace_na(feeding_swallow_dysfxn_ind, 0),
+      skeletal_changes_scoliosis = ifelse(skeletal_changes_scoliosis == "yes", 1, 0),
+      skeletal_changes_scoliosis = tidyr::replace_na(skeletal_changes_scoliosis, 0),
+      hip_luxation = ifelse(hip_luxation == "yes", 1, 0),
+      hip_luxation = tidyr::replace_na(hip_luxation, 0),
+      seizure = as.numeric(as.logical(seizure)),
+      seizure = tidyr::replace_na(seizure, 0),
+    )
+  
+}

### 2022-08-27 R/prep_survival_data.R

@@ -1,0 +1,7 @@
+prep_survival_data <- function(data, clusters, vars){
+  data |> 
+    mutate(cluster = as.factor(clusters)) |> 
+    mutate(deceased = as.numeric(deceased)) |> 
+    select(deceased, survival_time, all_of(vars))
+    
+  }

### 2022-08-27 _targets.R

@@ -1,0 +1,40 @@
+library(targets)
+library(tarchetypes)
+
+# Set target options
+tar_option_set(
+  packages = c("SMPD4", 'tidyverse', 'ggrepel', 'corrplot', 'cluster', 'survival', 'survminer', 'quarto', 'knitr', 'FactoMineR', 'factoextra'), 
+  format = "rds" 
+)
+
+# source files in /R
+tar_source()
+
+
+# Pipelinme
+list(
+  tar_target(subjects, prep_subjects_data(), cue = tar_cue(mode = "always")),
+  
+  # PCA
+  tar_target(mca_data, prep_mca_data(data = subjects, 
+                                    keep_cols = c(21:53, 57:60, 73, 77, 80, 86:152))),
+  tar_target(mca,  MCA(mca_data, graph = FALSE)),
+
+  # Clustering
+  tar_target(cluster_data, prep_cluster_data(data = subjects, 
+                                             keep_cols = c(21:53, 57:60, 73, 77, 80, 86:152))),
+  tar_target(clusters, hclust(dist(cluster_data, method = "binary"))),
+  tar_target(cluster_assign, cutree(clusters, k=3)),
+  tar_target(subjects_clustered, bind_cols(subjects, cluster = cluster_assign)),
+  
+  # # Survival Analysis
+  tar_target(survival_data, prep_survival_data(data = subjects_clustered,
+                                               clusters = cluster_assign,
+                                               vars = c("id", "cluster"))),
+  tar_target(kapm_baseline, survfit(Surv(survival_time, deceased) ~ 1, data = survival_data)),
+  tar_target(kapm, survfit(Surv(survival_time, deceased) ~ cluster, data = survival_data)),
+  tar_target(cox_ph, coxph(Surv(survival_time, deceased) ~ cluster, data = survival_data)),
+
+  # # Render Report
+  tar_quarto(report, path = "paper_draft.qmd")
+)

### 2022-08-27 _targets_packages.R

@@ -1,0 +1,20 @@
+# Generated by targets::tar_renv(): do not edit by hand
+library(bs4Dash)
+library(cluster)
+library(clustermq)
+library(corrplot)
+library(future)
+library(ggrepel)
+library(gt)
+library(markdown)
+library(pingr)
+library(quarto)
+library(rstudioapi)
+library(shiny)
+library(shinybusy)
+library(shinyWidgets)
+library(SMPD4)
+library(survival)
+library(survminer)
+library(tidyverse)
+library(visNetwork)

### 2022-08-27 proto.R

@@ -1,0 +1,22 @@
+
+
+
+
+clust <- mca_data |> 
+  select_if(\(x) length(unique(x)) == 2) 
+
+rownames(clust) <- subjects$id
+
+dm <- dist(clust, method = "binary")
+
+hc <- hclust(dm)
+plot(hc)
+cl <- cutree(hc, k=3)
+cl
+
+subjects$cluster <- cl
+
+subjects |> 
+  select(id, cluster, survival_time) |> View()
+  group_by(cluster) |> 
+  summarise(mean_st = mean(survival_time))

### 2022-08-27 run.R

@@ -1,0 +1,3 @@
+targets::tar_make()
+targets::tar_visnetwork() 
+

### 2022-08-29 paper_draft.qmd
more str changes

@@
 ---
+
+```{r echo=FALSE, message=FALSE}
+source('_targets_packages.R')
+library(targets)
+library(knitr)
+library(FactoMineR)
+library(factoextra)
+library(survival)
+library(clValid)
+
+source('R/utils.R')
+
+theme_set(theme_bw())
+```
 
@@
 Neurodevelopment disorders are a diverse and heterogeneous group of conditions impacting the development of the nervous system, brain function, physical development, emotional development and learning ability.
-A recently identified condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) (MIM:[618622](https://omim.org/entry/618622)) has been described in children with biallelic loss-of-function variants in *SMPD4* [@magini2019loss].
-
-Sphingomyelinases such as *SMPD4* hydrolyze sphingomyelin into ceramide and phosphorylcholine.
+A recently identified condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) (MIM:[618622](https://omim.org/entry/618622)) has been described in children with bi-allelic loss-of-function variants in *SMPD4* [@magini2019loss].
+Sphingomyelinases such as *SMPD4* play an important celluar role by hydrolyzing sphingomyelin into ceramide and phosphorylcholine.
 *SMPD4* specifically encodes one of 4 neutral sphinogmyelinases, nSMase3 (MIM: [610457](http://omim.org/entry/610457)).
 
-So far, only approximately 30 cases have been reported across over 12 families.
+So far, only approximately 30 cases of this rare disorder have been reported across over 12 families.
 Clinical phenotype data of these cases is largely heterogeneous with severe neurological complications and early demise a key feature.
@@
 In these cases individuals were able to develop independent motor skills, have mild intellectual disability and arthrogryposis without evidence of simplified gyral patterns on brain MRI.
-Other patients with truncating variants are shown to have more severe presentations, while a range of additional significant phenotypes were reported involving dysmorphic facial features, seizure, vocal cord paralysis and hearing impairment.
+Other patients with truncating variants are shown to have more severe presentations, while a range of additional significant clinical features were reported involving dysmorphic facial features, seizure, vocal cord paralysis and hearing impairment.
 
@@
 @ji2022case reports parallels with two cases showing the same homozygous null variant.
-The individual reported in @monies2019lessons presented with distinct symptoms of brain atrophy and skeletal dysplasia whereas the case in @magini2019loss with the same variant exhibited more typical clinical features.
+An individual reported in @monies2019lessons presented with distinct symptoms of brain atrophy and skeletal dysplasia whereas the case in @magini2019loss with the same variant exhibited more typical clinical features.
 
-Further work to collate and analyse the current body of research is challenging.
-While @magini2019loss cataloged a detailed clinical phenotype data set as a supplimentary to their study, these data are not suitable for statisical analysis.
-Many of the clinical features are represented as free text descriptions and a variety of non-standard terminology is applied between cases, making machine interpretation of the data difficult.
+Further work to collate and analyse data relating to NEDMABA is challenging.
+While @magini2019loss cataloged a detailed clinical phenotype data set as a supplementary data set to their study, these data are not suitable for statistical analysis as presented.
+Many of the clinical features are represented as free text descriptions and a variety of non-standard terminologies are applied between cases, making machine interpretation of the data difficult.
 Other studies [@ji2022case; @bijarnia2022growth] present only written case reports with some tabulated summaries.
-While larger studies [@ravenscroft2021neurogenetic; @monies2019lessons] offer data genotype analysis, there is less detail included on the clinical phenotype of indivuduals in the cohort.
+While larger studies [@ravenscroft2021neurogenetic; @monies2019lessons] focus more on genetic data, with less detail included on the clinical presentation of individuals in the cohort.
 
-Given the severe nature of this condition, often leading to early demise, a gap exists in understanding the expected survival time for individuals with bi-allelic null-variants in SMPD4.
-While a connection has been highlighted between some missense variants and a milder presentation with longer survival, this hypothesis has not been analysed further.
-Furthermore, the associations between diverse features of NEDMADA have not been analysed in detail, in-part due to the sparse and inconsistent collection of clinical and case report data for this new condition.
+\
+To overcome these challenges, @JSSv059i10 proposes a data structure know as 'tidy' data which organises each observation in a row, with each feature as its own column and each value in just one cell.
+In this case detailed text-based data can be tranformed to high dimensional binary indicators of key clinical features.
+This structure is conducive to effective statistical analysis and integrates deliberately with the *tidyverse* [@tidyverse] collection of packages for data analysis in R [@rbase].
+
+Further analysis to better describe rare genetic conditions was explored in @diaz2020phenotype in the context of large scale genotype-phenotype analysis.
+The authors argue patients with rare disorders (often in small samples) often present with varied symptoms that do not match exactly with the described phenotype.
+When considering this situation of low sample size data with many features, methods in the field of multivariate statistical analysis are commonly deployed.
+Here the aim is to find substructure in the data, or a simple representation of the multi-dimensional space.
+Methods, such as Multiple Correspondence Analysis [@le2010multiple] are appropriate for high dimensional data comprised of binary indicators.
+This technique is cited in many studies in the application of dimension reduction and clustering of comorbidities and phenotypes from disease [@han2018cluster; @costa2013use].
+
+When the focus is on survival or mortality, methods in statistical survival analysis are well suited and commonly used in the field of genomics [@chen2014survival].
+These methods in particular provide a mechanism for dealing with right-censoring, a common feature of clinical studies where the clinical outcome of interest may not be known by the end of the study period.
+Within the context of examining a single condition, @Crowe810 investigated comorbidity phenotypes and mortality risk in ischaemic heart disease patients.
+Here latent class analysis was performed to identify a small number of clusters in the patients which could then be included in survival analysis techniques to better understand the substructre and outcome of patients of this condition.
 
@@
 With current research highlighting a diverse and heterogeneous phenotype, the relationship and correlations between these diverse features is not well understood.
+Furthermore, the survival time of affected children is not well understood despite early-demise featuring as a severe outcome.
+While a connection has been highlighted between some missense variants and a milder presentation with longer survival, this hypothesis has not been analysed further.
 
 This research has three key aims.
-First, to collate and transform the variety of early case reports and studies on clinical phenotype data for this novel variant into an analysis-ready *tidy* dataset [@JSSv059i10].
-Secondly, to statistically quantify the expected survival time of children with NEDMABA based on the current case reports.
-Finally, detailed analysis will be conducted to further understand the associations between both patients and their clinical features.
+First, to collate and transform the variety of early case reports and studies on clinical phenotype data for this novel variant into an analysis-ready *tidy* data set.
+Secondly, to conduct analysis on the associations between both patients and their clinical features.
+Finally, to statistically quantify the expected survival time of children with NEDMABA based on the current case reports.
 
-While current research in this area has highlighted a diverse range of presentations, this will be the first in-depth analysis of early studies of this novel variant, which aims to produce statistical findings to better describe and understand this condition.
+This will be the first in-depth analysis of early studies of this novel variant, which aims to produce statistical findings to better describe and understand this condition.
 This research is significant as it will assist clinicians understand the typical and non-typical presentations of this challenging new condition.
@@
 
-The data contained in @magini2019loss "Summary of SMPD4-Related Clinical Phenotype" is an excel spreadsheet tabulating each of the 21 individuals from the study
-with clinical details recorded. The file has 21 columns (one for each individual), and 64 rows (one for each phenotype or clinical remark).
+The data contained in @magini2019loss "Summary of SMPD4-Related Clinical Phenotype" is an excel spreadsheet tabulating each of the 21 individuals from the study with clinical details recorded.
+The file has 21 columns (one for each individual), and 64 rows (one for each phenotype or clinical remark).
 
@@
 |                     | Family 1- Individual 1                                                                                     | Family 1- Individual 4                                      |
-|---------| -----------------------|------------------------|  
-Facial dysmorphisms | short palpebral fissures, large ears, simple helices, smooth philtrum, thin lips, bilateral simian creases | short palpebral fissures, receding forehead, thin upper lip |
+|------------------|--------------------------|----------------------------|
+| Facial dysmorphisms | short palpebral fissures, large ears, simple helices, smooth philtrum, thin lips, bilateral simian creases | short palpebral fissures, receding forehead, thin upper lip |
 
@@
 
-| id          | short palpebral fissures | large ears |simple helices | receding forehead | ... | 
-| ------------|------|------|------|------|-----|
-| Family 1- Individual 1  | 1 |  1 | 0 | ... |
+| id                     | short palpebral fissures | large ears | simple helices | receding forehead | ... |
+|------------|------------|------------|------------|------------|------------|
+| Family 1- Individual 1 | 1                        | 1          | 0              | ...               |     |
 | Family 1- Individual 4 | 1                        | 0          | 0              | 1                 | ... |
@@
 
-In the case where two variables were formed from phenotypes that are considered to be synonymous, these
-were merged into one indicator column to prevent duplication e.g. {bilateral_cleft_lips, bilateral_cleft_lip, cleft_lip_b_l}.
+In the case where two variables were formed from phenotypes that are considered to be synonymous, these were merged into one indicator column to prevent duplication e.g.
+{bilateral_cleft_lips, bilateral_cleft_lip, cleft_lip_b\_l}.
 
@@
 | Gender       | Birth Weight          |
-|--------|--------------|
+|--------------|-----------------------|
 | male         | 2175 grams (- 2.5 SD) |
@@
 
-This format was preserved and other case studies identified in the literature [@ravenscroft2021neurogenetic; @monies2019lessons;
-@bijarnia2022growth; @ji2022case] were manually entered to conform to this template to allow the data to be combined for further
-analysis.
+This format was preserved and other case studies identified in the literature [@ravenscroft2021neurogenetic; @monies2019lessons; @bijarnia2022growth; @ji2022case] were manually entered to conform to this template to allow the data to be combined for further analysis.
 
-These data sets were packaged into an R Package called SMPD4 [@smpd4-data] in order to allow for reproducibility and sharing. This can 
-be downloaded and installed from github at [https://github.com/deanmarchiori/SMPD4](https://github.com/deanmarchiori/SMPD4).
+These data sets were packaged into an R Package called SMPD4 [@smpd4-data] in order to allow for reproducibility and sharing.
+This can be downloaded and installed from github at <https://github.com/deanmarchiori/SMPD4>.
 
@@
 
-### Cluster Analysis
+Multiple Correspondence Analysis (MCA) is an analogy to Principle Component Analysis (PCA) for categorical data [@le2010multiple].
+
+We let a data set $\mathbf{X}$ be comprised of a set of individuals $I$ and a set of features $Q$ such that the $q_{th}$ feature has $K_q$ levels.
+The sum of all categories $K = \sum_{q=1}^{Q}K_q$ defines the dimensionality of $\mathbf{X}$ as an $I\times K$ matrix.
+
+Taking $\delta_{ik} = 1$ if subject $i$ has feature $k$ and $\delta_{ik} = 0$ if the subject does not, we are left with the completely disjnuctive table $\mathbf{X} = I\times K$ of $\{0,1\}$.
+Letting the sum of all entries of $\mathbf{X}$ be $N$, we have $\mathbf{Z} = N^{-1}\mathbf{X}$.
+
+We can introduce two diagonal matricies $\mathbf{D_r} = diag(\mathbf{r})$ and $\mathbf{D_c} = diag(\mathbf{c})$ where $\mathbf{r}$ and $\mathbf{c}$ are the vectors of row sums and column sums of $\mathbf{Z}$ respectively.
+
+Computing MCA involves taking the Singular Value Decomposition of:
+
+$$
+\mathbf{M = D_r^{-\frac{1}{2}}(Z-rc^T)D_c^{-\frac{1}{2}}} = \mathbf{P}\Delta\mathbf{Q^T}
+$$ where $\Delta$ are the singular values and $\Lambda = \Delta^2$ is the matrix of eigenvalues.
+
+This results in the row and column factor scores respectively as:
+
+$$
+\mathbf{F} = \mathbf{D_r^{-\frac{1}{2}}P}\Delta 
+$$
+
+$$
+\mathbf{G = D_c^{-\frac{1}{2}}Q}\Delta
+$$
+
+The combined clinical phenotype data was subsetted to include only those features that were encoded as a binary indicator variable.
+
+Only those features that appeared in more than one case were included to minimise the influence of non-related features in this analysis.
+
+This resulted in a dataset $X_{28 \times 61}$ indicator matrix of 61 clinical features across 28 individuals.
+
+Multiple Correspondence Analysis was then computed using the *FactoMineR* R package [@factominer].
 
@@
 
--  Explain why each procedure was done
+To model survival probability for all subjects in the combined data, the data are subsetted to include *survival_time* which is either the number of days the individual survived for, or the age in days at last follow up, and *deceased* a numeric indicator which is equal to 1 if the subject is deceased and 0 otherwise.
 
--   Experimental procedures and results are narrated in the past tense
+The Kaplan-Meier estimator [@kaplan1958nonparametric] is used to estimate the survival function of the data.
 
--   Mathematical equations and statistical tests are considered mathematical methods and should be described in this section along with the actual experimental work.
+The estimator is calcualted as:
+
+$$
+\hat{S}(t) = \prod_{t_i \leq t}\frac{n_i - d_i}{n_i} 
+$$ where $t_i$ is some event time, $d_i$ represents the number of events (here deceased subjects) and $n_i$ indicating the indivudals known to have survived up to $t_i$
+
+The baseline estimator of all individuals was calculated using the *survival* package in R [@survival-book].
+
+Next the Kaplan-Meier estimator was calculated, stratified by the reported type of Single nucleotide polymorphisms (SNP) reported in the literature.
+A test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* package.
+
+Finally, a Cox proportional hazards regression model [@andersen1982cox] was fit on survival time with the variant type as the sole covariate.
+This was to detect and measure the Hazard ratio of variant type on the survival time in order to test the hypothesis that misense variant types exhibit significantly greater survival compared to other variant types.
 
@@
 
+### MCA
+
+The first dimension ($\lambda_! = 0.19$) accounts for 19.29% of the variance in the data 
+with the first 4 dimensions of the MCA analysis accouting for 54% of the variance.
+
+```{r}
+### MCA  
+tar_load(mca)
+
+# eigenvalue summary 
+mca$eig |> 
+  kable(digits = 2, caption = "eigenvalues and percentage of variance explained by the MCA principal axes")
+```
+
+The contribution of the categories can be ranked within each of the first four
+MCA dimensions.
+
+The first dimension is dominated by highly specific facial dysmophisms such as 
+upward sweep of hair, priminent lower lip & bulbous nose. Also significant are cardiomyopathy and mild vermian hypoplasia. This principal dimension also notably 
+is categorized by the absence of simplified gyral pattern and abnormal EEG 
+with or without seizure.
+
+The second MCA dimension also has highly contributory facial features such as short philtrum, hairline related dysmorphisms and spastic tatraparesis.
+This dimension also has a notable absence of typical features such as microcephaly, arthrogryposis and hypomyelination.
+
+The third dimension is categorised by feeding and respiratory dysfunction with 
+vocal cord palsy, tracheostomy, tube feeding and GERD. The fourth dimension is best described by small for gestational age and structural brain anomalies. See @fig-ctr
+
+```{r ctr-plot, fig.height=10, fig.width=10, warning=FALSE}
+#| label: fig-ctr
+#| fig-cap: "Top 20 feature categories from MCA analysis for the first four MCA dimensions.
+The red dashed line indicates the mean contribution value.
+Each feature may be present or absent as indicated by 1 or 0 respectively"
+mca$var$contrib |> 
+  as_tibble(rownames = 'vars') |> 
+  select(1:5) |> 
+  pivot_longer(cols = -vars, names_to = "dim", values_to = "contribution") |> 
+  mutate(col = ifelse(str_detect(vars, "_0"), "absent", "present")) |> 
+  group_by(dim) |> 
+  top_n(20, wt = contribution) |> 
+  mutate(vars = fct_reorder(vars, contribution)) |> 
+  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim), fill = col)) +
+  geom_col() +
+  scale_y_reordered()+
+  scale_fill_grey(start = 0.8, end = 0.2) +
+  facet_wrap(~dim, scales = "free", ncol = 2) +
+  geom_vline(xintercept = 100/122, lty = 2, col = 'red') +
+  guides(fill=guide_legend(title="Condition")) +
+  labs(y = '')
+```
+
+```{r fig.height=8, fig.width=8,  warning=FALSE}
+fviz_mca_biplot(mca,  
+               repel = T) 
+```
+
+### Cluster Analysis
+
+```{r}
+### Clustering 
+tar_load(cluster_data)
+tar_load(cluster_select)
+tar_load(clustering)
+tar_load(cluster_assignments)
+```
+
+```{r}
+# CLValidation table or plot 
+summary(cluster_select)
+```
+
+```{r}
+cluster_select@measures |> 
+  as_tibble(rownames = 'measure') |> 
+  pivot_longer(cols =  -measure, names_to = 'cluster', values_to = "metric") |> 
+  mutate(cluster = as.numeric(str_remove(cluster, ".hierarchical"))) |> 
+  ggplot(aes(cluster, metric, colour = measure, group = measure)) +
+  geom_line()
+```
+
+```{r, fig.height=12, fig.width=10}
+# AGNES 
+plot(clustering, which.plot = 2)
+```
+
+```{r, fig.height=8, fig.width=8,  warning=FALSE}
+# plot MCA axes  
+cbind(cluster_data, clust = cluster_assignments) |> 
+  rownames_to_column(var = 'var') |> 
+  as_tibble() |> 
+  arrange(var, clust) |> 
+  ggplot(aes(dim_1, dim_2, colour = clust)) +
+  geom_point(position = "jitter") +
+  geom_label_repel(aes(label = var), size = 2, max.overlaps = 20)
+
+```
+
+```{r}
+# Profile table 
+cbind(cluster_data, clust = cluster_assignments) |> 
+  rownames_to_column(var = 'var') |> 
+  as_tibble() |> 
+  select(clust, var) |> 
+  arrange(clust) |> 
+  kable()
+```
+
+### Survival Analysis
+
+```{r}
+### Survival
+tar_load(survival_data)
+tar_load(kapm_baseline)
+tar_load(kapm)
+```
+
+```{r}
+# KM Curve
+x <- survminer::surv_summary(kapm_baseline, data = survival_data)
+kable(x, digits = 2)
+```
+
+```{r, fig.height=8, fig.width=8,  warning=FALSE}
+survminer::ggsurvplot_df(x,
+                         conf.int = TRUE,
+                         risk.table = TRUE, # Add risk table
+                         risk.table.col = "strata", # Change risk table color by groups
+                         linetype = "strata", # Change line type by groups
+                         surv.median.line = "hv", # Specify median survival
+                         ggtheme = theme_bw())
+
+```
+
+```{r, fig.height=8, fig.width=8,  warning=FALSE}
+# KM by variant
+
+# box plot of variant 
+ggplot(survival_data, aes(variant, survival_time)) +
+  geom_boxplot()
+```
+
+```{r}
+x2 <- survminer::surv_summary(kapm, data = survival_data)
+kable(x2, digits = 2)
+```
+
+```{r, fig.height=8, fig.width=8,  warning=FALSE}
+survminer::ggsurvplot_df(x2,
+                              conf.int = TRUE,
+          risk.table = TRUE, # Add risk table
+          risk.table.col = "strata", # Change risk table color by groups
+          linetype = "strata", # Change line type by groups
+          surv.median.line = "hv", # Specify median survival
+          ggtheme = theme_bw())
+
+```
+
+```{r}
+# log rank
+tar_load(logrank)
+logrank
+
+```
+
+```{r}
+# cox
+tar_load(cox_ph)
+cox_ph
+```
+
 ## Discussion
+
+```{r}
+# table with med gen 
+# cox power analysis
+# cluster for outlier
+# cluster for partitioning
+```
 
@@
 
-@bijarnia2022growth
+\newpage
 
 ## References
-
-```{r eval = FALSE}
-source('_targets_packages.R')
-library(targets)
-library(knitr)
-
-```
-
-```{r fig.height=8, fig.width=8, eval=FALSE}
-tar_load(subjects)
-tar_load(cluster_data)
-tar_load(clusters)
-tar_load(cluster_assign)
-
-cm <- as.matrix(cluster_data)
-# Check the most similar
-subjects[which(cm == min(cm[cm != min(cm)]), arr.ind = TRUE)[1, ], ]
-
-# Check the most dissimilar
-subjects[which(cm == max(cm[cm != max(cm)]), arr.ind = TRUE)[1, ], ]
-
-summary(clusters)
-plot(clusters)
-rect.hclust(clusters, k=4, border="red")
-
-tar_load(pca_data)
-
-pca_data |> 
-  bind_cols(cluster = cluster_assign) |> 
-  pivot_longer(cols = -c(cluster), names_to = 'a', values_to = 'v') |> 
-  group_by(cluster, a) |> 
-  summarise(n = sum(v)) |>  
-  arrange(n) |> 
-  ggplot(aes(n, a)) +
-  geom_col() +
-  facet_wrap(~cluster, scales = "free_x")
-
-tibble(n = 1:nrow(subjects), id = subjects$id, cluster  = cluster_assign) |> kable()
-
-tibble(n = 1:nrow(subjects), 
-       d = subjects$id, 
-       dec = subjects$deceased,
-       st = subjects$survival_time, 
-       cluster  = cluster_assign) |> 
-  group_by(cluster) |> 
-  summarise(ast = mean(st),
-            mst = max(st),
-            minst = min(st),
-            n = n(),
-            sd = sum(dec))
-
-```
-
-```{r fig.height=8, fig.width=8, eval=FALSE}
-tar_load(survival_data)
-tar_load(kapm)
-tar_load(cox_ph)
-kapm <- survfit(Surv(survival_time, deceased) ~ cluster, data = survival_data)
-summary(kapm, times= (0:60)*30, scale=30)
-x <- survminer::surv_summary(kapm, data = survival_data)
-plot(kapm)
-survminer::ggsurvplot_df(x,
-                              conf.int = TRUE,
-          risk.table = TRUE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw())
-
-survival::survdiff(Surv(survival_time, deceased) ~ cluster, data = survival_data)
-coxph(Surv(survival_time, deceased) ~ cluster, data = survival_data)
-summary(cox_ph)
-
-```
-
-```{r eval=FALSE}
-library(FactoMineR)
-library(factoextra)
-
-plot.MCA(mca,invisible=c("quali.sup","quanti.sup"),cex=0.7,  autoLab = 'yes', select  = "contrib 5", selectMod = "cos2 5") +
-  theme(panel.grid.major = element_blank(),
-        plot.title=element_text(size=14, color="blue"),
-        axis.title = element_text(size=12, color="red"))
-
-plotellipses(res.mca,keepvar=1:20)
-
-# eigenvalues
-fviz_screeplot(res.mca, addlabels = TRUE, ylim = c(0, 45))
-
-# biplot 
-fviz_mca_biplot(res.mca, 
-                repel = TRUE, # Avoid text overlapping (slow if many point)
-                ggtheme = theme_minimal())
-
-# variable plots
-
-fviz_mca_var(res.mca, choice = "var",repel = T,
-             ggtheme = theme_minimal())
-
-fviz_mca_var(res.mca, choice = "var.cat", repel = T,
-             ggtheme = theme_minimal())
-
-### quality 
-fviz_mca_var(res.mca, col.var = "cos2",
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
-             repel = TRUE, # Avoid text overlapping
-             ggtheme = theme_minimal())
-
-fviz_mca_var(res.mca, alpha.var="cos2",
-             repel = TRUE,
-             ggtheme = theme_minimal())
-
-fviz_cos2(res.mca, choice = "var", axes = 1:2)
-
-# indiv
-fviz_mca_ind(res.mca, 
-             label = "none", # hide individual labels
-             habillage = "deceased", # color by groups 
-             palette = c("#00AFBB", "#E7B800"),
-             addEllipses = TRUE, ellipse.type = "confidence",
-             ggtheme = theme_minimal())
-
-fviz_mca_ind(res.mca, col.ind = "cos2", 
-             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
-             repel = TRUE, # Avoid text overlapping (slow if many points)
-             ggtheme = theme_minimal())
-
-res.desc <- dimdesc(res.mca, axes = c(1, 2))
-
-res.desc[[2]]
-
-```

### 2022-08-29 R/prep_cluster_data.R

@@ -1,11 +1,18 @@
-prep_cluster_data <-  function(data,
-                               keep_cols) {
-  clust <- data |>
-    dplyr::select(all_of(keep_cols))  |> 
-    select_if(function(x) (length(unique(x)) == 2 & min(table(x))>1)) |> 
-    dplyr::mutate(across(everything(), as.factor)) 
+prep_cluster_data <- function(data, cols){
   
-  rownames(clust) <- data$id
+  vars <- data$var$coord |> 
+    as_tibble(rownames = 'var') |> 
+    filter(str_detect(var, "_1")) |> 
+    janitor::clean_names() |> 
+    select(var, all_of(cols)) |> 
+    as.data.frame()
   
-  return(clust)
-  }
+  rownames(vars) <- vars$var
+  
+  clust_dat <- vars[, -1]
+  
+  return(clust_dat)
+  
+}
+
+

### 2022-08-29 R/prep_subjects_data.R

@@ -7,13 +7,10 @@ prep_subjects_data <- function() {
     SMPD4::ravenscroft_2020
   ) |>
     dplyr::mutate(
-      feeding_swallow_dysfxn_ind = tidyr::replace_na(feeding_swallow_dysfxn_ind, 0),
       skeletal_changes_scoliosis = ifelse(skeletal_changes_scoliosis == "yes", 1, 0),
       skeletal_changes_scoliosis = tidyr::replace_na(skeletal_changes_scoliosis, 0),
       hip_luxation = ifelse(hip_luxation == "yes", 1, 0),
-      hip_luxation = tidyr::replace_na(hip_luxation, 0),
-      seizure = as.numeric(as.logical(seizure)),
-      seizure = tidyr::replace_na(seizure, 0),
+      hip_luxation = tidyr::replace_na(hip_luxation, 0)
     )
   
 }

### 2022-08-29 R/prep_survival_data.R

@@ -1,7 +1,12 @@
-prep_survival_data <- function(data, clusters, vars){
+prep_survival_data <- function(data, variant_data){
   data |> 
-    mutate(cluster = as.factor(clusters)) |> 
+    inner_join(variant_data, by = c("study", "family")) |> 
     mutate(deceased = as.numeric(deceased)) |> 
-    select(deceased, survival_time, all_of(vars))
+    group_by(id) |> 
+    mutate(ind = row_number()) |> 
+    pivot_wider(id_cols = c(id, variant_type.x, survival_time, deceased), names_from = ind, values_from = variant_type.y) |> 
+    unite(`1`, `2`, col = variant, na.rm = T, sep = "+") |> 
+    mutate(variant_2 = ifelse(str_detect(variant, "missense"), "missense", "other"))
     
-  }
+}
+

### 2022-08-29 R/utils.R

@@ -1,0 +1,16 @@
+reorder_within <- function(x, by, within, fun = mean, sep = "___", ...) {
+  new_x <- paste(x, within, sep = sep)
+  stats::reorder(new_x, by, FUN = fun)
+}
+
+
+scale_x_reordered <- function(..., sep = "___") {
+  reg <- paste0(sep, ".+$")
+  ggplot2::scale_x_discrete(labels = function(x) gsub(reg, "", x), ...)
+}
+
+
+scale_y_reordered <- function(..., sep = "___") {
+  reg <- paste0(sep, ".+$")
+  ggplot2::scale_y_discrete(labels = function(x) gsub(reg, "", x), ...)
+}

### 2022-08-29 _targets.R

@@ -3,7 +3,7 @@ library(tarchetypes)
 
 # Set target options
 tar_option_set(
-  packages = c("SMPD4", 'tidyverse', 'ggrepel', 'corrplot', 'cluster', 'survival', 'survminer', 'quarto', 'knitr', 'FactoMineR', 'factoextra'), 
+  packages = c("SMPD4", 'clValid', 'tidyverse', 'ggrepel', 'corrplot', 'cluster', 'survival', 'survminer', 'quarto', 'knitr', 'FactoMineR', 'factoextra'), 
   format = "rds" 
 )
 
@@ -15,26 +15,30 @@ tar_source()
 list(
   tar_target(subjects, prep_subjects_data(), cue = tar_cue(mode = "always")),
   
-  # PCA
+  # MCA
   tar_target(mca_data, prep_mca_data(data = subjects, 
                                     keep_cols = c(21:53, 57:60, 73, 77, 80, 86:152))),
   tar_target(mca,  MCA(mca_data, graph = FALSE)),
-
-  # Clustering
-  tar_target(cluster_data, prep_cluster_data(data = subjects, 
-                                             keep_cols = c(21:53, 57:60, 73, 77, 80, 86:152))),
-  tar_target(clusters, hclust(dist(cluster_data, method = "binary"))),
-  tar_target(cluster_assign, cutree(clusters, k=3)),
-  tar_target(subjects_clustered, bind_cols(subjects, cluster = cluster_assign)),
   
-  # # Survival Analysis
-  tar_target(survival_data, prep_survival_data(data = subjects_clustered,
-                                               clusters = cluster_assign,
-                                               vars = c("id", "cluster"))),
+  ## cluster analysis 
+  tar_target(cluster_data, prep_cluster_data(mca, cols = c('dim_1', 'dim_2'))),
+  tar_target(cluster_select, clValid(cluster_data,
+                                     nClust = 2:10, 
+                                     clMethods = "hierarchical", 
+                                     validation = "stability")),
+  tar_target(clustering, agnes(cluster_data)),
+  tar_target(cluster_assignments, factor(cutree(clustering, 9))),
+  
+  ## Survival Analysis
+  tar_target(survival_data, prep_survival_data(data = subjects,
+                                               variant_data = SMPD4::CADD)),
   tar_target(kapm_baseline, survfit(Surv(survival_time, deceased) ~ 1, data = survival_data)),
[…]

### 2022-08-29 proto.R

@@ -1,22 +1,0 @@
-
-
-
-
-clust <- mca_data |> 
-  select_if(\(x) length(unique(x)) == 2) 
-
-rownames(clust) <- subjects$id
-
-dm <- dist(clust, method = "binary")
-
-hc <- hclust(dm)
-plot(hc)
-cl <- cutree(hc, k=3)
-cl
-
-subjects$cluster <- cl
-
-subjects |> 
-  select(id, cluster, survival_time) |> View()
-  group_by(cluster) |> 
-  summarise(mean_st = mean(survival_time))

### 2022-08-29 run.R

@@ -1,3 +1,2 @@
 targets::tar_make()
 targets::tar_visnetwork() 
-

### 2022-09-01 paper_draft.qmd
first draft'

@@
 
+```{r}
+# table with med gen 
+# Better tables with gtsummary
+```
+
 ## Title
 
--   the name of the organism studied
-
--   the particular aspect or system studied
-
--   the variable(s) manipulated.
+Statistical Clinical Phenotype and Patient Survival Analysis in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA)
 
@@
 
-passive voice
+A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with in children with bi-allelic loss-of-function variants in *SMPD4*.
 
-(typically 150-300 words)
+Few case reports have been described to date, with a diverse and severe clinical phenotype in most reported cases, the progression of this condition is not well understood.
 
--   A brief introduction to the topic that you're investigating.
+A gap exists in the understanding the associations of the heterogenous features present in the clinical phenotype and the expected survival probabilities of indivudals with this condition.
+This is driven in part to the paucity of analysis-ready data on subjects.
 
--   Explanation of why the topic is important in your field/s.
+This research aims to collate and standardise available case reports, to analyse and identify
+meaningful clusters in the clinical phenotype and the quantify the survival probability
+function for children with NEDMABA.
 
--   Statement about what the gap is in the research.
+To overcome the challenge of sparse, multidimensional data on very few subjects, we 
+employ Multiple Correspondece Analysis as a dimension reduction technique, which is 
+then subject to heirarchical clustering and interpretation. To quantify more accurate 
+survival probabilities, kaplan-meier estimation is formulated to account for censoring
+in the data.
 
--   Your research question/s / aim/s.
+Several distinct clusters of clinical features were discovered in the current case reports
+which can help explain the variation present in the clinical phenotype for children 
+with this condition. Furthermore, survival probability for those affected declines
+sharply in early infancy, but exhibits a wide range of outcomes provisionally 
+associated with variant type.
 
--   An indication of your research methods and approach.
+We are able to identify the classic phenotype for this condition, as well as two
+other auxillary feature clusters. The first relates findings of vocal cord paralysis and
+swallowing dysfunction, the other relates to more complex brain anomalies. In addition, 
+several outlying or non-typical features are able to be classified. Despite strong
+indications of differences in survival outcomes based on variant type, interpretation
+of these results are guarded based on very low sample sizes.
 
--   Your key message.
-
--   A summary of your key findings.
-
--   An explanation of why your findings and key message contribute to the field/s
+Ultra rare conditions are a challenge for clinicians and families, with often limited
+information on diverse outcomes. Despite challenges with sparse and inconsistent data, 
+analytical techniques to better describe associations and trajectories of indivudals
+with these conditions can provide clinicians and genetic counsellors with 
+enhanced information to aide in decision making and support for families.
 
@@
 
+\[NEED TO INCLUDE CONTRIBUTION\]
+
 Multiple Correspondence Analysis was then computed using the *FactoMineR* R package [@factominer].
+
+### Cluster Analysis
+
+\[insert clvalid details\]
 
@@
 
-Next the Kaplan-Meier estimator was calculated, stratified by the reported type of Single nucleotide polymorphisms (SNP) reported in the literature.
+Next the Kaplan-Meier estimator was calculated, \[ON WHAT?\] stratified by the reported type of Single nucleotide polymorphisms (SNP) reported in the literature.
 A test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* package.
@@
 
-The function of this section is to summarize general trends in the data without comment, bias, or interpretation.
-The results of statistical tests applied to your data are reported in this section although conclusions about your original hypotheses are saved for the Discussion section.
-
 ### MCA
 
-The first dimension ($\lambda_! = 0.19$) accounts for 19.29% of the variance in the data 
-with the first 4 dimensions of the MCA analysis accouting for 54% of the variance.
+The initial results of the MCA were inspected with a bi-plot and identified two individuals (17, 18) from the data that were clear outliers and sat significatly beyound the 95% confidence ellipse.
+These individuals represented two twins from @magini2019loss who were described with a much milder phenotype and wer subject to a missense variation in SMPD4.
+Further analysis of this influence is conducted below.
+For now, these subjects were removed to allow for more meaningful analysis.
+
+Repeating MCA for the remaining subjects shows the first dimension ($\lambda_!
+= 0.18$) accounts for 17.88% of the variance in the data with the first 4 dimensions of the MCA analysis accounting for 54% of the variance.
+
+```{r}
+tar_load(mca_prelim)
+```
+
+```{r fig.height=8, fig.width=8,  warning=FALSE}
+#| label: fig-biplotprelim
+#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables"
+fviz_mca_biplot(
+  mca_prelim,
+  axes = c(1, 2),
+  geom = c("point", "text"),
+  geom.ind = c("point", "text"),
+  geom.var = c("point", "text"),
+  repel = TRUE,
+  label = "all",
+  invisible = "none",
+  habillage = "none",
+  addEllipses = T,
+  ellipse.level=0.95,
+  palette = "grey",
+  col.ind = "red",
+  col.var = "black",
+  arrows = c(FALSE, FALSE),
+  map = "symmetric",
+  title = ""
+)
+```
 
@@
 
-The contribution of the categories can be ranked within each of the first four
-MCA dimensions.
+The *contribution* of the categories were ranked within each of the first four MCA dimensions See @fig-ctr.
 
-The first dimension is dominated by highly specific facial dysmophisms such as 
-upward sweep of hair, priminent lower lip & bulbous nose. Also significant are cardiomyopathy and mild vermian hypoplasia. This principal dimension also notably 
-is categorized by the absence of simplified gyral pattern and abnormal EEG 
-with or without seizure.
+The first dimension is dominated by highly specific facial dysmophisms such as epicanthal folds, short philtrum and low anteriour hairline. Also significant are seizure and developmental delay.
 
-The second MCA dimension also has highly contributory facial features such as short philtrum, hairline related dysmorphisms and spastic tatraparesis.
-This dimension also has a notable absence of typical features such as microcephaly, arthrogryposis and hypomyelination.
+The second MCA dimension is categorised by feeding and respiratory dysfunction with vocal cord palsy, tracheostomy, tube feeding and GERD.
 
-The third dimension is categorised by feeding and respiratory dysfunction with 
-vocal cord palsy, tracheostomy, tube feeding and GERD. The fourth dimension is best described by small for gestational age and structural brain anomalies. See @fig-ctr
+Examining the bi-plot of the first two principal axes @fig-biplot shows contrasting sources of variation between conditions relating to airway and swallowing function and predominantly structural brain anomalies such as as cerebellar hypoplasia, agenesis of corpus collosum and development delay.
+
+The third dimension is characterised by complex non-typical features such as patent ductus arteriosus, respiratory infections, small for gestational age and further facial features.
+These are contrasted by the fourth principal dimension which is typical for respiratory distress including stridor and hypercapnia.
 
@@
   group_by(dim) |> 
-  top_n(20, wt = contribution) |> 
+  top_n(10, wt = contribution) |> 
   mutate(vars = fct_reorder(vars, contribution)) |> 
@@
 ```{r fig.height=8, fig.width=8,  warning=FALSE}
-fviz_mca_biplot(mca,  
-               repel = T) 
+#| label: fig-biplot
+#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables"
+fviz_mca_biplot(
+  mca,
+  axes = c(1, 2),
+  geom = c("point", "text"),
+  geom.ind = c("point"),
+  geom.var = c("point", "text"),
+  repel = TRUE,
+  label = "all",
+  invisible = "none",
+  habillage = "none",
+  addEllipses = T,
+  ellipse.level=0.95,
+  palette = "grey",
+  col.ind = "grey80",
+  col.var = "black",
+  arrows = c(FALSE, FALSE),
+  map = "symmetric",
+  title = ""
+)
+```
+
+```{r fig.height=8, fig.width=8,  warning=FALSE}
+#| label: fig-biplot2
+#| fig-cap: "A bi-plot of principal axes 3 and 4, with grey cicles indicating the individual's projected position and black triangles indicaating the variables"
+fviz_mca_biplot(
+  mca,
+  axes = c(3, 4),
+  geom = c("point", "text"),
+  geom.ind = "point",
+  geom.var = c("point", "text"),
+  repel = TRUE,
+  label = "all",
+  invisible = "none",
+  habillage = "none",
+  addEllipses = T,
+  ellipse.level=0.95,
+  palette = "grey",
+  col.ind = "grey80",
+  col.var = "black",
+  arrows = c(FALSE, FALSE),
+  map = "symmetric",
+  title = ""
+)
+
 ```
@@
 
-```{r}
-# CLValidation table or plot 
-summary(cluster_select)
-```
+Internal cluster validation measures resulted in between 2 and 6 clusters as the optimal choice.
+
+Proceeding with the optimal cluster selection, per the Silhouette coefficient of 6 clusters.
+
+A scatterplot of the cluster results @fig-clust projected on just the first two dimensions highlights a core cluster of classic features of NEDMABA.
+A separate cluster is formed of airway and feeding related conditions involing vocal cord paralysis, tracheostomy, tube feeding and enlarged ventricles.
+
+A cluster relating to structual brain anomalies is articulated with conditions such as cerebellar hypoplasia and agenesis of corpus collosum, along with small for gestational age.
+
+Some smaller satellite clusters exists where highly distinctive features were associated with an indivual or family.
+Also a singleton clusters exist for features that are distinct or unusual such as posteriorly angulated ears.
 
@@
   pivot_longer(cols =  -measure, names_to = 'cluster', values_to = "metric") |> 
+  mutate(`optimal cluster no.` = as.numeric(str_remove(cluster, ".hierarchical")))  |> 
+  select(-cluster) |> 
+  group_by(measure) |> 
+  filter(metric == min(metric)) |> 
+  kable(digits = 2, caption = "Optimal cluster numbers determined by four similar metrics measuring cluster analysis performance.")
+```
+
+```{r, fig.width=8,  warning=FALSE}
+#| label: fig-clvalid
+#| fig-cap: "cluster validation analysis showing various metrics for cluster evaluation from 2 to 10 clusters.
+The most optimal number of clusters are selected by picking the lowest value from one or many of these compariable metrics."
+cluster_select@measures |> 
+  as_tibble(rownames = 'measure') |> 
+  pivot_longer(cols =  -measure, names_to = 'cluster', values_to = "metric") |> 
   mutate(cluster = as.numeric(str_remove(cluster, ".hierarchical"))) |> 
   ggplot(aes(cluster, metric, colour = measure, group = measure)) +
-  geom_line()
+  geom_line() +
+  labs(x = 'number of clusters') +
+  facet_wrap(~measure, scales= 'free')
 ```
@@
 ```{r, fig.height=12, fig.width=10}
-# AGNES 
-plot(clustering, which.plot = 2)
+#| label: fig-dendro
+#| fig-cap: "Dendrogram of heirarchical clustering method"
+plot(clustering, which.plots = 2, main = NA, sub = NA, xlab = NA)
 ```
@@
 ```{r, fig.height=8, fig.width=8,  warning=FALSE}
-# plot MCA axes  
+#| label: fig-clust
+#| fig-cap: "A projection of features on the first two principal components with cluster assignment indicated by colour."
 cbind(cluster_data, clust = cluster_assignments) |> 
@@
   ggplot(aes(dim_1, dim_2, colour = clust)) +
-  geom_point(position = "jitter") +
-  geom_label_repel(aes(label = var), size = 2, max.overlaps = 20)
-
-```
-
-```{r}
-# Profile table 
-cbind(cluster_data, clust = cluster_assignments) |> 
-  rownames_to_column(var = 'var') |> 
-  as_tibble() |> 
-  select(clust, var) |> 
-  arrange(clust) |> 
-  kable()
+    geom_vline(xintercept = 0, lty = 2, colour = "grey50") +
+  geom_hline(yintercept = 0, lty = 2, colour = "grey50") +
+  geom_label_repel(aes(label = var), size = 2, max.overlaps = 20) +
+  geom_point(aes(colour = clust), position = "jitter") +
+  labs(x = "Dimension 1",
+       y = "Dimension 2",
+       colour = 'cluster')
 ```
@@
 
+The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with survival probability at time 0 of 0.82 (0.69, 0.98).
+Survival probability after one year at 0.403 (0.252, 0.644) with median survival of 150 days and survival probability at close to ten years of 0.24 (0.11, 0.52).
+
 ```{r}
@@
 x <- survminer::surv_summary(kapm_baseline, data = survival_data)
-kable(x, digits = 2)
+kable(x, digits = 2, caption = "Survival probabilities derived from Kaplan-Meier estimation. ")
 ```
@@
 ```{r, fig.height=8, fig.width=8,  warning=FALSE}
+#| label: fig-surv
+#| fig-cap: "Kaplan Meier survival curve"
 survminer::ggsurvplot_df(x,
+                         color = "black",
                          conf.int = TRUE,
+                         conf.int.fill = "gray",
                          risk.table = TRUE, # Add risk table
@@
                          surv.median.line = "hv", # Specify median survival
-                         ggtheme = theme_bw())
+                         ggtheme = theme_bw(),
+                         legend = "none") +
+  labs(x = "time (days)")
 
 ```
+
+A comparitive boxplot of survival times by variant type shows significant outliers for missense+splice and missense variations.
+However, even within missense hymozygous variants the survival time varies greatly.
+Other missense variants, such as heterozygous missense with gene deletion do not show the same increased survival time as indeed noted by @magini2019loss.
 
@@
 # box plot of variant 
-ggplot(survival_data, aes(variant, survival_time)) +
+ggplot(survival_data, aes(survival_time, variant,fill = variant_type.x)) +
   geom_boxplot()
@@
 
-```{r}
-x2 <- survminer::surv_summary(kapm, data = survival_data)
-kable(x2, digits = 2)
-```
-
-```{r, fig.height=8, fig.width=8,  warning=FALSE}
-survminer::ggsurvplot_df(x2,
-                              conf.int = TRUE,
-          risk.table = TRUE, # Add risk table
-          risk.table.col = "strata", # Change risk table color by groups
-          linetype = "strata", # Change line type by groups
-          surv.median.line = "hv", # Specify median survival
-          ggtheme = theme_bw())
-
-```
+The log rank test was carried out to compare the suvival curves of the various gene variants $\chi^2 = 14.8$ $(p = 0.04)$ however due to the small sample size, several variants had expected counts less than 5 which would create a violation in assumptions for the test.
 
@@
 logrank
-
-```
-
-```{r}
-# cox
-tar_load(cox_ph)
-cox_ph
 ```
@@
 
-```{r}
-# table with med gen 
-# cox power analysis
-# cluster for outlier
-# cluster for partitioning
-```
-
 -   the relationship between the results and the original hypothesis, i.e., whether they support the hypothesis, or cause it to be rejected or modified
 
--   an integration of your results with those of previous studies in order to arrive at explanations for the observed phenomena
+The Multiple Correspondence Analysis confirmed the heterogeneous variation in the data with the first four princicple components being required to explain more than 50% of the variation in the dataset.
 
--   possible explanations for unexpected results and observations, phrased as hypotheses that can be tested by realistic experimental procedures, which you should describe
+The examination of key contributing factors, as well as a visual inspection of the dat projected into this lower dimensional space, indicate the primary sources of variation in the clinical phenotype can be explained by few indivudals with very different clinical features.
+This has confounded the typical or classical presentation.
 
--   End the Discussion with a summary of the principal points you want the reader to remember
+Beyond these outliers, a interesting disticntion is made between the hallmark features of microcephaly, hypomyelination, arthrogryposis, simplified gyal pattern and developmental delay with specific airway and swallowing features such as vocal cord paralysis, tracheostomy, GERD and tube feeding.
+
+Further distictions are possible with more specific structural brainanomalies exhibiting as distinct sources of variation in the clininal phenotype.
+
+Many of the outlying features identifes are specific facial or anatomical dysmophims.
+While a noted and core feature is around course facial features, those with a milder course had clinical reports in the available data that articulated these dysmophims in much more detail.
+
+On a similar note, the reporting of 'talipes' as acondition seems distict as a source of variation when appearing as an isolated feature without the typical microcephasly and hypomyelination.
+However, there was significant heterogenity in how feet contractures were reported in clinical case reports.
+A distinction was made between reported cases with rocker bottom feed or congenital vertical talus and those with clubfeet/talipes.
+In practice its not clear how these diagnoses were arived at, and given the non-specific nature of this feature, care needs to be taken in distinguishing between similar manifesting features in the literaure where possible.
+
+The cluster analysis does not aim to group individuals that are alike, but rather the clinical features themselves.
+This highlights common sources of variation on the clinical phenotype of the condition itself, rather than typical coincident features of any one child.
+While the number of clusters chosen all represent valid partition of the data, a larger number was selected to align the goals of the analysis which was to uncover distinct groups.
+A smaller number would have likely yeilded isolated outliers from the main body of the features.
+This still occurs, however more substructure is detected with the higher cluster number.
+
+In terms of patient survival, the first few months show a sharp decline in survival probability confirming early demise in the infant period as a key factor.
+A small number of cases show much longer survival into the childhood and in some cases into the second decade of life.
+The current literature suggests the type of gene variant has an impact on surivival time, which seems to be supported in this work, however the limited sample size here prevents other meaningful analysis such as logrank tests or cox proprtional hazard modelling which are likely underpowered when comparing multiple groups.
+
+Despite NEDMABA exhibting a diverse and hetergeneous clinical phenotype, it is possible to identify meaningful substructures in this data through the use of dimension reduction techniques (MCA) and cluster analysis.
+This has isolated the classic NEDMABA phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyal pattern and developmental delay.
+Furthermore it has isolated additional distinct groupings of features that are exhibited in patients such as vocal cord paralysis with tracheostomy, swallowing dysfunction, tube feeding and GERD.
+In additional to typical MRI findings of symplified gyral pattern and hypomyelination, there exists clusters of structural anomalies of the corpus calllosum and cereballa hypoplasia.
+
+The survival of indivuduals with this condition is critical given early demise in infancy and childhood is a noted feature.
+This work identifies the survival probabilites and shows a one year survival rate of 0.403 (0.252, 0.644).
+A naive estimate, using just those number of deceased patients as proportion of the study population is 0.57.
+This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to overestimate the survival probability.
+
+\newpage
+
+## Appendix
+
+```{r}
+16/28
+# Profile table 
+cbind(cluster_data, clust = cluster_assignments) |> 
+  rownames_to_column(var = 'var') |> 
+  as_tibble() |> 
+  select(clust, var) |> 
+  arrange(clust) |> 
+  kable()
+```
 

### 2022-09-01 _targets.R

@@ -13,11 +13,14 @@ tar_source()
 
 # Pipelinme
 list(
-  tar_target(subjects, prep_subjects_data(), cue = tar_cue(mode = "always")),
+  tar_target(subjects, prep_subjects_data()),
   
   # MCA
-  tar_target(mca_data, prep_mca_data(data = subjects, 
+  tar_target(mca_data_prelim, prep_mca_data(data = subjects, 
                                     keep_cols = c(21:53, 57:60, 73, 77, 80, 86:152))),
+  tar_target(mca_prelim,  MCA(mca_data_prelim, graph = FALSE)),
+  tar_target(mca_data, prep_mca_data(data = slice(subjects, -c(17, 18)), 
+                                     keep_cols = c(21:53, 57:60, 73, 77, 80, 86:152))),
   tar_target(mca,  MCA(mca_data, graph = FALSE)),
   
   ## cluster analysis 
@@ -25,9 +28,9 @@ list(
   tar_target(cluster_select, clValid(cluster_data,
                                      nClust = 2:10, 
                                      clMethods = "hierarchical", 
-                                     validation = "stability")),
+                                     validation = "internal")),
   tar_target(clustering, agnes(cluster_data)),
-  tar_target(cluster_assignments, factor(cutree(clustering, 9))),
+  tar_target(cluster_assignments, factor(cutree(clustering, 6))),
   
   ## Survival Analysis
   tar_target(survival_data, prep_survival_data(data = subjects,

### 2022-09-05 paper.qmd
draft refinements

--- 2022-09-01 paper_draft.qmd
@@
 ---
-title: "paper_draft"
-author: "DM"
+title: "Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies"
+author: "Dean Marchiori"
 format: pdf
+abstract: "A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*.
+The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and diverse clinical phenotype.
+A gap exists in the understanding the associations of the heterogenous features present in the clinical phenotype and the expected survival probabilities of affected individuals, driven in part to the paucity of analysis-ready data.
+This research aims to collate and standardise available case reports to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA.
+To overcome the challenge of sparse, multidimensional data on very few subjects, we employ Multiple Correspondence Analysis as a dimension reduction technique, which is then subject to heirarchical clustering and interpretation.
+To quantify more accurate survival probabilities, Kaplan-Meier estimation is formulated to account for censoring in the data.
+The analysis identified the classic phenotype for this condition, as well as two other distinct feature clusters.
+The first relates to findings of vocal cord paralysis and swallowing dysfunction, the other relates to more complex brain anomalies.
+Furthermore, the survival probability for those affected was found to decline sharply in early infancy, but exhibited a wide range of outcomes provisionally associated with variant type.
+However, interpretation of these results are guarded based on very low sample sizes.
+Ultra rare conditions present a challenge for clinicians and families, with often limited information and diverse possible outcomes.
+Despite challenges with sparse and inconsistent data, this analysis helps describe associations and trajectories of indivudals with this condition and can provide clinicians and genetic counsellors with better information to aide in decision making and support for families."
 editor: visual
@@
 ```{r echo=FALSE, message=FALSE}
-source('_targets_packages.R')
 library(targets)
-library(knitr)
-library(FactoMineR)
-library(factoextra)
-library(survival)
-library(clValid)
-
-source('R/utils.R')
-
+source(here::here('_targets_packages.R'))
+source(here::here('R', 'utils.R'))
 theme_set(theme_bw())
@@
 
-## Title
-
-Statistical Clinical Phenotype and Patient Survival Analysis in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA)
-
-## Abstract
-
-A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with in children with bi-allelic loss-of-function variants in *SMPD4*.
-
-Few case reports have been described to date, with a diverse and severe clinical phenotype in most reported cases, the progression of this condition is not well understood.
-
-A gap exists in the understanding the associations of the heterogenous features present in the clinical phenotype and the expected survival probabilities of indivudals with this condition.
-This is driven in part to the paucity of analysis-ready data on subjects.
-
-This research aims to collate and standardise available case reports, to analyse and identify
-meaningful clusters in the clinical phenotype and the quantify the survival probability
-function for children with NEDMABA.
-
-To overcome the challenge of sparse, multidimensional data on very few subjects, we 
-employ Multiple Correspondece Analysis as a dimension reduction technique, which is 
-then subject to heirarchical clustering and interpretation. To quantify more accurate 
-survival probabilities, kaplan-meier estimation is formulated to account for censoring
-in the data.
-
-Several distinct clusters of clinical features were discovered in the current case reports
-which can help explain the variation present in the clinical phenotype for children 
-with this condition. Furthermore, survival probability for those affected declines
-sharply in early infancy, but exhibits a wide range of outcomes provisionally 
-associated with variant type.
-
-We are able to identify the classic phenotype for this condition, as well as two
-other auxillary feature clusters. The first relates findings of vocal cord paralysis and
-swallowing dysfunction, the other relates to more complex brain anomalies. In addition, 
-several outlying or non-typical features are able to be classified. Despite strong
-indications of differences in survival outcomes based on variant type, interpretation
-of these results are guarded based on very low sample sizes.
-
-Ultra rare conditions are a challenge for clinicians and families, with often limited
-information on diverse outcomes. Despite challenges with sparse and inconsistent data, 
-analytical techniques to better describe associations and trajectories of indivudals
-with these conditions can provide clinicians and genetic counsellors with 
-enhanced information to aide in decision making and support for families.
-
-## Introduction
+# Introduction
 
@@
 *SMPD4* specifically encodes one of 4 neutral sphinogmyelinases, nSMase3 (MIM: [610457](http://omim.org/entry/610457)).
-
-So far, only approximately 30 cases of this rare disorder have been reported across over 12 families.
+So far, less than 50 cases of this rare disorder have been reported in the literature.
 Clinical phenotype data of these cases is largely heterogeneous with severe neurological complications and early demise a key feature.
@@
 
-@ravenscroft2021neurogenetic performed a study over 190 probands with a diagnosis of arthrogryposis multiplex congenita, distal arthrogryposis, fetal akinesia deformation sequence or multiple pterygium syndrome.
-This study identified a novel missense variation in SMPD4 which at the time of whole exome sequencing was not well described in the literature.
+@ravenscroft2021neurogenetic performed a study over 190 probands and identified a novel missense variation in SMPD4 which at the time of whole exome sequencing was not well described in the literature.
 The impacted family from Melbourne exhibited features involving arthrogryposis multiplex congenita, complex brain malformations, small for gestation age and hypoplasia of the corpus callosum.
@@
 
-@monies2019lessons described two families with different homozygous truncating variants in SMPD4 as a syndrome of skeletal dysplasia with cerebella atrophy.
-The phenotypes described involved bilateral clenched hands, talipes, IUGR, partial absence of corpus callosum with family history of three neonatal deaths with similar features.
-
 A detailed study by @magini2019loss involved 12 unrelated familes with 32 individuals (21 with detailed clinical information).
-Hallmark presentations were of microcephaly, simplified gyration, hypomyelination, thin corpus callosum, mild cerebellar hypoplasia, brainstem hypoplasia, congenital arthrogryposis, diabetes mellitus, heart disease, severe encephalopathy and respiratory problems often leading to early demise.
+Presentations were of microcephaly, simplified gyration, hypomyelination, thin corpus callosum, mild cerebellar hypoplasia, brainstem hypoplasia, congenital arthrogryposis, diabetes mellitus, heart disease, severe encephalopathy and respiratory problems often leading to early demise.
 Despite this being the largest cohort studied, the clinical features and survival times among participants varied greatly.
@@
 
-A recent case study from China is described in @ji2022case involving a girl presenting in infancy with intrauterine growth restriction, microcephaly, postnatal developmental delay, arthrogryposis, hypertonicity, seizure, and hypomyelination on brain magnetic resonance imaging.
-@ji2022case argues the actual prevalence of this condition, based on the gene carrier rate is understated and may be attributable to typical symptoms of NEDMABA being non-specific, providing a diagnostic challenge in a clinical setting.
+A case study from China is described in @ji2022case involving a girl presenting in infancy with intrauterine growth restriction, microcephaly, postnatal developmental delay, arthrogryposis, hypertonicity, seizure, and hypomyelination on brain magnetic resonance imaging.
+The authors report parallels with two cases showing the same homozygous null variant.
+An individual reported in @monies2019lessons presented with distinct symptoms of brain atrophy and skeletal dysplasia whereas the case in @magini2019loss with the same variant exhibited more typical clinical features.
 
@@
 Notably, there is also additional non-typical clinical findings of nystagmus and visual impairment secondary to macular dystrophy and retinal pigment epithelial stippling at posterior pole.
-
-@ji2022case reports parallels with two cases showing the same homozygous null variant.
-An individual reported in @monies2019lessons presented with distinct symptoms of brain atrophy and skeletal dysplasia whereas the case in @magini2019loss with the same variant exhibited more typical clinical features.
 
@@
 
-\
-To overcome these challenges, @JSSv059i10 proposes a data structure know as 'tidy' data which organises each observation in a row, with each feature as its own column and each value in just one cell.
+To overcome these data challenges, @JSSv059i10 proposes a data structure know as 'tidy' data which organises each observation in a row, with each feature as its own column and each value in just one cell.
 In this case detailed text-based data can be tranformed to high dimensional binary indicators of key clinical features.
@@
 
-When the focus is on survival or mortality, methods in statistical survival analysis are well suited and commonly used in the field of genomics [@chen2014survival].
+WIn terms of patient survival, methods in statistical survival analysis are well suited and commonly used in the field of genomics [@chen2014survival].
 These methods in particular provide a mechanism for dealing with right-censoring, a common feature of clinical studies where the clinical outcome of interest may not be known by the end of the study period.
 Within the context of examining a single condition, @Crowe810 investigated comorbidity phenotypes and mortality risk in ischaemic heart disease patients.
-Here latent class analysis was performed to identify a small number of clusters in the patients which could then be included in survival analysis techniques to better understand the substructre and outcome of patients of this condition.
+Here latent class analysis was performed to identify a small number of clusters in the patients which could then be included in survival analysis techniques to better understand the comparative outcomes of patients of this condition.
 
@@
 Furthermore, the survival time of affected children is not well understood despite early-demise featuring as a severe outcome.
-While a connection has been highlighted between some missense variants and a milder presentation with longer survival, this hypothesis has not been analysed further.
+While a connection has been highlighted between some missense variants and a milder presentation with longer survival [@magini2019loss], this hypothesis has not been analysed further.
 
@@
 First, to collate and transform the variety of early case reports and studies on clinical phenotype data for this novel variant into an analysis-ready *tidy* data set.
-Secondly, to conduct analysis on the associations between both patients and their clinical features.
+Secondly, to conduct analysis on the associations between commonly reported clinical features.
 Finally, to statistically quantify the expected survival time of children with NEDMABA based on the current case reports.
@@
 This will be the first in-depth analysis of early studies of this novel variant, which aims to produce statistical findings to better describe and understand this condition.
-This research is significant as it will assist clinicians understand the typical and non-typical presentations of this challenging new condition.
+This research will assist clinicians understand the typical and non-typical presentations of this challenging new condition.
 In addition, genetic counselling of families with affected children will benefit from enhanced analysis on outcomes of existing reported cases.
 
-## Methods
+# Methods
 
-### SMPD4 Data Package
+## SMPD4 Data Package
 
@@
 
+This resulted in a long-formatted data set of 21 rows (individuals) and 64 columns (clinical features).
+
 In many cases, key clinical information was entered as free-text descriptions, which rendered any attempt of meaningful analysis impractical (@tbl-nontidy).
@@
 | id                     |      short palpebral fissures       | large ears | simple helices | receding forehead | ... |
-|------------|------------|------------|------------|------------|------------|
+|------------------------|-----------------|-------|--------|-------|-------|
 | Family 1- Individual 1 | 1                        | 1          | 0              | ...               |     |
@@
 In the case where two variables were formed from phenotypes that are considered to be synonymous, these were merged into one indicator column to prevent duplication e.g.
-{bilateral_cleft_lips, bilateral_cleft_lip, cleft_lip_b\_l}.
+{bilateral_cleft_lips, bilateral_cleft_lip, cleft_lip_b\_l} -\> {bilateral_cleft_lip}.
 
@@
 
-The final dataset consisted of 21 observations (one per individual) and 152 variables (one per clinical feature).
+| Gender | Birth Weight (g) |
+|--------|------------------|
+| male   | 2175             |
+| female | 2045             |
+| female | 2300             |
+| female | NA               |
+
+: Example of consistent and appropriate coding or data types {#tbl-data-types2}
 
@@
 
+The final dataset consisted of 21 observations (one per individual) and 152 variables (one per clinical feature).
+In some cases, were obvious, these variables were left as qualitative character descriptions, in other cases as noted above, distinct lists of features were tranformed into binary indicators.
+Other fields include background information on the patient and various biomarkers.
+A full list of this data set is available in @tbl-dict.
 These data sets were packaged into an R Package called SMPD4 [@smpd4-data] in order to allow for reproducibility and sharing.
-This can be downloaded and installed from github at <https://github.com/deanmarchiori/SMPD4>.
 
-### Multiple Correspondence Analysis
+## Multiple Correspondence Analysis
 
-Multiple Correspondence Analysis (MCA) is an analogy to Principle Component Analysis (PCA) for categorical data [@le2010multiple].
+The data on subjects introduced above is subset to include only variables that can be transformed into binary indicator variables that represent the presence or absence of a given clinical feature column.
+Only those features that appeared in more than one case were included to minimise the influence of non-related features in this analysis.
 
-We let a data set $\mathbf{X}$ be comprised of a set of individuals $I$ and a set of features $Q$ such that the $q_{th}$ feature has $K_q$ levels.
+Dimension reduction on this wide data set was performed using Multiple Correspondence Analysis (MCA) [@le2010multiple] in order to arrive at a set of meaningfully small dimensions that account for most of the variation in the data.
+Multiple Correspondence Analysis (MCA) is an analogy to Principle Component Analysis (PCA) for categorical data.
+
+We let these data be $\mathbf{X}$ which is comprised of a set of individuals $I$ and a set of features $Q$ such that the $q_{th}$ feature has $K_q$ levels.
 The sum of all categories $K = \sum_{q=1}^{Q}K_q$ defines the dimensionality of $\mathbf{X}$ as an $I\times K$ matrix.
-
+This resulted in a dataset $X_{28 \times 61}$ indicator matrix of 61 clinical features across 28 individuals.
 Taking $\delta_{ik} = 1$ if subject $i$ has feature $k$ and $\delta_{ik} = 0$ if the subject does not, we are left with the completely disjnuctive table $\mathbf{X} = I\times K$ of $\{0,1\}$.
 Letting the sum of all entries of $\mathbf{X}$ be $N$, we have $\mathbf{Z} = N^{-1}\mathbf{X}$.
-
 We can introduce two diagonal matricies $\mathbf{D_r} = diag(\mathbf{r})$ and $\mathbf{D_c} = diag(\mathbf{c})$ where $\mathbf{r}$ and $\mathbf{c}$ are the vectors of row sums and column sums of $\mathbf{Z}$ respectively.
 
-Computing MCA involves taking the Singular Value Decomposition of:
+Computing MCA involves taking the Singular Value Decomposition:
 
@@
 $$
-\mathbf{F} = \mathbf{D_r^{-\frac{1}{2}}P}\Delta 
+\begin{split}
+\mathbf{F} = \mathbf{D_r^{-\frac{1}{2}}P}\Delta \\
+\mathbf{G = D_c^{-\frac{1}{2}}Q}\Delta
+\end{split}
 $$
 
+This computation of the Multiple Correspondence Analysis was computed using the *FactoMineR* R package [@factominer].
+
+The *contribution* of each category is computed for each of the principal axis per @le2010multiple. This metric identifies the proportion of variance of that axis due to the point and is defined as:
+
 $$
-\mathbf{G = D_c^{-\frac{1}{2}}Q}\Delta
+Ctr_k = \frac{py_k^2}{\lambda}
 $$
+Where $p$ represents the point weighting, $y$ is the coordinate relative to the principal axes of variance $\lambda$.
 
-The combined clinical phenotype data was subsetted to include only those features that were encoded as a binary indicator variable.
+## Cluster Analysis
 
-Only those features that appeared in more than one case were included to minimise the influence of non-related features in this analysis.
+The MCA coordinates for the first 4 principal dimensions were subject to clustering in order to find groups of features
+that were internally homogeneous and externally heterogeneous. This was computed in R using the clValid package [@clvalid].
 
-This resulted in a dataset $X_{28 \times 61}$ indicator matrix of 61 clinical features across 28 individuals.
+A sensible range of clusters were trialed ranging from 2 to 7. Each of these settings were trialled across a range of clustering
+algorithms including AGNES hierarchical agglomerative [@kaufman2009finding], Kmeans [@hartigan1979algorithm] and self-organising maps (SOM) [@kohonen2012self].
 
-\[NEED TO INCLUDE CONTRIBUTION\]
+The results for each method were evaluated using three internal cluster validation measures, Connectivity [@handl2005computational], Sihouette width [@rousseeuw1987silhouettes] and Dunn Index [@dunn1974well].
 
-Multiple Correspondence Analysis was then computed using the *FactoMineR* R package [@factominer].
+The optimal method and cluster number was selected based on a trade off between the best performing method using the evaluation metrics
+and a sensible partioning of the data for this analysis.
 
-### Cluster Analysis
-
-\[insert clvalid details\]
-
-### Survival Analysis
+## Survival Analysis
 
@@
 
-The estimator is calcualted as:
+The estimator is calculated as:
 
@@
 
-## Results
+# Results
 
-### MCA
+## MCA
 
@@
 Repeating MCA for the remaining subjects shows the first dimension ($\lambda_!
-= 0.18$) accounts for 17.88% of the variance in the data with the first 4 dimensions of the MCA analysis accounting for 54% of the variance.
+= 0.18$) accounts for 17.88% of the variance in the data with the first 4 dimensions of the MCA analysis accounting for 54% of the variance (@tbl-eig).
 
@@
 ```{r}
+#| label: tbl-eig
+#| tbl-cap: Eigenvalues and percentage of variance explained by the MCA principal axes
+
 ### MCA  
@@
 
-# eigenvalue summary 
 mca$eig |> 
-  kable(digits = 2, caption = "eigenvalues and percentage of variance explained by the MCA principal axes")
+  tibble::as_tibble(rownames = 'dimension') |> 
+  gt() |> 
+  fmt_number(columns = 2,
+             decimals = 2) |> 
+  fmt_percent(columns = c(`percentage of variance`, `cumulative percentage of variance`),
+              scale_values = F) |> 
+   tab_header(
+    title = "MCA Eigenvalue Table",
+    subtitle = "Percentage variance explained by MCA for each dimension"
+  )
 ```
@@
 
-### Cluster Analysis
+## Cluster Analysis
 
@@
 ```{r}
-cluster_select@measures |> 
-  as_tibble(rownames = 'measure') |> 
-  pivot_longer(cols =  -measure, names_to = 'cluster', values_to = "metric") |> 
-  mutate(`optimal cluster no.` = as.numeric(str_remove(cluster, ".hierarchical")))  |> 
-  select(-cluster) |> 
-  group_by(measure) |> 
-  filter(metric == min(metric)) |> 
-  kable(digits = 2, caption = "Optimal cluster numbers determined by four similar metrics measuring cluster analysis performance.")
+#| label: tbl-opticlust
+#| tbl-cap: "Optimal clusters"
+optimalScores(cluster_select) |> 
+  as_tibble(rownames = 'Metric') |> 
+  gt() |> 
+  fmt_number(columns = Score, decimals = 2)
 ```
@@
 The most optimal number of clusters are selected by picking the lowest value from one or many of these compariable metrics."
+
 cluster_select@measures |>
@@
   pivot_longer(cols =  -measure, names_to = 'cluster', values_to = "metric") |>
-  mutate(cluster = as.numeric(str_remove(cluster, ".hierarchical"))) |> 
-  ggplot(aes(cluster, metric, colour = measure, group = measure)) +
+  separate(col = cluster, into = c('k', 'method'), sep = "\\.") |> 
+  mutate(k = as.numeric(k)) |>
+  ggplot(aes(k, metric, colour = method)) +
   geom_line() +
@@
 
-```{r, fig.height=12, fig.width=10}
-#| label: fig-dendro
-#| fig-cap: "Dendrogram of heirarchical clustering method"
-plot(clustering, which.plots = 2, main = NA, sub = NA, xlab = NA)
-```
-
 ```{r, fig.height=8, fig.width=8,  warning=FALSE}
@@
 #| fig-cap: "A projection of features on the first two principal components with cluster assignment indicated by colour."
-cbind(cluster_data, clust = cluster_assignments) |> 
+cbind(cluster_data, clust = factor(cluster_assignments)) |> 
   rownames_to_column(var = 'var') |> 
@@
 
-### Survival Analysis
+## Survival Analysis
 
@@
 
-```{r, fig.height=8, fig.width=8,  warning=FALSE}
+```{r, warning=FALSE}
 #| label: fig-surv
@@
 
-```{r, fig.height=8, fig.width=8,  warning=FALSE}
+```{r, warning=FALSE}
 # KM by variant
@@
 
-## Discussion
-
--   the relationship between the results and the original hypothesis, i.e., whether they support the hypothesis, or cause it to be rejected or modified
+# Discussion
 
@@
 
-## Appendix
+# Appendix
 
 ```{r}
-16/28
+#| label: tbl-dict
+#| tbl-cap: Data dictionary of the final dataset used for the analysis.
+This combines the case reports and supplied data for available and references studies.
+The data has been transformed into a tidy format with formatting applied.
+tar_load(subjects)
+make_data_dictionary(subjects)
+```
+
+```{r}
 # Profile table 
@@
 
-## References
+# References

### 2022-09-05 R/utils.R

@@ -14,3 +14,11 @@ scale_y_reordered <- function(..., sep = "___") {
   reg <- paste0(sep, ".+$")
   ggplot2::scale_y_discrete(labels = function(x) gsub(reg, "", x), ...)
 }
+
+make_data_dictionary <- function(data){
+  dd <- data.frame(variable = names(data), 
+                   type = unlist(lapply(data, class), use.names = FALSE), 
+                   description = ".")
+  gt::gt(dd) 
+}
+

### 2022-09-05 _targets.R

@@ -3,7 +3,9 @@ library(tarchetypes)
 
 # Set target options
 tar_option_set(
-  packages = c("SMPD4", 'clValid', 'tidyverse', 'ggrepel', 'corrplot', 'cluster', 'survival', 'survminer', 'quarto', 'knitr', 'FactoMineR', 'factoextra'), 
+  packages = c("SMPD4", 'clValid', 'tidyverse', 'ggrepel', 
+               'cluster', 'survival', 'survminer', 'quarto',
+               'knitr', 'FactoMineR', 'factoextra', 'gt', 'withr'), 
   format = "rds" 
 )
 
@@ -24,13 +26,14 @@ list(
   tar_target(mca,  MCA(mca_data, graph = FALSE)),
   
   ## cluster analysis 
-  tar_target(cluster_data, prep_cluster_data(mca, cols = c('dim_1', 'dim_2'))),
+  tar_target(cluster_data, prep_cluster_data(mca, cols = c('dim_1', 'dim_2', 'dim_3', 'dim_4'))),
   tar_target(cluster_select, clValid(cluster_data,
-                                     nClust = 2:10, 
-                                     clMethods = "hierarchical", 
-                                     validation = "internal")),
-  tar_target(clustering, agnes(cluster_data)),
-  tar_target(cluster_assignments, factor(cutree(clustering, 6))),
+                                     nClust = 2:5, 
+                                     clMethods = c("agnes", "kmeans", "som"), 
+                                     method = "average", 
+                                     validation = "stability")),
+  tar_target(clustering, withr::with_seed(seed = 1234, kmeans(cluster_data, centers = 5, nstart = 10))),
+  tar_target(cluster_assignments, clustering$cluster),
   
   ## Survival Analysis
   tar_target(survival_data, prep_survival_data(data = subjects,
@@ -41,7 +44,8 @@ list(
   tar_target(cox_ph, coxph(Surv(survival_time, deceased) ~ variant, data = survival_data)),
 
   # Render Report
-  tar_quarto(report, path = "paper_draft.qmd")
+  tar_quarto(report, path = "paper.qmd")
 )
+
 
 

### 2022-09-05 _targets_packages.R

@@ -2,10 +2,13 @@
 library(bs4Dash)
 library(cluster)
 library(clustermq)
-library(corrplot)
+library(clValid)
+library(factoextra)
+library(FactoMineR)
 library(future)
 library(ggrepel)
 library(gt)
+library(knitr)
 library(markdown)
 library(pingr)
 library(quarto)

### 2022-09-05 run.R

@@ -1,2 +1,1 @@
 targets::tar_make()
-targets::tar_visnetwork() 

### 2022-09-05 paper.qmd
updates to draft

@@
 abstract: "A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*.
-The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and diverse clinical phenotype.
-A gap exists in the understanding the associations of the heterogenous features present in the clinical phenotype and the expected survival probabilities of affected individuals, driven in part to the paucity of analysis-ready data.
+The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and clinically diverse phenotype.
+A gap exists in the understanding the associations of the heterogenous features present in the clinical phenotype and the expected survival probabilities of affected individuals.
+This is driven in part to the paucity of analysis-ready data in reported cases.
 This research aims to collate and standardise available case reports to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA.
-To overcome the challenge of sparse, multidimensional data on very few subjects, we employ Multiple Correspondence Analysis as a dimension reduction technique, which is then subject to heirarchical clustering and interpretation.
+To overcome the challenge of sparse, multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation.
 To quantify more accurate survival probabilities, Kaplan-Meier estimation is formulated to account for censoring in the data.
@@
 However, interpretation of these results are guarded based on very low sample sizes.
-Ultra rare conditions present a challenge for clinicians and families, with often limited information and diverse possible outcomes.
-Despite challenges with sparse and inconsistent data, this analysis helps describe associations and trajectories of indivudals with this condition and can provide clinicians and genetic counsellors with better information to aide in decision making and support for families."
+Despite challenges with sparse and inconsistent data, this analysis represents the first of its kind to help describe associations and trajectories of individuals with this condition and can provide clinicians and genetic counsellors with better information to aide in decision making and support for families."
 editor: visual
@@
 A detailed study by @magini2019loss involved 12 unrelated familes with 32 individuals (21 with detailed clinical information).
-Presentations were of microcephaly, simplified gyration, hypomyelination, thin corpus callosum, mild cerebellar hypoplasia, brainstem hypoplasia, congenital arthrogryposis, diabetes mellitus, heart disease, severe encephalopathy and respiratory problems often leading to early demise.
+Presentations were of microcephaly, simplified gyral pattern of the cortex, hypomyelination, cerebellar hypoplasia, congenital arthrogryposis, and early fetal/postnatal demise.
 Despite this being the largest cohort studied, the clinical features and survival times among participants varied greatly.
@@
 A recent study by @bijarnia2022growth presents the case of a 22-month old girl presenting with the typical phenotype of neurodevelopmental delay, prenatal onset growth failure, arthrogryposis, microcephaly and brain anomalies including severe hypomyelination, simplified gyral pattern and hypoplasia of corpus callosum and brainstem.
-Notably, there is also additional non-typical clinical findings of nystagmus and visual impairment secondary to macular dystrophy and retinal pigment epithelial stippling at posterior pole.
+Notably, there was also additional non-typical clinical findings of nystagmus and visual impairment secondary to macular dystrophy and retinal pigment epithelial stippling at posterior pole.
 
@@
 
-WIn terms of patient survival, methods in statistical survival analysis are well suited and commonly used in the field of genomics [@chen2014survival].
+In terms of patient survival, methods in statistical survival analysis are well suited and commonly used in the field of genomics [@chen2014survival].
 These methods in particular provide a mechanism for dealing with right-censoring, a common feature of clinical studies where the clinical outcome of interest may not be known by the end of the study period.
-Within the context of examining a single condition, @Crowe810 investigated comorbidity phenotypes and mortality risk in ischaemic heart disease patients.
-Here latent class analysis was performed to identify a small number of clusters in the patients which could then be included in survival analysis techniques to better understand the comparative outcomes of patients of this condition.
 
@@
 With current research highlighting a diverse and heterogeneous phenotype, the relationship and correlations between these diverse features is not well understood.
-Furthermore, the survival time of affected children is not well understood despite early-demise featuring as a severe outcome.
+Furthermore, the survival time of affected children is not well understood despite early-demise featuring as a severe outcome in many cases.
 While a connection has been highlighted between some missense variants and a milder presentation with longer survival [@magini2019loss], this hypothesis has not been analysed further.
@@
 
-This will be the first in-depth analysis of early studies of this novel variant, which aims to produce statistical findings to better describe and understand this condition.
+This will be the first in-depth analysis of early studies of this novel variant, which aims to produce statistical findings to better understand this newly described condition.
 This research will assist clinicians understand the typical and non-typical presentations of this challenging new condition.
@@
 
-The data were transposed to ensure it could be presented as *tidy* formatted data [@JSSv059i10]. This requires:
-
-1.  Each variable (clinical phenotype) forms a column.
-2.  Each observation (individual) forms a row.
-3.  Each type of observational unit forms a table (every cell has just one item)
+The data were transposed to ensure it could be presented as *tidy* formatted data [@JSSv059i10].
+This requires each variable (clinical phenotype) forms a column; each observation (individual) forms a row and each type of observational unit forms a table (every cell has just one item).
 
@@
 | id                     | short palpebral fissures | large ears | simple helices | receding forehead | ... |
-|------------------------|-----------------|-------|--------|-------|-------|
-| Family 1- Individual 1 | 1                        | 1          | 0              | ...               |     |
+|-----------------|-----------|-----------|-----------|-----------|-----------|
+| Family 1- Individual 1 | 1                        | 1          | 1              | 0                 | ... |
 | Family 1- Individual 4 | 1                        | 0          | 0              | 1                 | ... |
@@
 
-Some data type conversion and categorical level standardisation was performed to ensure the data were in consistent and appropriate data types.
+Data type conversion and categorical level standardisation was performed to ensure the data were in consistent and appropriate data types.
 For example, 'Gender' was not consistently coded, and 'Birth Weight' was encoded as a text string rather than a more useful numeric format.
@@
 
-The final dataset consisted of 21 observations (one per individual) and 152 variables (one per clinical feature).
-In some cases, were obvious, these variables were left as qualitative character descriptions, in other cases as noted above, distinct lists of features were tranformed into binary indicators.
-Other fields include background information on the patient and various biomarkers.
+The final dataset consisted of 28 observations (one per individual) and 152 variables (one per clinical feature).
+These variables were comprised of qualitative text descriptions, binary indicators of features as described above and other fields including background information on the patient and various biomarkers.
 A full list of this data set is available in @tbl-dict.
-These data sets were packaged into an R Package called SMPD4 [@smpd4-data] in order to allow for reproducibility and sharing.
+These data sets were compiled into a publicly available R Package called SMPD4 [@smpd4-data] in order to allow for reproducibility and sharing.
 
@@
 
-Dimension reduction on this wide data set was performed using Multiple Correspondence Analysis (MCA) [@le2010multiple] in order to arrive at a set of meaningfully small dimensions that account for most of the variation in the data.
-Multiple Correspondence Analysis (MCA) is an analogy to Principle Component Analysis (PCA) for categorical data.
+Dimension reduction on this wide data set was performed using Multiple Correspondence Analysis (MCA) (an analogy to Principle Component Analysis (PCA) for categorical data) [@le2010multiple] in order to arrive at a set of meaningfully small dimensions that account for most of the variation in the data.
 
@@
 Ctr_k = \frac{py_k^2}{\lambda}
-$$
-Where $p$ represents the point weighting, $y$ is the coordinate relative to the principal axes of variance $\lambda$.
+$$ Where $p$ represents the point weighting, $y$ is the coordinate relative to the principal axes of variance $\lambda$.
+
+The coordinates on the first few principal axes were inspected for any outliers or unusual data.
 
@@
 
-The MCA coordinates for the first 4 principal dimensions were subject to clustering in order to find groups of features
-that were internally homogeneous and externally heterogeneous. This was computed in R using the clValid package [@clvalid].
+The MCA coordinates for the first 4 principal dimensions were subject to clustering in order to find groups of features that were internally homogeneous and externally heterogeneous.
+This was computed in R using the clValid package [@clvalid].
 
-A sensible range of clusters were trialed ranging from 2 to 7. Each of these settings were trialled across a range of clustering
-algorithms including AGNES hierarchical agglomerative [@kaufman2009finding], Kmeans [@hartigan1979algorithm] and self-organising maps (SOM) [@kohonen2012self].
+A range of clustering algorithms were comparied including AGNES hierarchical agglomerative [@kaufman2009finding], k-means [@hartigan1979algorithm] and self-organising maps (SOM) [@kohonen2012self].
+A sensible range of clusters from 2-5 were trialed for each clustering algorithm.
+The results for each method were evaluated using three measures of stability: Average proportion of non-overlap (APN), Average Distance (AD) and Average Distance Between Means (ADM) [@datta2003comparisons].
+Stability measures compare the full clustering result with a result based on dropping meach of the columns one at a time.
+These metrics were selected as we are representing each data point (clinical feature) with a number of MCA dimensions.
+A good cluster solution is one that is robust and meaningful across all of these dimensions [@fig-clvalid].
 
-The results for each method were evaluated using three internal cluster validation measures, Connectivity [@handl2005computational], Sihouette width [@rousseeuw1987silhouettes] and Dunn Index [@dunn1974well].
-
-The optimal method and cluster number was selected based on a trade off between the best performing method using the evaluation metrics
-and a sensible partioning of the data for this analysis.
+The optimal method and cluster number was selected based on a trade off between the best performing method using the evaluation metrics and a sensible partioning of the data for this analysis [@tbl-eig].
 
@@
 To model survival probability for all subjects in the combined data, the data are subsetted to include *survival_time* which is either the number of days the individual survived for, or the age in days at last follow up, and *deceased* a numeric indicator which is equal to 1 if the subject is deceased and 0 otherwise.
-
 The Kaplan-Meier estimator [@kaplan1958nonparametric] is used to estimate the survival function of the data.
@@
 The baseline estimator of all individuals was calculated using the *survival* package in R [@survival-book].
-
-Next the Kaplan-Meier estimator was calculated, \[ON WHAT?\] stratified by the reported type of Single nucleotide polymorphisms (SNP) reported in the literature.
-A test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* package.
+Next the Kaplan-Meier estimator was calculated, stratified by the reported type of genetic variation reported in the literature.
+A logrank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* R package.
 
@@
 
-The initial results of the MCA were inspected with a bi-plot and identified two individuals (17, 18) from the data that were clear outliers and sat significatly beyound the 95% confidence ellipse.
-These individuals represented two twins from @magini2019loss who were described with a much milder phenotype and wer subject to a missense variation in SMPD4.
+The initial results of the MCA were inspected with a bi-plot and identified two subjects (17, 18) from the data that were clear outliers and sat significatly beyound the 95% confidence ellipse.
+These individuals represented two twins from @magini2019loss who were described with a much milder phenotype and were subject to a missense variation in *SMPD4*.
 Further analysis of this influence is conducted below.
@@
 
-```{r fig.height=8, fig.width=8,  warning=FALSE}
+```{r fig.height=5, fig.width=5,  warning=FALSE}
 #| label: fig-biplotprelim
-#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables"
+#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables.
+A 95% confidence ellipse is plotted in red around the origin."
 fviz_mca_biplot(
@@
   ellipse.level=0.95,
+  select.var = list(name = NULL, cos2 = 0.2, contrib = NULL),
   palette = "grey",
@@
   fmt_percent(columns = c(`percentage of variance`, `cumulative percentage of variance`),
-              scale_values = F) |> 
-   tab_header(
-    title = "MCA Eigenvalue Table",
-    subtitle = "Percentage variance explained by MCA for each dimension"
-  )
+              scale_values = F)
 ```
@@
 The *contribution* of the categories were ranked within each of the first four MCA dimensions See @fig-ctr.
-
-The first dimension is dominated by highly specific facial dysmophisms such as epicanthal folds, short philtrum and low anteriour hairline. Also significant are seizure and developmental delay.
+The first dimension is dominated by highly specific facial dysmophisms such as epicanthal folds, short philtrum and low anteriour hairline.
 
@@
 
-The third dimension is characterised by complex non-typical features such as patent ductus arteriosus, respiratory infections, small for gestational age and further facial features.
+The third dimension is characterised by complex features such as patent ductus arteriosus, respiratory infections, small for gestational age and further facial features.
 These are contrasted by the fourth principal dimension which is typical for respiratory distress including stridor and hypercapnia.
 
-```{r ctr-plot, fig.height=10, fig.width=10, warning=FALSE}
+```{r ctr-plot, fig.height=5, fig.width = 8, warning=FALSE}
 #| label: fig-ctr
-#| fig-cap: "Top 20 feature categories from MCA analysis for the first four MCA dimensions.
+#| fig-cap: "Top 10 feature categories from MCA analysis for the first four MCA dimensions.
 The red dashed line indicates the mean contribution value.
@@
   guides(fill=guide_legend(title="Condition")) +
+  theme(legend.position = 'bottom') +
   labs(y = '')
@@
 
-```{r fig.height=8, fig.width=8,  warning=FALSE}
+```{r fig.height=5, fig.width=5,  warning=FALSE}
 #| label: fig-biplot
-#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables"
+#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables.
+A 95% confidence ellipse is plotted in grey around the origin."
 fviz_mca_biplot(
@@
   ellipse.level=0.95,
+  select.var = list(name = NULL, cos2 = 0.4, contrib = NULL),
   palette = "grey",
@@
 
-```{r fig.height=8, fig.width=8,  warning=FALSE}
+```{r fig.height=5, fig.width=5,  warning=FALSE}
 #| label: fig-biplot2
-#| fig-cap: "A bi-plot of principal axes 3 and 4, with grey cicles indicating the individual's projected position and black triangles indicaating the variables"
+#| fig-cap: "A bi-plot of principal axes 3 and 4, with grey cicles indicating the individual's projected position and black triangles indicaating the variables.
+A 95% confidence ellipse is plotted in grey around the origin."
 fviz_mca_biplot(
@@
 
-A cluster relating to structual brain anomalies is articulated with conditions such as cerebellar hypoplasia and agenesis of corpus collosum, along with small for gestational age.
-
 Some smaller satellite clusters exists where highly distinctive features were associated with an indivual or family.
 Also a singleton clusters exist for features that are distinct or unusual such as posteriorly angulated ears.
+
+A full articulation of feature to cluster assignment is given in @tbl-profile.
 
@@
 
-```{r, fig.width=8,  warning=FALSE}
+```{r, warning=FALSE}
 #| label: fig-clvalid
@@
 
-```{r, fig.height=8, fig.width=8,  warning=FALSE}
+```{r, fig.height=6, fig.width=6,  warning=FALSE}
 #| label: fig-clust
-#| fig-cap: "A projection of features on the first two principal components with cluster assignment indicated by colour."
-cbind(cluster_data, clust = factor(cluster_assignments)) |> 
+#| fig-cap: "A projection of features on the first two principal components with k-means cluster assignment indicated by colour."
+
+d <- cbind(cluster_data, clust = factor(cluster_assignments)) |> 
   rownames_to_column(var = 'var') |> 
   as_tibble() |> 
-  arrange(var, clust) |> 
-  ggplot(aes(dim_1, dim_2, colour = clust)) +
+  arrange(var, clust)
+
+hull <- d %>%
+  group_by(clust) |> 
+  slice(chull(dim_1, dim_2))
+
+  ggplot(d, aes(dim_1, dim_2, colour = clust)) +
+  geom_polygon(data = hull, alpha = 0.2, aes(fill = clust)) +
   geom_vline(xintercept = 0, lty = 2, colour = "grey50") +
   geom_hline(yintercept = 0, lty = 2, colour = "grey50") +
-  geom_label_repel(aes(label = var), size = 2, max.overlaps = 20) +
-  geom_point(aes(colour = clust), position = "jitter") +
+  geom_label_repel(data = d |> group_by(clust) |> slice_sample(n=4), aes(label = var), size = 3, show.legend = F) +
+  geom_point(aes(colour = clust), position = "jitter", show.legend = F) +
   labs(x = "Dimension 1",
        y = "Dimension 2",
-       colour = 'cluster')
+       fill = 'cluster') +
+    guides(color='none')
 ```
@@
 #| label: fig-surv
-#| fig-cap: "Kaplan Meier survival curve"
+#| fig-cap: "Kaplan Meier curve with the black line representing the survival probability and with grey shading indicating the 95% confidence interval."
 survminer::ggsurvplot_df(x,
@@
 However, even within missense hymozygous variants the survival time varies greatly.
-Other missense variants, such as heterozygous missense with gene deletion do not show the same increased survival time as indeed noted by @magini2019loss.
+Other missense variants, such as heterozygous missense with gene deletion do not show the same increased survival time as noted by @magini2019loss.
 
-```{r, warning=FALSE}
+```{r,fig.width=7, warning=FALSE}
 # KM by variant
@@
 ggplot(survival_data, aes(survival_time, variant,fill = variant_type.x)) +
-  geom_boxplot()
+  geom_boxplot() +
+  theme(legend.position = 'bottom') +
+  labs(fill = 'Variant Type',
+       x = 'Survival Time')
 ```
 
-The log rank test was carried out to compare the suvival curves of the various gene variants $\chi^2 = 14.8$ $(p = 0.04)$ however due to the small sample size, several variants had expected counts less than 5 which would create a violation in assumptions for the test.
+The logrank test was carried out to compare the suvival curves of the various gene variants $\chi^2 = 14.8$ $(p = 0.04)$ however due to the small sample size, several variants had expected counts less than 5 which would create a violation in assumptions for the test.
 
@@
 tar_load(logrank)
-logrank
 ```
@@
 
-Further distictions are possible with more specific structural brainanomalies exhibiting as distinct sources of variation in the clininal phenotype.
-
 Many of the outlying features identifes are specific facial or anatomical dysmophims.
 While a noted and core feature is around course facial features, those with a milder course had clinical reports in the available data that articulated these dysmophims in much more detail.
-
 On a similar note, the reporting of 'talipes' as acondition seems distict as a source of variation when appearing as an isolated feature without the typical microcephasly and hypomyelination.
@@
 
-The cluster analysis does not aim to group individuals that are alike, but rather the clinical features themselves.
+The cluster analysis does not aim to group individuals that are alike, but rather the clinical features themselves and how these contrasting clusters account for variation in the data.
 This highlights common sources of variation on the clinical phenotype of the condition itself, rather than typical coincident features of any one child.
-While the number of clusters chosen all represent valid partition of the data, a larger number was selected to align the goals of the analysis which was to uncover distinct groups.
-A smaller number would have likely yeilded isolated outliers from the main body of the features.
-This still occurs, however more substructure is detected with the higher cluster number.
+While the number of clusters chosen all represent valid partition of the data, some metrics indicated an optimal selection of 2 clusters.
+This type of partition is useful when segregating outliers from the rest of the data and is valid in some application.
+In this study, a larger number was selected to align the goals of the analysis which was to uncover distinct groups of features.
 
@@
 A small number of cases show much longer survival into the childhood and in some cases into the second decade of life.
-The current literature suggests the type of gene variant has an impact on surivival time, which seems to be supported in this work, however the limited sample size here prevents other meaningful analysis such as logrank tests or cox proprtional hazard modelling which are likely underpowered when comparing multiple groups.
+This work identifies a one year survival rate of 0.403 (0.252, 0.644).
+A naive estimate, using just those number of deceased patients as proportion of the study population is 0.57.
+This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to erroneously overestimate the survival probability.
+The current literature suggests the type of gene variant has an impact on surivival time, which seems to be supported in this work through a comparison of multiple survival surves, stratified by variation type, however the limited sample size here prevents other meaningful analysis such as logrank tests or cox proprtional hazard modelling which are likely underpowered when comparing multiple groups.
 
@@
 Furthermore it has isolated additional distinct groupings of features that are exhibited in patients such as vocal cord paralysis with tracheostomy, swallowing dysfunction, tube feeding and GERD.
-In additional to typical MRI findings of symplified gyral pattern and hypomyelination, there exists clusters of structural anomalies of the corpus calllosum and cereballa hypoplasia.
-
-The survival of indivuduals with this condition is critical given early demise in infancy and childhood is a noted feature.
-This work identifies the survival probabilites and shows a one year survival rate of 0.403 (0.252, 0.644).
-A naive estimate, using just those number of deceased patients as proportion of the study population is 0.57.
-This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to overestimate the survival probability.
+Furthermore, we were able to quanitfy the expected baseline survival probabilities for children with this condition as well as quantifying the significant variation in these estimates.
+Additional work is required with the benefit of more cases, to better understand the differences in survival curves for other factors associated with genotype.
 
@@
 ```{r}
-# Profile table 
-cbind(cluster_data, clust = cluster_assignments) |> 
-  rownames_to_column(var = 'var') |> 
+#| label: tbl-profile
+#| tbl-cap: Assignment of clinical features with their cluster.
+These clusters represent similar features projected onto the first four MCA principal axes.
+Clusters that are on opposing sides of an axes contract with each other and account for the largest amount of variation in the data.
+cbind(cluster_data, cluster = cluster_assignments) |> 
+  rownames_to_column(var = 'variable') |> 
   as_tibble() |> 
-  select(clust, var) |> 
-  arrange(clust) |> 
+  select(cluster, variable) |> 
+  arrange(cluster) |> 
   kable()

### 2022-09-05 R/utils.R

@@ -17,8 +17,7 @@ scale_y_reordered <- function(..., sep = "___") {
 
 make_data_dictionary <- function(data){
   dd <- data.frame(variable = names(data), 
-                   type = unlist(lapply(data, class), use.names = FALSE), 
-                   description = ".")
+                   type = unlist(lapply(data, class), use.names = FALSE))
   gt::gt(dd) 
 }
 

### 2022-09-07 paper.qmd
draft review

@@
 To quantify more accurate survival probabilities, Kaplan-Meier estimation is formulated to account for censoring in the data.
-The analysis identified the classic phenotype for this condition, as well as two other distinct feature clusters.
-The first relates to findings of vocal cord paralysis and swallowing dysfunction, the other relates to more complex brain anomalies.
-Furthermore, the survival probability for those affected was found to decline sharply in early infancy, but exhibited a wide range of outcomes provisionally associated with variant type.
-However, interpretation of these results are guarded based on very low sample sizes.
+The analysis identified the classic phenotype for this condition, as well as one other distinct feature cluster relating to findings of vocal cord paralysis and respiratory failure.
+The survival probability for those affected was found to decline sharply in early infancy with median survival of 150 days, but with some as high as 12.5 years.
+This wide range of outcomes is provisionally associated with variant type however interpretation of these results are guarded based on very low sample sizes.
 Despite challenges with sparse and inconsistent data, this analysis represents the first of its kind to help describe associations and trajectories of individuals with this condition and can provide clinicians and genetic counsellors with better information to aide in decision making and support for families."
@@
 # table with med gen 
-# Better tables with gtsummary
 ```
@@
 
-Neurodevelopment disorders are a diverse and heterogeneous group of conditions impacting the development of the nervous system, brain function, physical development, emotional development and learning ability.
+Neurodevelopment disorders are a diverse and heterogeneous group of conditions impacting the development of the nervous system, brain function, physical development and learning ability.
 A recently identified condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) (MIM:[618622](https://omim.org/entry/618622)) has been described in children with bi-allelic loss-of-function variants in *SMPD4* [@magini2019loss].
-Sphingomyelinases such as *SMPD4* play an important celluar role by hydrolyzing sphingomyelin into ceramide and phosphorylcholine.
+Sphingomyelinases such as *SMPD4* play an important cellular role by hydrolyzing sphingomyelin into ceramide and phosphorylcholine.
 *SMPD4* specifically encodes one of 4 neutral sphinogmyelinases, nSMase3 (MIM: [610457](http://omim.org/entry/610457)).
@@
 
-A detailed study by @magini2019loss involved 12 unrelated familes with 32 individuals (21 with detailed clinical information).
+A detailed study by @magini2019loss involved 12 unrelated families with 32 individuals (21 with detailed clinical information).
 Presentations were of microcephaly, simplified gyral pattern of the cortex, hypomyelination, cerebellar hypoplasia, congenital arthrogryposis, and early fetal/postnatal demise.
@@
 To overcome these data challenges, @JSSv059i10 proposes a data structure know as 'tidy' data which organises each observation in a row, with each feature as its own column and each value in just one cell.
-In this case detailed text-based data can be tranformed to high dimensional binary indicators of key clinical features.
+In this case detailed text-based data can be transformed to high dimensional binary indicators of key clinical features.
 This structure is conducive to effective statistical analysis and integrates deliberately with the *tidyverse* [@tidyverse] collection of packages for data analysis in R [@rbase].
@@
 
-An open research question exists around the clinical pathway and survival time of children exhibiting bi-allelic loss-of-function variants in SMPD4.
+An open research question exists around the clinical pathway and survival time of children exhibiting bi-allelic loss-of-function variants in *SMPD4*.
 With current research highlighting a diverse and heterogeneous phenotype, the relationship and correlations between these diverse features is not well understood.
-Furthermore, the survival time of affected children is not well understood despite early-demise featuring as a severe outcome in many cases.
+Furthermore, the survival time of affected children has not been extensively analysed despite early-demise featuring as a severe outcome in many cases.
 While a connection has been highlighted between some missense variants and a milder presentation with longer survival [@magini2019loss], this hypothesis has not been analysed further.
@@
 
-This will be the first in-depth analysis of early studies of this novel variant, which aims to produce statistical findings to better understand this newly described condition.
-This research will assist clinicians understand the typical and non-typical presentations of this challenging new condition.
+This will be the first in-depth analysis of this novel variant, which aims to produce statistical findings to better understand this newly described condition.
+This research will assist clinicians understand the various presentations of this challenging new condition.
 In addition, genetic counselling of families with affected children will benefit from enhanced analysis on outcomes of existing reported cases.
@@
 
-The data were read into R [@rbase] without any changes, so as to preserve the reproducibility of the data transformation steps.
-
+The data were read into `R` [@rbase] without any changes, so as to preserve the reproducibility of the data transformation steps.
 The data were transposed to ensure it could be presented as *tidy* formatted data [@JSSv059i10].
 This requires each variable (clinical phenotype) forms a column; each observation (individual) forms a row and each type of observational unit forms a table (every cell has just one item).
-
 This resulted in a long-formatted data set of 21 rows (individuals) and 64 columns (clinical features).
-
 In many cases, key clinical information was entered as free-text descriptions, which rendered any attempt of meaningful analysis impractical (@tbl-nontidy).
@@
 |                     | Family 1- Individual 1                                                                                     | Family 1- Individual 4                                      |
-|------------------|--------------------------|----------------------------|
+|---------------|-------------------------------------|--------------------|
 | Facial dysmorphisms | short palpebral fissures, large ears, simple helices, smooth philtrum, thin lips, bilateral simian creases | short palpebral fissures, receding forehead, thin upper lip |
@@
 | id                     | short palpebral fissures | large ears | simple helices | receding forehead | ... |
-|-----------------|-----------|-----------|-----------|-----------|-----------|
+|------------|-------------|------------|------------|------------|------------|
 | Family 1- Individual 1 | 1                        | 1          | 1              | 0                 | ... |
@@
 
-In the case where two variables were formed from phenotypes that are considered to be synonymous, these were merged into one indicator column to prevent duplication e.g.
+In the case where two variables were formed from features that are considered to be synonymous, these were merged into one indicator column to prevent duplication e.g.
 {bilateral_cleft_lips, bilateral_cleft_lip, cleft_lip_b\_l} -\> {bilateral_cleft_lip}.
@@
 
-The final dataset consisted of 28 observations (one per individual) and 152 variables (one per clinical feature).
-These variables were comprised of qualitative text descriptions, binary indicators of features as described above and other fields including background information on the patient and various biomarkers.
+The final data set consisted of 28 observations (one per individual) and 152 variables (one per clinical feature).
+These variables were comprised of qualitative text descriptions, binary indicators of features as described above and other fields including background information on the patient and various bio markers.
 A full list of this data set is available in @tbl-dict.
-These data sets were compiled into a publicly available R Package called SMPD4 [@smpd4-data] in order to allow for reproducibility and sharing.
+These data sets were compiled into a publicly available `R` Package called `SMPD4` [@smpd4-data] in order to allow for reproducibility and sharing.
 
@@
 The sum of all categories $K = \sum_{q=1}^{Q}K_q$ defines the dimensionality of $\mathbf{X}$ as an $I\times K$ matrix.
-This resulted in a dataset $X_{28 \times 61}$ indicator matrix of 61 clinical features across 28 individuals.
-Taking $\delta_{ik} = 1$ if subject $i$ has feature $k$ and $\delta_{ik} = 0$ if the subject does not, we are left with the completely disjnuctive table $\mathbf{X} = I\times K$ of $\{0,1\}$.
+This resulted in a data set $X_{28 \times 61}$ indicator matrix of 61 clinical features across 28 individuals.
+Taking $\delta_{ik} = 1$ if subject $i$ has feature $k$ and $\delta_{ik} = 0$ if the subject does not, we are left with the completely disjunctive table $\mathbf{X} = I\times K$ of $\{0,1\}$.
 Letting the sum of all entries of $\mathbf{X}$ be $N$, we have $\mathbf{Z} = N^{-1}\mathbf{X}$.
-We can introduce two diagonal matricies $\mathbf{D_r} = diag(\mathbf{r})$ and $\mathbf{D_c} = diag(\mathbf{c})$ where $\mathbf{r}$ and $\mathbf{c}$ are the vectors of row sums and column sums of $\mathbf{Z}$ respectively.
+We can introduce two diagonal matrices $\mathbf{D_r} = diag(\mathbf{r})$ and $\mathbf{D_c} = diag(\mathbf{c})$ where $\mathbf{r}$ and $\mathbf{c}$ are the vectors of row sums and column sums of $\mathbf{Z}$ respectively.
 
@@
 
-The *contribution* of each category is computed for each of the principal axis per @le2010multiple. This metric identifies the proportion of variance of that axis due to the point and is defined as:
+The *contribution* of each category was computed for each of the principal axis per @le2010multiple. This metric identifies the proportion of variance of that axis due to the point and is defined as:
 
@@
 The MCA coordinates for the first 4 principal dimensions were subject to clustering in order to find groups of features that were internally homogeneous and externally heterogeneous.
-This was computed in R using the clValid package [@clvalid].
+This was computed in R using the `clValid` package [@clvalid].
 
-A range of clustering algorithms were comparied including AGNES hierarchical agglomerative [@kaufman2009finding], k-means [@hartigan1979algorithm] and self-organising maps (SOM) [@kohonen2012self].
+A range of clustering algorithms were compared including AGNES hierarchical agglomerative [@kaufman2009finding], k-means [@hartigan1979algorithm] and self-organising maps (SOM) [@kohonen2012self].
 A sensible range of clusters from 2-5 were trialed for each clustering algorithm.
 The results for each method were evaluated using three measures of stability: Average proportion of non-overlap (APN), Average Distance (AD) and Average Distance Between Means (ADM) [@datta2003comparisons].
-Stability measures compare the full clustering result with a result based on dropping meach of the columns one at a time.
-These metrics were selected as we are representing each data point (clinical feature) with a number of MCA dimensions.
+Stability measures compare the full clustering result with a result based on dropping each of the columns one at a time.
+These metrics were selected to represent each data point (clinical feature) with a number of MCA dimensions.
 A good cluster solution is one that is robust and meaningful across all of these dimensions [@fig-clvalid].
-
-The optimal method and cluster number was selected based on a trade off between the best performing method using the evaluation metrics and a sensible partioning of the data for this analysis [@tbl-eig].
+The optimal method and cluster number was selected based on a trade off between the best performing method using the evaluation metrics and a sensible partitioning of the data for this analysis [@tbl-eig].
 
@@
 
-To model survival probability for all subjects in the combined data, the data are subsetted to include *survival_time* which is either the number of days the individual survived for, or the age in days at last follow up, and *deceased* a numeric indicator which is equal to 1 if the subject is deceased and 0 otherwise.
-The Kaplan-Meier estimator [@kaplan1958nonparametric] is used to estimate the survival function of the data.
+To model survival probability for all subjects in the combined data, the data were subsetted to include *survival_time* which is either the number of days the individual survived for, or the age in days at last follow up, and *deceased* a numeric event indicator which is equal to 1 if the subject is deceased and 0 if the subject was still alive at the time the respective study had concluded.
+The Kaplan-Meier estimator [@kaplan1958nonparametric] was used to estimate the survival function of the data.
 
@@
 \hat{S}(t) = \prod_{t_i \leq t}\frac{n_i - d_i}{n_i} 
-$$ where $t_i$ is some event time, $d_i$ represents the number of events (here deceased subjects) and $n_i$ indicating the indivudals known to have survived up to $t_i$
+$$ where $t_i$ is some event time, $d_i$ represents the number of events (here deceased subjects) and $n_i$ indicating the individuals known to have survived up to $t_i$
 
 The baseline estimator of all individuals was calculated using the *survival* package in R [@survival-book].
-Next the Kaplan-Meier estimator was calculated, stratified by the reported type of genetic variation reported in the literature.
-A logrank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* R package.
-
-Finally, a Cox proportional hazards regression model [@andersen1982cox] was fit on survival time with the variant type as the sole covariate.
-This was to detect and measure the Hazard ratio of variant type on the survival time in order to test the hypothesis that misense variant types exhibit significantly greater survival compared to other variant types.
+Next the Kaplan-Meier estimator was stratified by the type of genetic variation reported in the literature.
+A log-rank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* R package.
 
@@
 
-The initial results of the MCA were inspected with a bi-plot and identified two subjects (17, 18) from the data that were clear outliers and sat significatly beyound the 95% confidence ellipse.
+The initial results of the MCA were inspected with a bi-plot and identified two subjects (17, 18) from the data that were clear outliers and sat significantly beyond the 95% confidence ellipse.
 These individuals represented two twins from @magini2019loss who were described with a much milder phenotype and were subject to a missense variation in *SMPD4*.
@@
 #| label: fig-biplotprelim
-#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables.
+#| fig-cap: "A bi-plot of the first two principal axes with grey circles indicating the individual's projected position and black triangles indicating the variables.
 A 95% confidence ellipse is plotted in red around the origin."
@@
 
-The *contribution* of the categories were ranked within each of the first four MCA dimensions See @fig-ctr.
-The first dimension is dominated by highly specific facial dysmophisms such as epicanthal folds, short philtrum and low anteriour hairline.
+The *contribution* of the categories were ranked within each of the first four MCA dimensions (@fig-ctr).
+The first dimension is dominated by highly specific facial dysmorphisms such as epicanthal folds, short philtrum and low anterior hairline.
 
@@
 
-Examining the bi-plot of the first two principal axes @fig-biplot shows contrasting sources of variation between conditions relating to airway and swallowing function and predominantly structural brain anomalies such as as cerebellar hypoplasia, agenesis of corpus collosum and development delay.
+Examining the bi-plot of the first two principal axes @fig-biplot illustrates the contrasting sources of variation between conditions relating to airway and swallowing function and predominantly facial features and structural brain anomalies such as as cerebellar hypoplasia, development delay.
 
-The third dimension is characterised by complex features such as patent ductus arteriosus, respiratory infections, small for gestational age and further facial features.
-These are contrasted by the fourth principal dimension which is typical for respiratory distress including stridor and hypercapnia.
+The third dimension is characterised by atypical, complex features such as patent ductus arteriosus, respiratory infections, small for gestational age and further facial features.
+These are contrasted by the fourth principal dimension which is strongly categorised by respiratory distress including stridor and hypercapnia in the absence of typical presentations of simplified gyral pattern, intrauterine growth retardation and microcephaly.
 
-```{r ctr-plot, fig.height=5, fig.width = 8, warning=FALSE}
+```{r ctr-plot, fig.height=6, fig.width = 8, warning=FALSE}
 #| label: fig-ctr
@@
   mutate(vars = fct_reorder(vars, contribution)) |> 
+  mutate(vars = str_remove(vars, "_[:digit:]")) |> 
+  mutate(vars = str_remove(vars, "_ind")) |> 
+  mutate(vars = str_replace_all(vars, "_", " ")) |> 
   ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim), fill = col)) +
@@
 #| label: fig-biplot
-#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables.
+#| fig-cap: "A bi-plot of the first two principal axes with grey circles indicating the individual's projected position and black triangles indicating the variables.
 A 95% confidence ellipse is plotted in grey around the origin."
@@
 #| label: fig-biplot2
-#| fig-cap: "A bi-plot of principal axes 3 and 4, with grey cicles indicating the individual's projected position and black triangles indicaating the variables.
+#| fig-cap: "A bi-plot of principal axes 3 and 4, with grey circles indicating the individual's projected position and black triangles indicating the variables.
 A 95% confidence ellipse is plotted in grey around the origin."
@@
 
-Internal cluster validation measures resulted in between 2 and 6 clusters as the optimal choice.
+Cluster validation measures resulted in either 2 or 5 clusters as the optimal choice.
 
-Proceeding with the optimal cluster selection, per the Silhouette coefficient of 6 clusters.
+In order to partition the data into meaningful groupings, k-means with 5 clusters was selected. This reflects the optimal choice for both AD and FOM cluster metrics.
 
-A scatterplot of the cluster results @fig-clust projected on just the first two dimensions highlights a core cluster of classic features of NEDMABA.
-A separate cluster is formed of airway and feeding related conditions involing vocal cord paralysis, tracheostomy, tube feeding and enlarged ventricles.
+A scatter plot of the cluster results @fig-clust projected on just the first two dimensions highlights a core cluster of classic features of NEDMABA such as microcephaly, hypomyelination, arthrogryposis and feet deformity and structural brain anomalies.
+A separate cluster is formed of airway and feeding related conditions involving vocal cord paralysis, tracheostomy, tube feeding and enlarged ventricles.
 
-Some smaller satellite clusters exists where highly distinctive features were associated with an indivual or family.
-Also a singleton clusters exist for features that are distinct or unusual such as posteriorly angulated ears.
-
+Three other smaller satellite clusters exists where highly distinctive features were associated with an individual or family.
 A full articulation of feature to cluster assignment is given in @tbl-profile.
@@
 #| fig-cap: "cluster validation analysis showing various metrics for cluster evaluation from 2 to 10 clusters.
-The most optimal number of clusters are selected by picking the lowest value from one or many of these compariable metrics."
+The most optimal number of clusters are selected by picking the lowest value from one or many of these comparable metrics."
 
@@
 #| label: fig-clust
-#| fig-cap: "A projection of features on the first two principal components with k-means cluster assignment indicated by colour."
+#| fig-cap: "A projection of features on the first two principal components with k-means cluster assignment indicated by colour.
+A random sample of three labels per cluster are printed to provide a representation of each cluster's characteristics."
 
@@
   as_tibble() |> 
-  arrange(var, clust)
+  arrange(var, clust) |> 
+  mutate(clust = fct_recode(clust,
+    Core = "4",
+    Adjacent = "2",
+    Outlying_3 = "3",
+    Outlying_5 = "5",
+    Outlying_1 = "1",
+    ))
 
@@
   geom_hline(yintercept = 0, lty = 2, colour = "grey50") +
-  geom_label_repel(data = d |> group_by(clust) |> slice_sample(n=4), aes(label = var), size = 3, show.legend = F) +
+  geom_label_repel(data = d |> group_by(clust) |> slice_sample(n=3), aes(label = var), size = 3, show.legend = F) +
   geom_point(aes(colour = clust), position = "jitter", show.legend = F) +
@@
 
-The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with survival probability at time 0 of 0.82 (0.69, 0.98).
-Survival probability after one year at 0.403 (0.252, 0.644) with median survival of 150 days and survival probability at close to ten years of 0.24 (0.11, 0.52).
+The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with median survival time calculated as 150 days.
+At time 0 the survival probability was 0.82 (0.69, 0.98).
+Survival probability after one year at 0.403 (0.252, 0.644) and survival probability at close to ten years of 0.24 (0.11, 0.52).
 
@@
 x <- survminer::surv_summary(kapm_baseline, data = survival_data)
-kable(x, digits = 2, caption = "Survival probabilities derived from Kaplan-Meier estimation. ")
+kable(x, digits = 2, caption = "Survival probabilities derived from Kaplan-Meier estimation for the range of recorded survival times, including 95% Confidence Interval.")
 ```
@@
 #| label: fig-surv
-#| fig-cap: "Kaplan Meier curve with the black line representing the survival probability and with grey shading indicating the 95% confidence interval."
+#| fig-cap: "Kaplan Meier curve with the black line representing the survival probability and with grey shading indicating the 95% confidence interval. Cross icons represent right-censoring"
 survminer::ggsurvplot_df(x,
@@
 
-A comparitive boxplot of survival times by variant type shows significant outliers for missense+splice and missense variations.
-However, even within missense hymozygous variants the survival time varies greatly.
+A comparative box plot of survival times by variant type shows significant outliers for missense+splice and missense variations.
+However, even within missense homozygous variants the survival time varies greatly.
 Other missense variants, such as heterozygous missense with gene deletion do not show the same increased survival time as noted by @magini2019loss.
@@
   labs(fill = 'Variant Type',
-       x = 'Survival Time')
+       x = 'Survival Time',
+       y = '')
 ```
 
-The logrank test was carried out to compare the suvival curves of the various gene variants $\chi^2 = 14.8$ $(p = 0.04)$ however due to the small sample size, several variants had expected counts less than 5 which would create a violation in assumptions for the test.
+A log-rank test was carried out to compare the survival curves of the various gene variants and showed a significant difference in the stratified survival curves ($\chi^2 = 14.8, p = 0.04)$ however due to the small sample size, several variants had expected counts less than 5 which would create a violation in assumptions for the test.
 
@@
 
-The Multiple Correspondence Analysis confirmed the heterogeneous variation in the data with the first four princicple components being required to explain more than 50% of the variation in the dataset.
-
-The examination of key contributing factors, as well as a visual inspection of the dat projected into this lower dimensional space, indicate the primary sources of variation in the clinical phenotype can be explained by few indivudals with very different clinical features.
+The use of Multiple Correspondence Analysis confirmed the heterogeneous variation in the data with the first four principal axes being required to explain more than 50% of the variation in the data set.
+The examination of key contributing factors, as well as a visual inspection of the data projected into this lower dimensional space indicate the primary sources of variation in the clinical phenotype are explained by a few individuals with very different clinical features.
 This has confounded the typical or classical presentation.
-
-Beyond these outliers, a interesting disticntion is made between the hallmark features of microcephaly, hypomyelination, arthrogryposis, simplified gyal pattern and developmental delay with specific airway and swallowing features such as vocal cord paralysis, tracheostomy, GERD and tube feeding.
-
-Many of the outlying features identifes are specific facial or anatomical dysmophims.
-While a noted and core feature is around course facial features, those with a milder course had clinical reports in the available data that articulated these dysmophims in much more detail.
-On a similar note, the reporting of 'talipes' as acondition seems distict as a source of variation when appearing as an isolated feature without the typical microcephasly and hypomyelination.
-However, there was significant heterogenity in how feet contractures were reported in clinical case reports.
-A distinction was made between reported cases with rocker bottom feed or congenital vertical talus and those with clubfeet/talipes.
-In practice its not clear how these diagnoses were arived at, and given the non-specific nature of this feature, care needs to be taken in distinguishing between similar manifesting features in the literaure where possible.
 
@@
 This highlights common sources of variation on the clinical phenotype of the condition itself, rather than typical coincident features of any one child.
+
+After excluding two outlying, remarkable cases this analysis clearly identifies the hallmark features of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay.
+Of significant note is the presence of an adjacent cluster of related features that are not as well described in The Human Phenotype Ontology [@kohler2021human] as shown in @tbl-hpo.
+We identify a prolific feature of 'respiratory distress' as an expression more associated with other features such as vocal cord palsy, tracheostomy, tube feeding and GERD which are present in the case reported data, but not well described in HPO.
+This represents a related but distinct grouping of severe features that associated with this condition.
+
+Other cluster groupings identify outlying features that are expressed in individuals or specific families.
+Many of the outlying features are specific facial or anatomical dysmorphisms.
+While course facial features are a noted and core feature, those with a milder course had clinical reports that articulated these dysmorphisms in much more detail.
+On a similar note, the reporting of 'talipes' as a condition which seems distinct as a source of variation appearing as an isolated feature without the typical microcephaly and hypomyelination.
+However, there was significant heterogeneity in how feet contractures were reported in clinical case reports.
+In the data pre-processing, a separate feature was created for reported cases with rocker bottom feet or congenital vertical talus versus those with clubfeet/talipes.
+In practice its not clear how these diagnoses were arrived at, and given the non-specific nature of this feature, care needs to be taken in distinguishing between similar manifesting features in the literature where possible.
+This is particularly clear as rocker bottom feet was found closely related in our analysis to other core features despite the prevalence in HPO being quite low (2/21), suggesting possible under-reporting or confounded reporting of this feature.
+Another feature that clusters within the core identified features is bilateral cleft lip.
+This condition has a reported frequency of just 2/21 (@tbl-hpo) and may be an emerging feature for a more typical presentation given its close relationship with other classic features.
+Not all features in the data were included in this analysis, particular if isolated to just one case.
+In this circumstance, these terms in @tbl-hpo have been labelled as 'Isolated'.
+
 While the number of clusters chosen all represent valid partition of the data, some metrics indicated an optimal selection of 2 clusters.
-This type of partition is useful when segregating outliers from the rest of the data and is valid in some application.
+This type of partition is useful when segregating outliers from the rest of the data and is valid in some applications.
 In this study, a larger number was selected to align the goals of the analysis which was to uncover distinct groups of features.
@@
 This work identifies a one year survival rate of 0.403 (0.252, 0.644).
-A naive estimate, using just those number of deceased patients as proportion of the study population is 0.57.
+A naive estimate, using just the number of deceased patients at this time as proportion of the study population is 0.57.
 This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to erroneously overestimate the survival probability.
-The current literature suggests the type of gene variant has an impact on surivival time, which seems to be supported in this work through a comparison of multiple survival surves, stratified by variation type, however the limited sample size here prevents other meaningful analysis such as logrank tests or cox proprtional hazard modelling which are likely underpowered when comparing multiple groups.
+The current literature suggests the type of gene variant has an impact on survival time, which seems to be supported in this work.
+Through a comparison of multiple survival curves, stratified by variation type, the limited sample size prevented other meaningful analysis such as log-rank tests or cox proportional hazard modelling which are likely under-powered when comparing multiple groups.
 
-Despite NEDMABA exhibting a diverse and hetergeneous clinical phenotype, it is possible to identify meaningful substructures in this data through the use of dimension reduction techniques (MCA) and cluster analysis.
-This has isolated the classic NEDMABA phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyal pattern and developmental delay.
+Despite NEDMABA exhibiting a diverse and heterogeneous clinical phenotype, our analysis was able to identify meaningful substructures in this data through the use of dimension reduction techniques (MCA) and cluster analysis.
+This has isolated the classic NEDMABA phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay.
 Furthermore it has isolated additional distinct groupings of features that are exhibited in patients such as vocal cord paralysis with tracheostomy, swallowing dysfunction, tube feeding and GERD.
-Furthermore, we were able to quanitfy the expected baseline survival probabilities for children with this condition as well as quantifying the significant variation in these estimates.
-Additional work is required with the benefit of more cases, to better understand the differences in survival curves for other factors associated with genotype.
+Furthermore, we were able to quantify the expected baseline survival probabilities for children with this condition as well as quantifying the significant variation in these estimates.
+Future work is required with the benefit of more cases, to better understand the differences in survival curves for other factors associated with genotype.
+
+```{r}
+#| label: tbl-hpo
+#| tbl-cap: "A mapping table of phenotype terms from The Human Phenotype Ontology (HPO) with clusters defined in this analysis.
+Not all clinical features defined in the analysis are present in the HPO terms"
+#| 
+hpo <- tibble::tribble(
+                           ~`HPO Term Name`,               ~Category,   ~Cluster, ~Frequency,
+                       "Adducted thumb",                 "Limbs", "Isolated",     "1/21",
+             "Simplified gyral pattern",        "Nervous System",     "Core",    "12/18",
+                           "Hypertonia",           "Musculature",     "Core",     "9/15",
+                              "Seizure",        "Nervous System",     "Core",    "10/17",
+                            "Hypotonia",           "Musculature",     "Core",     "4/15",
+           "Global developmental delay",        "Nervous System",     "Core",      "7/7",
+                "Highly arched eyebrow", "Skin, Hair, and Nails", "Isolated",     "2/21",
+                "Cerebellar hypoplasia",        "Nervous System",     "Core",    "10/21",
+        "Neonatal respiratory distress",    "Respiratory System", "Adjacent",    "14/17",
+                          "Bifid uvula",         "Head and neck", "Isolated",     "1/21",
+                        "Deep philtrum",         "Head and neck", "Isolated",     "1/21",
+                  "CNS hypomyelination",        "Nervous System",     "Core",     "7/19",
+                "Lumbar hypertrichosis", "Skin, Hair, and Nails", "Isolated",     "1/21",
+  "Birth length less than 3rd percentile",                "Growth",     "Core",     "6/11",
+          "Hypoplasia of the brainstem",        "Nervous System", "Isolated",     "3/20",
+           "Tented upper lip vermilion",         "Head and neck", "Isolated",     "3/21",
+                         "Hypotelorism",                   "Eye", "Isolated",     "1/21",
+              "Short palpebral fissure",         "Head and neck", "Outlying",     "4/21",
+                  "Bilateral cleft lip",         "Head and neck",     "Core",     "2/21",
+                    "Diabetes mellitus",             "Endocrine", "Isolated",     "2/19",
+      "Single transverse palmar crease", "Skin, Hair, and Nails", "Isolated",     "1/21",
+                           "Epicanthus",         "Head and neck", "Outlying",     "5/21",
+                "Low anterior hairline",         "Head and neck", "Outlying",     "2/21",
+   "Arthrogryposis multiplex congenita",     "Connective tissue",     "Core",    "17/20",
+             "Progressive microcephaly",        "Nervous System",     "Core",     "9/10",
+                         "Microcephaly",        "Nervous System",     "Core",    "15/21",
+             "Thin upper lip vermilion",         "Head and neck", "Outlying",     "2/21",
+                "Thin vermilion border",         "Head and neck", "Isolated",     "3/21",
+      "Intrauterine growth retardation",                "Growth",     "Core",    "12/19",
+             "Posteriorly rotated ears",                   "Ear", "Adjacent",     "2/21",
+                      "Narrow forehead",         "Head and neck", "Isolated",     "1/21",
+                     "Sloping forehead",         "Head and neck", "Isolated",     "4/21",
+                      "Smooth philtrum",         "Head and neck", "Isolated",     "2/21",
+                       "Short philtrum",         "Head and neck", "Outlying",     "2/21",
+                      "Premature birth",    "Prenatal and Birth",     "Core",     "7/18",
+                "Mandibular prognathia",         "Head and neck", "Outlying",     "2/21",
+         "Wide intermamillary distance",                "Breast", "Isolated",     "1/21",
+               "Depressed nasal bridge",         "Head and neck", "Outlying",     "1/21",
+       "Downslanted palpebral fissures",         "Head and neck", "Outlying",     "2/21",
+                           "Short neck",         "Head and neck", "Isolated",     "1/21",
+                         "Bulbous nose",         "Head and neck", "Isolated",     "2/21",
+                       "Protruding ear",                   "Ear", "Outlying",     "2/21",
+                   "Rocker bottom foot",                 "Limbs",     "Core",     "2/21"
+  )
+
+hpo |> 
+  gt() |> 
+   tab_source_note(
+    source_note = "Source: The Human Phenotype Ontology (HPO) https://hpo.jax.org/app/browse/disease/OMIM:618622 "
+  ) |> 
+  cols_move_to_start(
+    columns = c(`HPO Term Name`, Category, Frequency)
+  ) |> 
+   cols_label(
+     Cluster = md("**Cluster**")
+  ) |> 
+   tab_spanner(
+    label = "HPO Data",
+    columns = c(`HPO Term Name`, Category, Frequency)
+  ) |> 
+  tab_spanner(
+    label = "Derived Clusters",
+    columns = c(Cluster)
+  ) |> 
+  tab_source_note(
+    source_note = md("Reference: Köhler S, Gargano M, Matentzoglu N, Carmody LC, Lewis-Smith D, Vasilevsky NA, Danis D, Balagura G, Baynam G, Brower AM, Callahan TJ, Chute CG, Est JL, Galer PD, Ganesan S, Griese M, Haimel M, Pazmandi J, Hanauer M, Harris NL, Hartnett MJ, Hastreiter M, Hauck F, He Y, Jeske T, Kearney H, Kindle G, Klein C, Knoflach K, Krause R, Lagorce D, McMurry JA, Miller JA, Munoz-Torres MC, Peters RL, Rapp CK, Rath AM, Rind SA, Rosenberg AZ, Segal MM, Seidel MG, Smedley D, Talmy T, Thomas Y, Wiafe SA, Xian J, Yüksel Z, Helbig I, Mungall CJ, Haendel MA, Robinson PN.
+*The Human Phenotype Ontology in 2021.* Nucleic Acids Res.
+2021 Jan 8;49(D1):D1207-D1217.
+doi: [10.1093/nar/gkaa1043](10.1093/nar/gkaa1043).
+PMID: 33264411; PMCID: PMC7778952.")
+  ) |> 
+  tab_footnote(
+    footnote = "Clusters defined in this analysis mapped to HPO Terms",
+    locations = cells_column_labels(columns = "Cluster")
+  )
+```
 
@@
 #| label: tbl-dict
-#| tbl-cap: Data dictionary of the final dataset used for the analysis.
+#| tbl-cap: Data dictionary of the final data set used for the analysis.
 This combines the case reports and supplied data for available and references studies.
@@
   select(cluster, variable) |> 
+  mutate(variable = str_remove(variable, "_[:digit:]")) |> 
+  mutate(variable = str_remove(variable, "_ind")) |> 
   arrange(cluster) |> 

### 2022-09-07 test/paper.qmd
movign paper into own dir

--- 2022-09-07 paper.qmd
@@
 title: "Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies"
-author: "Dean Marchiori"
+author:
+  - id: dm
+    number: 1
+    name: Dean Marchiori
+    email: deanmarchiori@gmail.com
+    corresponding: TRUE
+date: "September 2022"
 format: pdf
@@
 |                     | Family 1- Individual 1                                                                                     | Family 1- Individual 4                                      |
-|---------------|-------------------------------------|--------------------|
+|------------------|-----------------------------------|-------------------|
 | Facial dysmorphisms | short palpebral fissures, large ears, simple helices, smooth philtrum, thin lips, bilateral simian creases | short palpebral fissures, receding forehead, thin upper lip |
@@
 | id                     | short palpebral fissures | large ears | simple helices | receding forehead | ... |
-|------------|-------------|------------|------------|------------|------------|
+|------------|------------|------------|------------|------------|------------|
 | Family 1- Individual 1 | 1                        | 1          | 1              | 0                 | ... |

### 2022-09-07 _targets.R

@@ -44,7 +44,7 @@ list(
   tar_target(cox_ph, coxph(Surv(survival_time, deceased) ~ variant, data = survival_data)),
 
   # Render Report
-  tar_quarto(report, path = "paper.qmd")
+  tar_quarto(report, path = "test/paper.qmd")
 )
 
 

### 2022-09-07 paper/paper.qmd
updated template

--- 2022-09-07 paper.qmd
@@
 ---
-title: "Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies"
-author: "Dean Marchiori"
-format: pdf
-abstract: "A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*.
+title: Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies
+author:
+  - name: Dean Marchiori
+    email: deanmarchiori@gmail.com
+    attributes:
+        corresponding: true
+abstract: |
+  A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*.
 The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and clinically diverse phenotype.
@@
 This wide range of outcomes is provisionally associated with variant type however interpretation of these results are guarded based on very low sample sizes.
-Despite challenges with sparse and inconsistent data, this analysis represents the first of its kind to help describe associations and trajectories of individuals with this condition and can provide clinicians and genetic counsellors with better information to aide in decision making and support for families."
-editor: visual
-bibliography: refs.bib
+Despite challenges with sparse and inconsistent data, this analysis represents the first of its kind to help describe associations and trajectories of individuals with this condition and can provide clinicians and genetic counsellors with better information to aide in decision making and support for families.
+keywords: 
+  - NEDMABA
+  - SMPD4
+#journal: "An awesome journal"
+date: last-modified
 execute: 
   echo: false 
+bibliography: bibliography.bib
+format:
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      formatting: preprint
+      model: 3p
+      cite-style: authoryear
 ---
@@
 theme_set(theme_bw())
-```
-
-```{r}
-# table with med gen 
 ```
@@
 
+\newpage
+
+# Appendix
+
 ```{r}
@@
 
-\newpage
-
-# Appendix
-
-```{r}
-#| label: tbl-dict
-#| tbl-cap: Data dictionary of the final data set used for the analysis.
-This combines the case reports and supplied data for available and references studies.
-The data has been transformed into a tidy format with formatting applied.
-tar_load(subjects)
-make_data_dictionary(subjects)
-```
-
 ```{r}
@@
 
+```{r}
+#| label: tbl-dict
+#| tbl-cap: Data dictionary of the final data set used for the analysis.
+This combines the case reports and supplied data for available and references studies.
+The data has been transformed into a tidy format with formatting applied.
+tar_load(subjects)
+make_data_dictionary(subjects)
+```
+
 \newpage
 
-# References
+# References {-}

### 2022-09-07 paper/paper.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint,
  3p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies},
  pdfauthor={Dean Marchiori},
  pdfkeywords={NEDMABA, SMPD4},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{Statistical Analysis of Clinical Phenotype and Patient Survival
in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and
Structural Brain Anomalies}
\author[]{Dean Marchiori%
\corref{cor1}%
}
 \ead{deanmarchiori@gmail.com}

\cortext[cor1]{Corresponding author}

\begin{abstract}
A recently described, rare genetic condition known as Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies (NEDMABA) has been identified in children with bi-allelic
loss-of-function variants in \emph{SMPD4}. The progression of this
condition is not well understood with the limited case reports described
so far exhibiting a severe and clinically diverse phenotype. A gap
exists in the understanding the associations of the heterogenous
features present in the clinical phenotype and the expected survival
probabilities of affected individuals. This is driven in part to the
paucity of analysis-ready data in reported cases. This research aims to
collate and standardise available case reports to analyse and identify
meaningful clusters in the clinical phenotype, and to quantify the
survival probability for children with NEDMABA. To overcome the
challenge of sparse, multidimensional data on very few subjects, we
employ Multiple Correspondence Analysis (MCA) as a dimension reduction
technique, which is then subject to cluster analysis and interpretation.
To quantify more accurate survival probabilities, Kaplan-Meier
estimation is formulated to account for censoring in the data. The
analysis identified the classic phenotype for this condition, as well as
one other distinct feature cluster relating to findings of vocal cord
paralysis and respiratory failure. The survival probability for those
affected was found to decline sharply in early infancy with median
survival of 150 days, but with some as high as 12.5 years. This wide
range of outcomes is provisionally associated with variant type however
interpretation of these results are guarded based on very low sample
sizes. Despite challenges with sparse and inconsistent data, this
analysis represents the first of its kind to help describe associations
and trajectories of individuals with this condition and can provide
clinicians and genetic counsellors with better information to aide in
decision making and support for families.
\end{abstract}

\begin{keyword}
    NEDMABA \sep 
    SMPD4
\end{keyword}
\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Neurodevelopment disorders are a diverse and heterogeneous group of
conditions impacting the development of the nervous system, brain
function, physical development and learning ability. A recently
identified condition known as Neurodevelopmental Disorder with
Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA)
(MIM:\href{https://omim.org/entry/618622}{618622}) has been described in
children with bi-allelic loss-of-function variants in \emph{SMPD4}
\citep{magini2019loss}. Sphingomyelinases such as \emph{SMPD4} play an
important cellular role by hydrolyzing sphingomyelin into ceramide and
phosphorylcholine. \emph{SMPD4} specifically encodes one of 4 neutral
sphinogmyelinases, nSMase3 (MIM:
\href{http://omim.org/entry/610457}{610457}). So far, less than 50 cases
of this rare disorder have been reported in the literature. Clinical
phenotype data of these cases is largely heterogeneous with severe
neurological complications and early demise a key feature. This makes
the identification, diagnosis and ongoing management of cases a
challenge for patients, their families and medical practitioners.

\citet{ravenscroft2021neurogenetic} performed a study over 190 probands
and identified a novel missense variation in SMPD4 which at the time of
whole exome sequencing was not well described in the literature. The
impacted family from Melbourne exhibited features involving
arthrogryposis multiplex congenita, complex brain malformations, small
for gestation age and hypoplasia of the corpus callosum. In two of the
three related cases were additional features of microcephaly, congenital
encephalopathy, cerebellar malformation and hypoplasia and
hypomyelination.

A detailed study by \citet{magini2019loss} involved 12 unrelated
families with 32 individuals (21 with detailed clinical information).
Presentations were of microcephaly, simplified gyral pattern of the
cortex, hypomyelination, cerebellar hypoplasia, congenital
arthrogryposis, and early fetal/postnatal demise. Despite this being the
largest cohort studied, the clinical features and survival times among
participants varied greatly. Three missense changes were noted in the
study with affected children in these families often showing a milder
presentation suggestive of possible residual function. In these cases
individuals were able to develop independent motor skills, have mild
intellectual disability and arthrogryposis without evidence of
simplified gyral patterns on brain MRI. Other patients with truncating
variants are shown to have more severe presentations, while a range of
additional significant clinical features were reported involving
dysmorphic facial features, seizure, vocal cord paralysis and hearing
impairment.

A case study from China is described in \citet{ji2022case} involving a
girl presenting in infancy with intrauterine growth restriction,
microcephaly, postnatal developmental delay, arthrogryposis,
hypertonicity, seizure, and hypomyelination on brain magnetic resonance
imaging. The authors report parallels with two cases showing the same
homozygous null variant. An individual reported in
\citet{monies2019lessons} presented with distinct symptoms of brain
atrophy and skeletal dysplasia whereas the case in
\citet{magini2019loss} with the same variant exhibited more typical
clinical features.

A recent study by \citet{bijarnia2022growth} presents the case of a
22-month old girl presenting with the typical phenotype of
neurodevelopmental delay, prenatal onset growth failure, arthrogryposis,
microcephaly and brain anomalies including severe hypomyelination,
simplified gyral pattern and hypoplasia of corpus callosum and
brainstem. Notably, there was also additional non-typical clinical
findings of nystagmus and visual impairment secondary to macular
dystrophy and retinal pigment epithelial stippling at posterior pole.

Further work to collate and analyse data relating to NEDMABA is
challenging. While \citet{magini2019loss} cataloged a detailed clinical
phenotype data set as a supplementary data set to their study, these
data are not suitable for statistical analysis as presented. Many of the
clinical features are represented as free text descriptions and a
variety of non-standard terminologies are applied between cases, making
machine interpretation of the data difficult. Other studies
\citep{ji2022case, bijarnia2022growth} present only written case reports
with some tabulated summaries. While larger studies
\citep{ravenscroft2021neurogenetic, monies2019lessons} focus more on
genetic data, with less detail included on the clinical presentation of
individuals in the cohort.

To overcome these data challenges, \citet{JSSv059i10} proposes a data
structure know as `tidy' data which organises each observation in a row,
with each feature as its own column and each value in just one cell. In
this case detailed text-based data can be transformed to high
dimensional binary indicators of key clinical features. This structure
is conducive to effective statistical analysis and integrates
deliberately with the \emph{tidyverse} \citep{tidyverse} collection of
packages for data analysis in R \citep{rbase}.

Further analysis to better describe rare genetic conditions was explored
in \citet{diaz2020phenotype} in the context of large scale
genotype-phenotype analysis. The authors argue patients with rare
disorders (often in small samples) often present with varied symptoms
that do not match exactly with the described phenotype. When considering
this situation of low sample size data with many features, methods in
the field of multivariate statistical analysis are commonly deployed.
Here the aim is to find substructure in the data, or a simple
representation of the multi-dimensional space. Methods, such as Multiple
Correspondence Analysis \citep{le2010multiple} are appropriate for high
dimensional data comprised of binary indicators. This technique is cited
in many studies in the application of dimension reduction and clustering
of comorbidities and phenotypes from disease
\citep{han2018cluster, costa2013use}.

In terms of patient survival, methods in statistical survival analysis
are well suited and commonly used in the field of genomics
\citep{chen2014survival}. These methods in particular provide a
mechanism for dealing with right-censoring, a common feature of clinical
studies where the clinical outcome of interest may not be known by the
end of the study period.

An open research question exists around the clinical pathway and
survival time of children exhibiting bi-allelic loss-of-function
variants in \emph{SMPD4}. With current research highlighting a diverse
and heterogeneous phenotype, the relationship and correlations between
these diverse features is not well understood. Furthermore, the survival
time of affected children has not been extensively analysed despite
early-demise featuring as a severe outcome in many cases. While a
connection has been highlighted between some missense variants and a
milder presentation with longer survival \citep{magini2019loss}, this
hypothesis has not been analysed further.

This research has three key aims. First, to collate and transform the
variety of early case reports and studies on clinical phenotype data for
this novel variant into an analysis-ready \emph{tidy} data set.
Secondly, to conduct analysis on the associations between commonly
reported clinical features. Finally, to statistically quantify the
expected survival time of children with NEDMABA based on the current
case reports.

This will be the first in-depth analysis of this novel variant, which
aims to produce statistical findings to better understand this newly
described condition. This research will assist clinicians understand the
various presentations of this challenging new condition. In addition,
genetic counselling of families with affected children will benefit from
enhanced analysis on outcomes of existing reported cases.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{smpd4-data-package}{%
\subsection{SMPD4 Data Package}\label{smpd4-data-package}}

The data contained in \citet{magini2019loss} ``Summary of SMPD4-Related
Clinical Phenotype'' is an excel spreadsheet tabulating each of the 21
individuals from the study with clinical details recorded. The file has
21 columns (one for each individual), and 64 rows (one for each
phenotype or clinical remark).

The data were read into \texttt{R} \citep{rbase} without any changes, so
as to preserve the reproducibility of the data transformation steps. The
data were transposed to ensure it could be presented as \emph{tidy}
formatted data \citep{JSSv059i10}. This requires each variable (clinical
phenotype) forms a column; each observation (individual) forms a row and
each type of observational unit forms a table (every cell has just one
item). This resulted in a long-formatted data set of 21 rows
(individuals) and 64 columns (clinical features). In many cases, key
clinical information was entered as free-text descriptions, which
rendered any attempt of meaningful analysis impractical
(Table~\ref{tbl-nontidy}). In these cases, the text was tokenised by
separating the list of clinical observations at each comma and forming a
binary indicator column noting its presence `1' or absence `0'
(Table~\ref{tbl-tidy}).

\hypertarget{tbl-nontidy}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5139}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2778}}@{}}
\caption{\label{tbl-nontidy}Example of non-tidy free-text descriptions
of features}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Family 1- Individual 1
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Family 1- Individual 4
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Family 1- Individual 1
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Family 1- Individual 4
\end{minipage} \\
\midrule()
\endhead
Facial dysmorphisms & short palpebral fissures, large ears, simple
helices, smooth philtrum, thin lips, bilateral simian creases & short
palpebral fissures, receding forehead, thin upper lip \\
\bottomrule()
\end{longtable}

\hypertarget{tbl-tidy}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1781}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}@{}}
\caption{\label{tbl-tidy}Example of tidy formatted data where
individuals are transposed into rows and text into binary
indicators}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
id
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
short palpebral fissures
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
large ears
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
simple helices
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
receding forehead
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\ldots{}
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
id
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
short palpebral fissures
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
large ears
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
simple helices
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
receding forehead
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\ldots{}
\end{minipage} \\
\midrule()
\endhead
Family 1- Individual 1 & 1 & 1 & 1 & 0 & \ldots{} \\
Family 1- Individual 4 & 1 & 0 & 0 & 1 & \ldots{} \\
\bottomrule()
\end{longtable}

In the case where two variables were formed from features that are
considered to be synonymous, these were merged into one indicator column
to prevent duplication e.g.~\{bilateral\_cleft\_lips,
bilateral\_cleft\_lip, cleft\_lip\_b\_l\} -\textgreater{}
\{bilateral\_cleft\_lip\}.

Data type conversion and categorical level standardisation was performed
to ensure the data were in consistent and appropriate data types. For
example, `Gender' was not consistently coded, and `Birth Weight' was
encoded as a text string rather than a more useful numeric format.
(Table~\ref{tbl-data-types}).

\hypertarget{tbl-data-types}{}
\begin{longtable}[]{@{}ll@{}}
\caption{\label{tbl-data-types}Example of inconsistent coding or
sub-optimal data types}\tabularnewline
\toprule()
Gender & Birth Weight \\
\midrule()
\endfirsthead
\toprule()
Gender & Birth Weight \\
\midrule()
\endhead
male & 2175 grams (- 2.5 SD) \\
female & 2045 g (-3 SD) \\
Female & 2300 gram (-2 SD) \\
Female fetus & n.a. \\
\bottomrule()
\end{longtable}

\hypertarget{tbl-data-types2}{}
\begin{longtable}[]{@{}ll@{}}
\caption{\label{tbl-data-types2}Example of consistent and appropriate
coding or data types}\tabularnewline
\toprule()
Gender & Birth Weight (g) \\
\midrule()
\endfirsthead
\toprule()
Gender & Birth Weight (g) \\
\midrule()
\endhead
male & 2175 \\
female & 2045 \\
female & 2300 \\
female & NA \\
\bottomrule()
\end{longtable}

This format was preserved and other case studies identified in the
literature
\citep{ravenscroft2021neurogenetic, monies2019lessons, bijarnia2022growth, ji2022case}
were manually entered to conform to this template to allow the data to
be combined for further analysis.

The final data set consisted of 28 observations (one per individual) and
152 variables (one per clinical feature). These variables were comprised
of qualitative text descriptions, binary indicators of features as
described above and other fields including background information on the
patient and various bio markers. A full list of this data set is
available in Table~\ref{tbl-dict}. These data sets were compiled into a
publicly available \texttt{R} Package called \texttt{SMPD4}
\citep{smpd4-data} in order to allow for reproducibility and sharing.

\hypertarget{multiple-correspondence-analysis}{%
\subsection{Multiple Correspondence
Analysis}\label{multiple-correspondence-analysis}}

The data on subjects introduced above is subset to include only
variables that can be transformed into binary indicator variables that
represent the presence or absence of a given clinical feature column.
Only those features that appeared in more than one case were included to
minimise the influence of non-related features in this analysis.

Dimension reduction on this wide data set was performed using Multiple
Correspondence Analysis (MCA) (an analogy to Principle Component
Analysis (PCA) for categorical data) \citep{le2010multiple} in order to
arrive at a set of meaningfully small dimensions that account for most
of the variation in the data.

We let these data be \(\mathbf{X}\) which is comprised of a set of
individuals \(I\) and a set of features \(Q\) such that the \(q_{th}\)
feature has \(K_q\) levels. The sum of all categories
\(K = \sum_{q=1}^{Q}K_q\) defines the dimensionality of \(\mathbf{X}\)
as an \(I\times K\) matrix. This resulted in a data set
\(X_{28 \times 61}\) indicator matrix of 61 clinical features across 28
individuals. Taking \(\delta_{ik} = 1\) if subject \(i\) has feature
\(k\) and \(\delta_{ik} = 0\) if the subject does not, we are left with
the completely disjunctive table \(\mathbf{X} = I\times K\) of
\(\{0,1\}\). Letting the sum of all entries of \(\mathbf{X}\) be \(N\),
we have \(\mathbf{Z} = N^{-1}\mathbf{X}\). We can introduce two diagonal
matrices \(\mathbf{D_r} = diag(\mathbf{r})\) and
\(\mathbf{D_c} = diag(\mathbf{c})\) where \(\mathbf{r}\) and
\(\mathbf{c}\) are the vectors of row sums and column sums of
\(\mathbf{Z}\) respectively.

Computing MCA involves taking the Singular Value Decomposition:

\[
\mathbf{M = D_r^{-\frac{1}{2}}(Z-rc^T)D_c^{-\frac{1}{2}}} = \mathbf{P}\Delta\mathbf{Q^T}
\] where \(\Delta\) are the singular values and \(\Lambda = \Delta^2\)
is the matrix of eigenvalues.

This results in the row and column factor scores respectively as:

\[
\begin{split}
\mathbf{F} = \mathbf{D_r^{-\frac{1}{2}}P}\Delta \\
\mathbf{G = D_c^{-\frac{1}{2}}Q}\Delta
\end{split}
\]

This computation of the Multiple Correspondence Analysis was computed
using the \emph{FactoMineR} R package \citep{factominer}.

The \emph{contribution} of each category was computed for each of the
principal axis per \citet{le2010multiple}. This metric identifies the
proportion of variance of that axis due to the point and is defined as:

\[
Ctr_k = \frac{py_k^2}{\lambda}
\] Where \(p\) represents the point weighting, \(y\) is the coordinate
relative to the principal axes of variance \(\lambda\).

The coordinates on the first few principal axes were inspected for any
outliers or unusual data.

\hypertarget{cluster-analysis}{%
\subsection{Cluster Analysis}\label{cluster-analysis}}

The MCA coordinates for the first 4 principal dimensions were subject to
clustering in order to find groups of features that were internally
homogeneous and externally heterogeneous. This was computed in R using
the \texttt{clValid} package \citep{clvalid}.

A range of clustering algorithms were compared including AGNES
hierarchical agglomerative \citep{kaufman2009finding}, k-means
\citep{hartigan1979algorithm} and self-organising maps (SOM)
\citep{kohonen2012self}. A sensible range of clusters from 2-5 were
trialed for each clustering algorithm. The results for each method were
evaluated using three measures of stability: Average proportion of
non-overlap (APN), Average Distance (AD) and Average Distance Between
Means (ADM) \citep{datta2003comparisons}. Stability measures compare the
full clustering result with a result based on dropping each of the
columns one at a time. These metrics were selected to represent each
data point (clinical feature) with a number of MCA dimensions. A good
cluster solution is one that is robust and meaningful across all of
these dimensions Figure~\ref{fig-clvalid}. The optimal method and
cluster number was selected based on a trade off between the best
performing method using the evaluation metrics and a sensible
partitioning of the data for this analysis Table~\ref{tbl-eig}.

\hypertarget{survival-analysis}{%
\subsection{Survival Analysis}\label{survival-analysis}}

To model survival probability for all subjects in the combined data, the
data were subsetted to include \emph{survival\_time} which is either the
number of days the individual survived for, or the age in days at last
follow up, and \emph{deceased} a numeric event indicator which is equal
to 1 if the subject is deceased and 0 if the subject was still alive at
the time the respective study had concluded. The Kaplan-Meier estimator
\citep{kaplan1958nonparametric} was used to estimate the survival
function of the data.

The estimator is calculated as:

\[
\hat{S}(t) = \prod_{t_i \leq t}\frac{n_i - d_i}{n_i} 
\] where \(t_i\) is some event time, \(d_i\) represents the number of
events (here deceased subjects) and \(n_i\) indicating the individuals
known to have survived up to \(t_i\)

The baseline estimator of all individuals was calculated using the
\emph{survival} package in R \citep{survival-book}. Next the
Kaplan-Meier estimator was stratified by the type of genetic variation
reported in the literature. A log-rank test was performed to detect
differences in the survival curves using methods from
\citet{harrington1982class}, again implemented in the \emph{survival} R
package.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{mca}{%
\subsection{MCA}\label{mca}}

The initial results of the MCA were inspected with a bi-plot and
identified two subjects (17, 18) from the data that were clear outliers
and sat significantly beyond the 95\% confidence ellipse. These
individuals represented two twins from \citet{magini2019loss} who were
described with a much milder phenotype and were subject to a missense
variation in \emph{SMPD4}. Further analysis of this influence is
conducted below. For now, these subjects were removed to allow for more
meaningful analysis.

Repeating MCA for the remaining subjects shows the first dimension
(\(\lambda_! = 0.18\)) accounts for 17.88\% of the variance in the data
with the first 4 dimensions of the MCA analysis accounting for 54\% of
the variance (Table~\ref{tbl-eig}).

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-biplotprelim-1.pdf}

}

\caption{\label{fig-biplotprelim}A bi-plot of the first two principal
axes with grey circles indicating the individual's projected position
and black triangles indicating the variables. A 95\% confidence ellipse
is plotted in red around the origin.}

\end{figure}

\hypertarget{tbl-eig}{}
\begin{longtable}{lrrr}
\caption{\label{tbl-eig}Eigenvalues and percentage of variance explained by the MCA principal
axes }\tabularnewline

\toprule
dimension & eigenvalue & percentage of variance & cumulative percentage of variance \\ 
\midrule
dim 1 & $0.18$ & $17.88\%$ & $17.88\%$ \\ 
dim 2 & $0.14$ & $13.81\%$ & $31.70\%$ \\ 
dim 3 & $0.12$ & $12.32\%$ & $44.01\%$ \\ 
dim 4 & $0.10$ & $9.66\%$ & $53.67\%$ \\ 
dim 5 & $0.07$ & $7.49\%$ & $61.16\%$ \\ 
dim 6 & $0.06$ & $5.97\%$ & $67.13\%$ \\ 
dim 7 & $0.05$ & $4.90\%$ & $72.03\%$ \\ 
dim 8 & $0.04$ & $4.20\%$ & $76.23\%$ \\ 
dim 9 & $0.04$ & $4.08\%$ & $80.31\%$ \\ 
dim 10 & $0.03$ & $3.47\%$ & $83.78\%$ \\ 
dim 11 & $0.03$ & $2.92\%$ & $86.70\%$ \\ 
dim 12 & $0.02$ & $2.49\%$ & $89.20\%$ \\ 
dim 13 & $0.02$ & $2.15\%$ & $91.34\%$ \\ 
dim 14 & $0.02$ & $1.82\%$ & $93.16\%$ \\ 
dim 15 & $0.02$ & $1.65\%$ & $94.81\%$ \\ 
dim 16 & $0.01$ & $1.12\%$ & $95.93\%$ \\ 
dim 17 & $0.01$ & $1.00\%$ & $96.93\%$ \\ 
dim 18 & $0.01$ & $0.93\%$ & $97.85\%$ \\ 
dim 19 & $0.01$ & $0.67\%$ & $98.53\%$ \\ 
dim 20 & $0.01$ & $0.63\%$ & $99.16\%$ \\ 
dim 21 & $0.00$ & $0.41\%$ & $99.57\%$ \\ 
dim 22 & $0.00$ & $0.31\%$ & $99.88\%$ \\ 
dim 23 & $0.00$ & $0.12\%$ & $100.00\%$ \\ 
dim 24 & $0.00$ & $0.00\%$ & $100.00\%$ \\ 
dim 25 & $0.00$ & $0.00\%$ & $100.00\%$ \\ 
\bottomrule
\end{longtable}

The \emph{contribution} of the categories were ranked within each of the
first four MCA dimensions (Figure~\ref{fig-ctr}). The first dimension is
dominated by highly specific facial dysmorphisms such as epicanthal
folds, short philtrum and low anterior hairline.

The second MCA dimension is categorised by feeding and respiratory
dysfunction with vocal cord palsy, tracheostomy, tube feeding and GERD.

Examining the bi-plot of the first two principal axes
Figure~\ref{fig-biplot} illustrates the contrasting sources of variation
between conditions relating to airway and swallowing function and
predominantly facial features and structural brain anomalies such as as
cerebellar hypoplasia, development delay.

The third dimension is characterised by atypical, complex features such
as patent ductus arteriosus, respiratory infections, small for
gestational age and further facial features. These are contrasted by the
fourth principal dimension which is strongly categorised by respiratory
distress including stridor and hypercapnia in the absence of typical
presentations of simplified gyral pattern, intrauterine growth
retardation and microcephaly.

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-ctr-1.pdf}

}

\caption{\label{fig-ctr}Top 10 feature categories from MCA analysis for
the first four MCA dimensions. The red dashed line indicates the mean
contribution value. Each feature may be present or absent as indicated
by 1 or 0 respectively}

\end{figure}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-biplot-1.pdf}

}

\caption{\label{fig-biplot}A bi-plot of the first two principal axes
with grey circles indicating the individual's projected position and
black triangles indicating the variables. A 95\% confidence ellipse is
plotted in grey around the origin.}

\end{figure}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-biplot2-1.pdf}

}

\caption{\label{fig-biplot2}A bi-plot of principal axes 3 and 4, with
grey circles indicating the individual's projected position and black
triangles indicating the variables. A 95\% confidence ellipse is plotted
in grey around the origin.}

\end{figure}

\hypertarget{cluster-analysis-1}{%
\subsection{Cluster Analysis}\label{cluster-analysis-1}}

Cluster validation measures resulted in either 2 or 5 clusters as the
optimal choice.

In order to partition the data into meaningful groupings, k-means with 5
clusters was selected. This reflects the optimal choice for both AD and
FOM cluster metrics.

A scatter plot of the cluster results Figure~\ref{fig-clust} projected
on just the first two dimensions highlights a core cluster of classic
features of NEDMABA such as microcephaly, hypomyelination,
arthrogryposis and feet deformity and structural brain anomalies. A
separate cluster is formed of airway and feeding related conditions
involving vocal cord paralysis, tracheostomy, tube feeding and enlarged
ventricles.

Three other smaller satellite clusters exists where highly distinctive
features were associated with an individual or family. A full
articulation of feature to cluster assignment is given in
Table~\ref{tbl-profile}.

\hypertarget{tbl-opticlust}{}
\begin{longtable}{lrlr}
\caption{\label{tbl-opticlust}Optimal clusters }\tabularnewline

\toprule
Metric & Score & Method & Clusters \\ 
\midrule
APN & $0.07$ & agnes & 2 \\ 
AD & $1.38$ & kmeans & 5 \\ 
ADM & $0.33$ & kmeans & 2 \\ 
FOM & $0.65$ & kmeans & 5 \\ 
\bottomrule
\end{longtable}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-clvalid-1.pdf}

}

\caption{\label{fig-clvalid}cluster validation analysis showing various
metrics for cluster evaluation from 2 to 10 clusters. The most optimal
number of clusters are selected by picking the lowest value from one or
many of these comparable metrics.}

\end{figure}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-clust-1.pdf}

}

\caption{\label{fig-clust}A projection of features on the first two
principal components with k-means cluster assignment indicated by
colour. A random sample of three labels per cluster are printed to
provide a representation of each cluster's characteristics.}

\end{figure}

\hypertarget{survival-analysis-1}{%
\subsection{Survival Analysis}\label{survival-analysis-1}}

The Kaplan-Meier estimator shows a range of survival times (0, 4562
days) with median survival time calculated as 150 days. At time 0 the
survival probability was 0.82 (0.69, 0.98). Survival probability after
one year at 0.403 (0.252, 0.644) and survival probability at close to
ten years of 0.24 (0.11, 0.52).

\begin{longtable}[]{@{}rrrrrrrr@{}}
\caption{Survival probabilities derived from Kaplan-Meier estimation for
the range of recorded survival times, including 95\% Confidence
Interval.}\tabularnewline
\toprule()
time & n.risk & n.event & n.censor & surv & std.err & upper & lower \\
\midrule()
\endfirsthead
\toprule()
time & n.risk & n.event & n.censor & surv & std.err & upper & lower \\
\midrule()
\endhead
0.0 & 28 & 5 & 0 & 0.82 & 0.09 & 0.98 & 0.69 \\
15.0 & 23 & 1 & 0 & 0.79 & 0.10 & 0.95 & 0.65 \\
30.0 & 22 & 2 & 0 & 0.71 & 0.12 & 0.90 & 0.57 \\
47.0 & 20 & 1 & 0 & 0.68 & 0.13 & 0.88 & 0.53 \\
60.0 & 19 & 1 & 1 & 0.64 & 0.14 & 0.85 & 0.49 \\
67.0 & 17 & 1 & 0 & 0.61 & 0.15 & 0.82 & 0.45 \\
90.0 & 16 & 0 & 1 & 0.61 & 0.15 & 0.82 & 0.45 \\
120.0 & 15 & 2 & 0 & 0.52 & 0.18 & 0.75 & 0.37 \\
150.0 & 13 & 1 & 0 & 0.48 & 0.20 & 0.72 & 0.33 \\
180.0 & 12 & 1 & 0 & 0.44 & 0.22 & 0.68 & 0.29 \\
240.0 & 11 & 1 & 0 & 0.40 & 0.24 & 0.64 & 0.25 \\
270.0 & 10 & 0 & 1 & 0.40 & 0.24 & 0.64 & 0.25 \\
365.0 & 9 & 0 & 1 & 0.40 & 0.24 & 0.64 & 0.25 \\
750.0 & 8 & 1 & 0 & 0.35 & 0.27 & 0.60 & 0.21 \\
1095.0 & 7 & 1 & 0 & 0.30 & 0.31 & 0.56 & 0.16 \\
1825.0 & 6 & 0 & 1 & 0.30 & 0.31 & 0.56 & 0.16 \\
2190.0 & 5 & 1 & 0 & 0.24 & 0.39 & 0.52 & 0.11 \\
3285.0 & 4 & 0 & 2 & 0.24 & 0.39 & 0.52 & 0.11 \\
4380.0 & 2 & 1 & 0 & 0.12 & 0.81 & 0.59 & 0.02 \\
4562.5 & 1 & 1 & 0 & 0.00 & Inf & NA & NA \\
\bottomrule()
\end{longtable}

\begin{figure}

{\centering \includegraphics{paper_files/figure-pdf/fig-surv-1.pdf}

}

\caption{\label{fig-surv}Kaplan Meier curve with the black line
representing the survival probability and with grey shading indicating
the 95\% confidence interval. Cross icons represent right-censoring}

\end{figure}

A comparative box plot of survival times by variant type shows
significant outliers for missense+splice and missense variations.
However, even within missense homozygous variants the survival time
varies greatly. Other missense variants, such as heterozygous missense
with gene deletion do not show the same increased survival time as noted
by \citet{magini2019loss}.

\includegraphics{paper_files/figure-pdf/unnamed-chunk-14-1.pdf}

A log-rank test was carried out to compare the survival curves of the
various gene variants and showed a significant difference in the
stratified survival curves (\(\chi^2 = 14.8, p = 0.04)\) however due to
the small sample size, several variants had expected counts less than 5
which would create a violation in assumptions for the test.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The use of Multiple Correspondence Analysis confirmed the heterogeneous
variation in the data with the first four principal axes being required
to explain more than 50\% of the variation in the data set. The
examination of key contributing factors, as well as a visual inspection
of the data projected into this lower dimensional space indicate the
primary sources of variation in the clinical phenotype are explained by
a few individuals with very different clinical features. This has
confounded the typical or classical presentation.

The cluster analysis does not aim to group individuals that are alike,
but rather the clinical features themselves and how these contrasting
clusters account for variation in the data. This highlights common
sources of variation on the clinical phenotype of the condition itself,
rather than typical coincident features of any one child.

After excluding two outlying, remarkable cases this analysis clearly
identifies the hallmark features of microcephaly, hypomyelination,
arthrogryposis, simplified gyral pattern and developmental delay. Of
significant note is the presence of an adjacent cluster of related
features that are not as well described in The Human Phenotype Ontology
\citep{kohler2021human} as shown in Table~\ref{tbl-hpo}. We identify a
prolific feature of `respiratory distress' as an expression more
associated with other features such as vocal cord palsy, tracheostomy,
tube feeding and GERD which are present in the case reported data, but
not well described in HPO. This represents a related but distinct
grouping of severe features that associated with this condition.

Other cluster groupings identify outlying features that are expressed in
individuals or specific families. Many of the outlying features are
specific facial or anatomical dysmorphisms. While course facial features
are a noted and core feature, those with a milder course had clinical
reports that articulated these dysmorphisms in much more detail. On a
similar note, the reporting of `talipes' as a condition which seems
distinct as a source of variation appearing as an isolated feature
without the typical microcephaly and hypomyelination. However, there was
significant heterogeneity in how feet contractures were reported in
clinical case reports. In the data pre-processing, a separate feature
was created for reported cases with rocker bottom feet or congenital
vertical talus versus those with clubfeet/talipes. In practice its not
clear how these diagnoses were arrived at, and given the non-specific
nature of this feature, care needs to be taken in distinguishing between
similar manifesting features in the literature where possible. This is
particularly clear as rocker bottom feet was found closely related in
our analysis to other core features despite the prevalence in HPO being
quite low (2/21), suggesting possible under-reporting or confounded
reporting of this feature. Another feature that clusters within the core
identified features is bilateral cleft lip. This condition has a
reported frequency of just 2/21 (Table~\ref{tbl-hpo}) and may be an
emerging feature for a more typical presentation given its close
relationship with other classic features. Not all features in the data
were included in this analysis, particular if isolated to just one case.
In this circumstance, these terms in Table~\ref{tbl-hpo} have been
labelled as `Isolated'.

While the number of clusters chosen all represent valid partition of the
data, some metrics indicated an optimal selection of 2 clusters. This
type of partition is useful when segregating outliers from the rest of
the data and is valid in some applications. In this study, a larger
number was selected to align the goals of the analysis which was to
uncover distinct groups of features.

In terms of patient survival, the first few months show a sharp decline
in survival probability confirming early demise in the infant period as
a key factor. A small number of cases show much longer survival into the
childhood and in some cases into the second decade of life. This work
identifies a one year survival rate of 0.403 (0.252, 0.644). A naive
estimate, using just the number of deceased patients at this time as
proportion of the study population is 0.57. This highlights the
importance of applying a survival analysis technique that incorporates
censoring so as not to erroneously overestimate the survival
probability. The current literature suggests the type of gene variant
has an impact on survival time, which seems to be supported in this
work. Through a comparison of multiple survival curves, stratified by
variation type, the limited sample size prevented other meaningful
analysis such as log-rank tests or cox proportional hazard modelling
which are likely under-powered when comparing multiple groups.

Despite NEDMABA exhibiting a diverse and heterogeneous clinical
phenotype, our analysis was able to identify meaningful substructures in
this data through the use of dimension reduction techniques (MCA) and
cluster analysis. This has isolated the classic NEDMABA phenotype of
microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern
and developmental delay. Furthermore it has isolated additional distinct
groupings of features that are exhibited in patients such as vocal cord
paralysis with tracheostomy, swallowing dysfunction, tube feeding and
GERD. Furthermore, we were able to quantify the expected baseline
survival probabilities for children with this condition as well as
quantifying the significant variation in these estimates. Future work is
required with the benefit of more cases, to better understand the
differences in survival curves for other factors associated with
genotype.

\newpage

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

\hypertarget{tbl-hpo}{}
\setlength{\LTpost}{0mm}
\begin{longtable}{llrl}
\caption{\label{tbl-hpo}A mapping table of phenotype terms from The Human Phenotype Ontology
(HPO) with clusters defined in this analysis. Not all clinical features
defined in the analysis are present in the HPO terms }\tabularnewline

\toprule
\multicolumn{3}{c}{HPO Data} & Derived Clusters \\ 
\cmidrule(lr){1-3} \cmidrule(lr){4-4}
HPO Term Name & Category & Frequency & \textbf{Cluster}\textsuperscript{1} \\ 
\midrule
Adducted thumb & Limbs & 1/21 & Isolated \\ 
Simplified gyral pattern & Nervous System & 12/18 & Core \\ 
Hypertonia & Musculature & 9/15 & Core \\ 
Seizure & Nervous System & 10/17 & Core \\ 
Hypotonia & Musculature & 4/15 & Core \\ 
Global developmental delay & Nervous System & 7/7 & Core \\ 
Highly arched eyebrow & Skin, Hair, and Nails & 2/21 & Isolated \\ 
Cerebellar hypoplasia & Nervous System & 10/21 & Core \\ 
Neonatal respiratory distress & Respiratory System & 14/17 & Adjacent \\ 
Bifid uvula & Head and neck & 1/21 & Isolated \\ 
Deep philtrum & Head and neck & 1/21 & Isolated \\ 
CNS hypomyelination & Nervous System & 7/19 & Core \\ 
Lumbar hypertrichosis & Skin, Hair, and Nails & 1/21 & Isolated \\ 
Birth length less than 3rd percentile & Growth & 6/11 & Core \\ 
Hypoplasia of the brainstem & Nervous System & 3/20 & Isolated \\ 
Tented upper lip vermilion & Head and neck & 3/21 & Isolated \\ 
Hypotelorism & Eye & 1/21 & Isolated \\ 
Short palpebral fissure & Head and neck & 4/21 & Outlying \\ 
Bilateral cleft lip & Head and neck & 2/21 & Core \\ 
Diabetes mellitus & Endocrine & 2/19 & Isolated \\ 
Single transverse palmar crease & Skin, Hair, and Nails & 1/21 & Isolated \\ 
Epicanthus & Head and neck & 5/21 & Outlying \\ 
Low anterior hairline & Head and neck & 2/21 & Outlying \\ 
Arthrogryposis multiplex congenita & Connective tissue & 17/20 & Core \\ 
Progressive microcephaly & Nervous System & 9/10 & Core \\ 
Microcephaly & Nervous System & 15/21 & Core \\ 
Thin upper lip vermilion & Head and neck & 2/21 & Outlying \\ 
Thin vermilion border & Head and neck & 3/21 & Isolated \\ 
Intrauterine growth retardation & Growth & 12/19 & Core \\ 
Posteriorly rotated ears & Ear & 2/21 & Adjacent \\ 
Narrow forehead & Head and neck & 1/21 & Isolated \\ 
Sloping forehead & Head and neck & 4/21 & Isolated \\ 
Smooth philtrum & Head and neck & 2/21 & Isolated \\ 
Short philtrum & Head and neck & 2/21 & Outlying \\ 
Premature birth & Prenatal and Birth & 7/18 & Core \\ 
Mandibular prognathia & Head and neck & 2/21 & Outlying \\ 
Wide intermamillary distance & Breast & 1/21 & Isolated \\ 
Depressed nasal bridge & Head and neck & 1/21 & Outlying \\ 
Downslanted palpebral fissures & Head and neck & 2/21 & Outlying \\ 
Short neck & Head and neck & 1/21 & Isolated \\ 
Bulbous nose & Head and neck & 2/21 & Isolated \\ 
Protruding ear & Ear & 2/21 & Outlying \\ 
Rocker bottom foot & Limbs & 2/21 & Core \\ 
\bottomrule
\end{longtable}
\begin{minipage}{\linewidth}
\textsuperscript{1}Clusters defined in this analysis mapped to HPO Terms\\
Source: The Human Phenotype Ontology (HPO) https://hpo.jax.org/app/browse/disease/OMIM:618622 \\
Reference: Köhler S, Gargano M, Matentzoglu N, Carmody LC, Lewis-Smith D, Vasilevsky NA, Danis D, Balagura G, Baynam G, Brower AM, Callahan TJ, Chute CG, Est JL, Galer PD, Ganesan S, Griese M, Haimel M, Pazmandi J, Hanauer M, Harris NL, Hartnett MJ, Hastreiter M, Hauck F, He Y, Jeske T, Kearney H, Kindle G, Klein C, Knoflach K, Krause R, Lagorce D, McMurry JA, Miller JA, Munoz-Torres MC, Peters RL, Rapp CK, Rath AM, Rind SA, Rosenberg AZ, Segal MM, Seidel MG, Smedley D, Talmy T, Thomas Y, Wiafe SA, Xian J, Yüksel Z, Helbig I, Mungall CJ, Haendel MA, Robinson PN. \emph{The Human Phenotype Ontology in 2021.} Nucleic Acids Res. 2021 Jan 8;49(D1):D1207-D1217. doi: {10.1093/nar/gkaa1043}. PMID: 33264411; PMCID: PMC7778952.\\
\end{minipage}

\hypertarget{tbl-profile}{}
\begin{longtable}[]{@{}rl@{}}
\caption{\label{tbl-profile}Assignment of clinical features with their
cluster. These clusters represent similar features projected onto the
first four MCA principal axes. Clusters that are on opposing sides of an
axes contract with each other and account for the largest amount of
variation in the data.}\tabularnewline
\toprule()
cluster & variable \\
\midrule()
\endfirsthead
\toprule()
cluster & variable \\
\midrule()
\endhead
1 & polyhydramnios \\
1 & short\_palpebral\_fissures \\
1 & patent\_ductus\_arteriosus \\
1 & thin\_lips \\
1 & prognathism \\
1 & pectus\_excavatum \\
1 & respiratory\_infections \\
2 & vocal\_cord\_palsy \\
2 & respiratory\_distress \\
2 & tracheostomy \\
2 & posteriorly\_angulated\_ears \\
2 & upper\_lip\_dysmorph \\
2 & tube\_feed \\
2 & gerd \\
2 & decreased\_craniofacial\_ratio \\
2 & short\_upturned\_nose \\
2 & borderline\_small\_brainstem \\
2 & enlarged\_ventricles \\
3 & epicanthal\_folds \\
3 & receding\_forehead \\
3 & spastic\_tetraparesis \\
3 & intubation \\
3 & low\_anterior\_hairline \\
3 & short\_philtrum \\
3 & hirsutism \\
4 & microcephaly \\
4 & sgp \\
4 & iugr \\
4 & talipes \\
4 & rocker\_bottom\_feet \\
4 & contractures\_hands\_deform \\
4 & arthrogryposis\_contractures\_other \\
4 & cleft\_lip \\
4 & coarse\_facial\_features \\
4 & hypomyelination \\
4 & cerebellar\_hypoplasia \\
4 & small\_anterior\_fontanel \\
4 & thin\_corpus\_collosum \\
4 & feeding\_swallow\_dysfxn \\
4 & hearing \\
4 & seizure \\
4 & agenesis\_of\_corpus\_callosum \\
4 & severe\_developmental\_delay \\
4 & at\_birth\_sga \\
4 & lissencephaly\_and\_cerebellar\_hypoplasia \\
4 & eeg\_normal \\
5 & hypercapnia \\
5 & inspiratory\_stridor \\
5 & oligohydramnios \\
5 & prominent\_ears \\
5 & dysmorphic\_features \\
5 & tracheo\_laryngomalacia \\
\bottomrule()
\end{longtable}

\hypertarget{tbl-dict}{}
\begin{longtable}{ll}
\caption{\label{tbl-dict}Data dictionary of the final data set used for the analysis. This
combines the case reports and supplied data for available and references
studies. The data has been transformed into a tidy format with
formatting applied. }\tabularnewline

\toprule
variable & type \\ 
\midrule
study & character \\ 
id & character \\ 
family & character \\ 
individual & factor \\ 
deceased & logical \\ 
survival\_time & numeric \\ 
variant\_type & factor \\ 
locus\_1 & numeric \\ 
locus\_2 & numeric \\ 
gender & factor \\ 
ethnicity & character \\ 
consanguineous & logical \\ 
termination & logical \\ 
birth\_gestation & numeric \\ 
route\_vaginal\_vs\_c\_sect & factor \\ 
birth\_weight & numeric \\ 
birth\_ofc & numeric \\ 
birth\_length & numeric \\ 
age\_at\_demise & numeric \\ 
age\_at\_last\_follow\_up & numeric \\ 
microcephaly\_ind & factor \\ 
sgp\_ind & factor \\ 
iugr\_ind & factor \\ 
talipes\_ind & factor \\ 
rocker\_bottom\_feet\_ind & factor \\ 
contractures\_hands\_deform\_ind & factor \\ 
vocal\_cord\_palsy\_ind & factor \\ 
arthrogryposis\_contractures\_other\_ind & factor \\ 
hypercapnia\_ind & factor \\ 
cleft\_lip\_ind & factor \\ 
inspiratory\_stridor\_ind & factor \\ 
polyhydramnios\_ind & factor \\ 
coarse\_facial\_features\_ind & factor \\ 
oligohydramnios\_ind & factor \\ 
respiratory\_distress\_ind & factor \\ 
tracheostomy\_ind & factor \\ 
prominent\_ears\_ind & factor \\ 
posteriorly\_angulated\_ears\_ind & factor \\ 
prominent\_philtrum\_ind & factor \\ 
upper\_lip\_dysmorph\_ind & factor \\ 
hypomyelination\_ind & factor \\ 
cerebellar\_hypoplasia\_ind & factor \\ 
short\_palpebral\_fissures\_ind & factor \\ 
epicanthal\_folds\_ind & factor \\ 
tube\_feed\_ind & factor \\ 
dysmorphic\_features\_ind & factor \\ 
patent\_ductus\_arteriosus\_ind & factor \\ 
small\_anterior\_fontanel\_ind & factor \\ 
receding\_forehead\_ind & factor \\ 
tracheo\_laryngomalacia\_ind & factor \\ 
gerd\_ind & factor \\ 
thin\_corpus\_collosum\_ind & factor \\ 
feeding\_swallow\_dysfxn\_ind & factor \\ 
seizures\_age\_started & character \\ 
seizure\_type\_s & character \\ 
eeg & character \\ 
sleeping\_problem & character \\ 
behavioral\_issues & character \\ 
skeletal\_changes\_scoliosis & numeric \\ 
hip\_luxation & numeric \\ 
development\_head\_control & character \\ 
sitting\_unsupported & character \\ 
crawling & character \\ 
walking & character \\ 
speech & character \\ 
iq\_int\_disability & character \\ 
cp & character \\ 
performance & character \\ 
neuro\_exam\_eye\_mvts\_nystagmus & character \\ 
strabismus & character \\ 
feeding\_swallow\_dysfxn & character \\ 
gag & character \\ 
tone & character \\ 
strength & character \\ 
sensory\_deficits & character \\ 
abnormal\_mvts & character \\ 
dtr & character \\ 
other\_neuro\_signs & character \\ 
sensory\_vision & character \\ 
hearing & character \\ 
other\_organs & character \\ 
endocrinology & character \\ 
functional\_tests & character \\ 
metabolic\_investigation & character \\ 
cardiology & character \\ 
seizure & numeric \\ 
agenesis\_of\_corpus\_callosum & factor \\ 
abnormal\_pattern\_of\_cerebral\_sulci & factor \\ 
left\_sided\_gallbladder & factor \\ 
suspected\_brain\_stem\_dysfunction & factor \\ 
spastic\_tetraparesis & factor \\ 
severe\_developmental\_delay & factor \\ 
acidosis & factor \\ 
small\_thorax & factor \\ 
cyanosis & factor \\ 
intubation & factor \\ 
apnea & factor \\ 
mr\_is\_reportedly\_with\_migrational\_defect & factor \\ 
at\_birth\_sga & factor \\ 
epilepsy\_and\_apnea & factor \\ 
choanal\_atresia\_surgeries\_for\_transposition\_of\_the\_great\_arteries & factor \\ 
interventricular\_and\_interatrial\_defects & factor \\ 
smooth\_philtrum & factor \\ 
thin\_lips & factor \\ 
bilateral\_simian\_creases & factor \\ 
low\_anterior\_hairline & factor \\ 
short\_philtrum & factor \\ 
prominent\_nose\_bridge & factor \\ 
depressed\_nasal\_bridge & factor \\ 
prominent\_suture\_ridges & factor \\ 
hypotelorism & factor \\ 
hypertrichosis\_of\_lower\_back & factor \\ 
decreased\_craniofacial\_ratio & factor \\ 
there\_is\_overlapping\_of\_the\_sutures & factor \\ 
bitemporal\_narrowing & factor \\ 
short\_upturned\_nose & factor \\ 
short\_neck & factor \\ 
upward\_sweep\_of\_hair & factor \\ 
arched\_eye\_brows & factor \\ 
bulbous\_nose & factor \\ 
prominent\_lower\_lip & factor \\ 
broad\_chin & factor \\ 
prognathism & factor \\ 
asd & factor \\ 
left\_multicystic\_dysplastic\_kidney & factor \\ 
anteriorly\_placed\_anus & factor \\ 
dilated\_cardiomyopathy & factor \\ 
cryptorchidism & factor \\ 
hirsutism & factor \\ 
intestinal\_malrotation & factor \\ 
dysphagia & factor \\ 
intermittent\_sinus\_bradycardia & factor \\ 
bilateral\_staphylomas & factor \\ 
lissencephaly & factor \\ 
echogenic\_choroid\_plexus\_on\_head\_us & factor \\ 
head\_us\_mild\_ventriculomegaly & factor \\ 
brain\_parenchyma\_appears\_homogeneous\_without\_evidence\_of\_focal\_lesion & factor \\ 
hypoplastic\_cerebellum & factor \\ 
borderline\_small\_brainstem & factor \\ 
abnormal\_cerebellar\_folia & factor \\ 
mild\_vermian\_hypoplasia & factor \\ 
asymmetry\_of\_lateral\_ventricles & factor \\ 
lissencephaly\_and\_cerebellar\_hypoplasia & factor \\ 
signs\_of\_hypoxic\_encephalopathy & factor \\ 
mild\_dilation\_of\_virchow\_robins\_spaces & factor \\ 
pectus\_excavatum & factor \\ 
widely\_spaced\_nipples & factor \\ 
respiratory\_infections & factor \\ 
enlarged\_ventricles & factor \\ 
eeg\_normal & numeric \\ 
bifid\_uvula & factor \\ 
under\_developed\_cerebellar\_inferior\_vermis & factor \\ 
\bottomrule
\end{longtable}

\newpage

\renewcommand\refname{References}
  \bibliography{bibliography.bib}

\end{document}

### 2022-09-07 paper/paper.qmd
rebuild with edits

@@
 The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and clinically diverse phenotype.
-A gap exists in the understanding the associations of the heterogenous features present in the clinical phenotype and the expected survival probabilities of affected individuals.
-This is driven in part to the paucity of analysis-ready data in reported cases.
-This research aims to collate and standardise available case reports to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA.
-To overcome the challenge of sparse, multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation.
+A gap exists in the understanding of associations present in the heterogenous features of the clinical phenotype, and the expected survival probabilities of affected individuals.
+This is driven in part to the paucity of analysis-ready data on reported cases.
+This research aims to collate and standardise available case reports, to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA.
+To overcome the challenge of multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation.
 To quantify more accurate survival probabilities, Kaplan-Meier estimation is formulated to account for censoring in the data.
-The analysis identified the classic phenotype for this condition, as well as one other distinct feature cluster relating to findings of vocal cord paralysis and respiratory failure.
-The survival probability for those affected was found to decline sharply in early infancy with median survival of 150 days, but with some as high as 12.5 years.
+The analysis correctly identified the classic phenotype for this condition, as well as one other distinct feature cluster relating to findings of vocal cord paralysis and respiratory failure.
+Several additional outlying and isolated clinical features were also identified.
+The survival probability for those affected was found to decline sharply in early infancy with median survival of 150 days, but with some surviving as long as 12.5 years.
 This wide range of outcomes is provisionally associated with variant type however interpretation of these results are guarded based on very low sample sizes.
-Despite challenges with sparse and inconsistent data, this analysis represents the first of its kind to help describe associations and trajectories of individuals with this condition and can provide clinicians and genetic counsellors with better information to aide in decision making and support for families.
+This analysis represents the first of its kind to help describe associations and trajectories of individuals with this condition, despite challenges with sparse and inconsistent data.
+This analysis can provide clinicians and genetic counsellors with better information to aide in decision making and support for families with this rare condition.
 keywords: 
@@
 
-The data on subjects introduced above is subset to include only variables that can be transformed into binary indicator variables that represent the presence or absence of a given clinical feature column.
-Only those features that appeared in more than one case were included to minimise the influence of non-related features in this analysis.
+The data on subjects introduced above is subset to include only variables that can be transformed into binary indicator variables that represent the presence or absence of a given clinical feature.
+Only those features that appeared in more than one case were included to minimise the influence of isolated features in this analysis.
 
@@
 The MCA coordinates for the first 4 principal dimensions were subject to clustering in order to find groups of features that were internally homogeneous and externally heterogeneous.
-This was computed in R using the `clValid` package [@clvalid].
+This was computed in `R` using the `clValid` package [@clvalid].
 
@@
 
-The baseline estimator of all individuals was calculated using the *survival* package in R [@survival-book].
+The baseline estimator of all individuals was calculated using the *survival* package in `R` [@survival-book].
 Next the Kaplan-Meier estimator was stratified by the type of genetic variation reported in the literature.
-A log-rank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* R package.
+A log-rank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* `R` package.
 
@@
 
-Repeating MCA for the remaining subjects shows the first dimension ($\lambda_!
-= 0.18$) accounts for 17.88% of the variance in the data with the first 4 dimensions of the MCA analysis accounting for 54% of the variance (@tbl-eig).
+Repeating MCA for the remaining subjects shows the first dimension ($\lambda_1 = 0.18$) accounts for 17.88% of the variance in the data with the first 4 dimensions of the MCA analysis accounting for 54% of the variance (@tbl-eig).
 
@@
 
-Three other smaller satellite clusters exists where highly distinctive features were associated with an individual or family.
+Three other smaller satellite clusters exists where highly distinctive features were associated with a single individual or family.
 A full articulation of feature to cluster assignment is given in @tbl-profile.
@@
 #| label: fig-clvalid
-#| fig-cap: "cluster validation analysis showing various metrics for cluster evaluation from 2 to 10 clusters.
+#| fig-cap: "cluster validation analysis showing various metrics for cluster evaluation from 2 to 5 clusters.
 The most optimal number of clusters are selected by picking the lowest value from one or many of these comparable metrics."
@@
 The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with median survival time calculated as 150 days.
-At time 0 the survival probability was 0.82 (0.69, 0.98).
-Survival probability after one year at 0.403 (0.252, 0.644) and survival probability at close to ten years of 0.24 (0.11, 0.52).
+At time 0 the survival probability was 0.82 (95% CI: 0.69, 0.98).
+Survival probability after one year at 0.403 (95% CI: 0.252, 0.644) and survival probability at close to ten years of 0.24 (95% CI: 0.11, 0.52).
 
@@
 The cluster analysis does not aim to group individuals that are alike, but rather the clinical features themselves and how these contrasting clusters account for variation in the data.
-This highlights common sources of variation on the clinical phenotype of the condition itself, rather than typical coincident features of any one child.
+This highlights common sources of variation in the clinical phenotype of the condition itself, rather than typical coincident features of any one child.
 
-After excluding two outlying, remarkable cases this analysis clearly identifies the hallmark features of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay.
+After excluding two outlying, remarkable cases this analysis clearly identifies a core cluster of the hallmark features of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay.
 Of significant note is the presence of an adjacent cluster of related features that are not as well described in The Human Phenotype Ontology [@kohler2021human] as shown in @tbl-hpo.
 We identify a prolific feature of 'respiratory distress' as an expression more associated with other features such as vocal cord palsy, tracheostomy, tube feeding and GERD which are present in the case reported data, but not well described in HPO.
-This represents a related but distinct grouping of severe features that associated with this condition.
+This represents a related but distinct grouping of severe features associated with this condition.
 
-Other cluster groupings identify outlying features that are expressed in individuals or specific families.
-Many of the outlying features are specific facial or anatomical dysmorphisms.
-While course facial features are a noted and core feature, those with a milder course had clinical reports that articulated these dysmorphisms in much more detail.
-On a similar note, the reporting of 'talipes' as a condition which seems distinct as a source of variation appearing as an isolated feature without the typical microcephaly and hypomyelination.
+Other smaller clusters identify outlying features that are expressed in individuals or specific families.
+Many of the outlying features are highly specific facial or anatomical dysmorphisms.
+While coarse facial features are a noted and core presentation, those with a milder course had clinical reports that articulated these dysmorphisms in much more detail.
+On a similar note, the reporting of 'talipes' as a condition which seems distinct as a source of variation, appearing as an isolated feature without the typical microcephaly and hypomyelination.
 However, there was significant heterogeneity in how feet contractures were reported in clinical case reports.
 In the data pre-processing, a separate feature was created for reported cases with rocker bottom feet or congenital vertical talus versus those with clubfeet/talipes.
-In practice its not clear how these diagnoses were arrived at, and given the non-specific nature of this feature, care needs to be taken in distinguishing between similar manifesting features in the literature where possible.
+In practice its not clear how these diagnoses were arrived at, and given the non-specific nature of these features, care needs to be taken in distinguishing between similar manifesting features in the literature where possible.
 This is particularly clear as rocker bottom feet was found closely related in our analysis to other core features despite the prevalence in HPO being quite low (2/21), suggesting possible under-reporting or confounded reporting of this feature.
@@
 In terms of patient survival, the first few months show a sharp decline in survival probability confirming early demise in the infant period as a key factor.
-A small number of cases show much longer survival into the childhood and in some cases into the second decade of life.
-This work identifies a one year survival rate of 0.403 (0.252, 0.644).
+A small number of cases show much longer survival into childhood and in some cases into the second decade of life.
+This work identifies a one year survival rate of 0.403 (95% CI: 0.252, 0.644).
 A naive estimate, using just the number of deceased patients at this time as proportion of the study population is 0.57.

### 2022-09-07 paper/paper.tex
rebuild with edits

@@
 so far exhibiting a severe and clinically diverse phenotype. A gap
-exists in the understanding the associations of the heterogenous
-features present in the clinical phenotype and the expected survival
+exists in the understanding of associations present in the heterogenous
+features of the clinical phenotype, and the expected survival
 probabilities of affected individuals. This is driven in part to the
-paucity of analysis-ready data in reported cases. This research aims to
-collate and standardise available case reports to analyse and identify
+paucity of analysis-ready data on reported cases. This research aims to
+collate and standardise available case reports, to analyse and identify
 meaningful clusters in the clinical phenotype, and to quantify the
 survival probability for children with NEDMABA. To overcome the
-challenge of sparse, multidimensional data on very few subjects, we
-employ Multiple Correspondence Analysis (MCA) as a dimension reduction
+challenge of multidimensional data on very few subjects, we employ
+Multiple Correspondence Analysis (MCA) as a dimension reduction
 technique, which is then subject to cluster analysis and interpretation.
@@
 estimation is formulated to account for censoring in the data. The
-analysis identified the classic phenotype for this condition, as well as
-one other distinct feature cluster relating to findings of vocal cord
-paralysis and respiratory failure. The survival probability for those
-affected was found to decline sharply in early infancy with median
-survival of 150 days, but with some as high as 12.5 years. This wide
-range of outcomes is provisionally associated with variant type however
-interpretation of these results are guarded based on very low sample
-sizes. Despite challenges with sparse and inconsistent data, this
-analysis represents the first of its kind to help describe associations
-and trajectories of individuals with this condition and can provide
-clinicians and genetic counsellors with better information to aide in
-decision making and support for families.
+analysis correctly identified the classic phenotype for this condition,
+as well as one other distinct feature cluster relating to findings of
+vocal cord paralysis and respiratory failure. Several additional
+outlying and isolated clinical features were also identified. The
+survival probability for those affected was found to decline sharply in
+early infancy with median survival of 150 days, but with some surviving
+as long as 12.5 years. This wide range of outcomes is provisionally
+associated with variant type however interpretation of these results are
+guarded based on very low sample sizes. This analysis represents the
+first of its kind to help describe associations and trajectories of
+individuals with this condition, despite challenges with sparse and
+inconsistent data. This analysis can provide clinicians and genetic
+counsellors with better information to aide in decision making and
+support for families with this rare condition.
 \end{abstract}
@@
 \end{keyword}
-\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 variables that can be transformed into binary indicator variables that
-represent the presence or absence of a given clinical feature column.
-Only those features that appeared in more than one case were included to
-minimise the influence of non-related features in this analysis.
+represent the presence or absence of a given clinical feature. Only
+those features that appeared in more than one case were included to
+minimise the influence of isolated features in this analysis.
 
@@
 clustering in order to find groups of features that were internally
-homogeneous and externally heterogeneous. This was computed in R using
-the \texttt{clValid} package \citep{clvalid}.
+homogeneous and externally heterogeneous. This was computed in
+\texttt{R} using the \texttt{clValid} package \citep{clvalid}.
 
@@
 The baseline estimator of all individuals was calculated using the
-\emph{survival} package in R \citep{survival-book}. Next the
+\emph{survival} package in \texttt{R} \citep{survival-book}. Next the
 Kaplan-Meier estimator was stratified by the type of genetic variation
@@
 differences in the survival curves using methods from
-\citet{harrington1982class}, again implemented in the \emph{survival} R
-package.
+\citet{harrington1982class}, again implemented in the \emph{survival}
+\texttt{R} package.
 
@@
 Repeating MCA for the remaining subjects shows the first dimension
-(\(\lambda_! = 0.18\)) accounts for 17.88\% of the variance in the data
+(\(\lambda_1 = 0.18\)) accounts for 17.88\% of the variance in the data
 with the first 4 dimensions of the MCA analysis accounting for 54\% of
@@
 Three other smaller satellite clusters exists where highly distinctive
-features were associated with an individual or family. A full
+features were associated with a single individual or family. A full
 articulation of feature to cluster assignment is given in
@@
 \caption{\label{fig-clvalid}cluster validation analysis showing various
-metrics for cluster evaluation from 2 to 10 clusters. The most optimal
+metrics for cluster evaluation from 2 to 5 clusters. The most optimal
 number of clusters are selected by picking the lowest value from one or
@@
 days) with median survival time calculated as 150 days. At time 0 the
-survival probability was 0.82 (0.69, 0.98). Survival probability after
-one year at 0.403 (0.252, 0.644) and survival probability at close to
-ten years of 0.24 (0.11, 0.52).
+survival probability was 0.82 (95\% CI: 0.69, 0.98). Survival
+probability after one year at 0.403 (95\% CI: 0.252, 0.644) and survival
+probability at close to ten years of 0.24 (95\% CI: 0.11, 0.52).
 
@@
 clusters account for variation in the data. This highlights common
-sources of variation on the clinical phenotype of the condition itself,
+sources of variation in the clinical phenotype of the condition itself,
 rather than typical coincident features of any one child.
@@
 After excluding two outlying, remarkable cases this analysis clearly
-identifies the hallmark features of microcephaly, hypomyelination,
-arthrogryposis, simplified gyral pattern and developmental delay. Of
-significant note is the presence of an adjacent cluster of related
-features that are not as well described in The Human Phenotype Ontology
-\citep{kohler2021human} as shown in Table~\ref{tbl-hpo}. We identify a
-prolific feature of `respiratory distress' as an expression more
-associated with other features such as vocal cord palsy, tracheostomy,
-tube feeding and GERD which are present in the case reported data, but
-not well described in HPO. This represents a related but distinct
-grouping of severe features that associated with this condition.
-
-Other cluster groupings identify outlying features that are expressed in
+identifies a core cluster of the hallmark features of microcephaly,
+hypomyelination, arthrogryposis, simplified gyral pattern and
+developmental delay. Of significant note is the presence of an adjacent
+cluster of related features that are not as well described in The Human
+Phenotype Ontology \citep{kohler2021human} as shown in
+Table~\ref{tbl-hpo}. We identify a prolific feature of `respiratory
+distress' as an expression more associated with other features such as
+vocal cord palsy, tracheostomy, tube feeding and GERD which are present
+in the case reported data, but not well described in HPO. This
+represents a related but distinct grouping of severe features associated
+with this condition.
+
+Other smaller clusters identify outlying features that are expressed in
 individuals or specific families. Many of the outlying features are
-specific facial or anatomical dysmorphisms. While course facial features
-are a noted and core feature, those with a milder course had clinical
-reports that articulated these dysmorphisms in much more detail. On a
-similar note, the reporting of `talipes' as a condition which seems
-distinct as a source of variation appearing as an isolated feature
-without the typical microcephaly and hypomyelination. However, there was
-significant heterogeneity in how feet contractures were reported in
-clinical case reports. In the data pre-processing, a separate feature
-was created for reported cases with rocker bottom feet or congenital
-vertical talus versus those with clubfeet/talipes. In practice its not
-clear how these diagnoses were arrived at, and given the non-specific
-nature of this feature, care needs to be taken in distinguishing between
-similar manifesting features in the literature where possible. This is
-particularly clear as rocker bottom feet was found closely related in
-our analysis to other core features despite the prevalence in HPO being
-quite low (2/21), suggesting possible under-reporting or confounded
-reporting of this feature. Another feature that clusters within the core
-identified features is bilateral cleft lip. This condition has a
-reported frequency of just 2/21 (Table~\ref{tbl-hpo}) and may be an
-emerging feature for a more typical presentation given its close
-relationship with other classic features. Not all features in the data
-were included in this analysis, particular if isolated to just one case.
-In this circumstance, these terms in Table~\ref{tbl-hpo} have been
-labelled as `Isolated'.
+highly specific facial or anatomical dysmorphisms. While coarse facial
+features are a noted and core presentation, those with a milder course
+had clinical reports that articulated these dysmorphisms in much more
+detail. On a similar note, the reporting of `talipes' as a condition
+which seems distinct as a source of variation, appearing as an isolated
+feature without the typical microcephaly and hypomyelination. However,
+there was significant heterogeneity in how feet contractures were
+reported in clinical case reports. In the data pre-processing, a
+separate feature was created for reported cases with rocker bottom feet
+or congenital vertical talus versus those with clubfeet/talipes. In
+practice its not clear how these diagnoses were arrived at, and given
+the non-specific nature of these features, care needs to be taken in
+distinguishing between similar manifesting features in the literature
+where possible. This is particularly clear as rocker bottom feet was
+found closely related in our analysis to other core features despite the
+prevalence in HPO being quite low (2/21), suggesting possible
+under-reporting or confounded reporting of this feature. Another feature
+that clusters within the core identified features is bilateral cleft
+lip. This condition has a reported frequency of just 2/21
+(Table~\ref{tbl-hpo}) and may be an emerging feature for a more typical
+presentation given its close relationship with other classic features.
+Not all features in the data were included in this analysis, particular
+if isolated to just one case. In this circumstance, these terms in
+Table~\ref{tbl-hpo} have been labelled as `Isolated'.
 
@@
 in survival probability confirming early demise in the infant period as
-a key factor. A small number of cases show much longer survival into the
+a key factor. A small number of cases show much longer survival into
 childhood and in some cases into the second decade of life. This work
-identifies a one year survival rate of 0.403 (0.252, 0.644). A naive
-estimate, using just the number of deceased patients at this time as
-proportion of the study population is 0.57. This highlights the
+identifies a one year survival rate of 0.403 (95\% CI: 0.252, 0.644). A
+naive estimate, using just the number of deceased patients at this time
+as proportion of the study population is 0.57. This highlights the
 importance of applying a survival analysis technique that incorporates

### 2022-09-07 _targets.R

@@ -5,7 +5,7 @@ library(tarchetypes)
 tar_option_set(
   packages = c("SMPD4", 'clValid', 'tidyverse', 'ggrepel', 
                'cluster', 'survival', 'survminer', 'quarto',
-               'knitr', 'FactoMineR', 'factoextra', 'gt', 'withr'), 
+               'knitr', 'FactoMineR', 'factoextra', 'gt', 'withr', 'kohonen'), 
   format = "rds" 
 )
 

### 2022-09-07 _targets_packages.R

@@ -9,6 +9,7 @@ library(future)
 library(ggrepel)
 library(gt)
 library(knitr)
+library(kohonen)
 library(markdown)
 library(pingr)
 library(quarto)
@@ -21,3 +22,4 @@ library(survival)
 library(survminer)
 library(tidyverse)
 library(visNetwork)
+library(withr)

### 2022-09-09 paper/paper.qmd
more edits

@@
 ---
-title: Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies
+title: Clinical Phenotype and Survival Analysis in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in *SMPD4*
 author:
@@
 This is driven in part to the paucity of analysis-ready data on reported cases.
-This research aims to collate and standardise available case reports, to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA.
+This research aims to collate and standardise available case reports into a common dataset, to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA.
 To overcome the challenge of multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation.
-To quantify more accurate survival probabilities, Kaplan-Meier estimation is formulated to account for censoring in the data.
-The analysis correctly identified the classic phenotype for this condition, as well as one other distinct feature cluster relating to findings of vocal cord paralysis and respiratory failure.
+To account for censoring in the data Kaplan-Meier estimation is formulated calculate patient survival time.
+The analysis correctly detected the classic phenotype for this condition, as well as an additional distinct feature-cluster relating to findings of vocal cord paralysis, feeding dysfunction and respiratory failure.
 Several additional outlying and isolated clinical features were also identified.
 The survival probability for those affected was found to decline sharply in early infancy with median survival of 150 days, but with some surviving as long as 12.5 years.
-This wide range of outcomes is provisionally associated with variant type however interpretation of these results are guarded based on very low sample sizes.
-This analysis represents the first of its kind to help describe associations and trajectories of individuals with this condition, despite challenges with sparse and inconsistent data.
+This wide range of outcomes is provisionally associated with different variant types however this conclusion could not be validated based on very low sample sizes.
+This analysis represents the first of its kind to help describe associations and trajectories of individuals with this newly reported condition, despite challenges with sparse and inconsistent data.
 This analysis can provide clinicians and genetic counsellors with better information to aide in decision making and support for families with this rare condition.
@@
   - SMPD4  
-#journal: "An awesome journal"
+#journal: "Preprint"    
 date: last-modified
@@
 Further analysis to better describe rare genetic conditions was explored in @diaz2020phenotype in the context of large scale genotype-phenotype analysis.
-The authors argue patients with rare disorders (often in small samples) often present with varied symptoms that do not match exactly with the described phenotype.
+The authors argue patients with rare disorders (often in small samples) frequently present with varied symptoms that do not match exactly with the described phenotype.
 When considering this situation of low sample size data with many features, methods in the field of multivariate statistical analysis are commonly deployed.
@@
 
-In the case where two variables were formed from features that are considered to be synonymous, these were merged into one indicator column to prevent duplication e.g.
+In the case where synonymous terms are present, these indicator columns are consolidated and merged into one indicator column to prevent duplication e.g.
 {bilateral_cleft_lips, bilateral_cleft_lip, cleft_lip_b\_l} -\> {bilateral_cleft_lip}.
@@
 
-This format was preserved and other case studies identified in the literature [@ravenscroft2021neurogenetic; @monies2019lessons; @bijarnia2022growth; @ji2022case] were manually entered to conform to this template to allow the data to be combined for further analysis.
+This format was used as a template to align other written case reports identified in the literature [@ravenscroft2021neurogenetic; @monies2019lessons; @bijarnia2022growth; @ji2022case].
+These were manually entered to conform to this template to allow the data to be combined for further analysis.
 
@@
 The data on subjects introduced above is subset to include only variables that can be transformed into binary indicator variables that represent the presence or absence of a given clinical feature.
-Only those features that appeared in more than one case were included to minimise the influence of isolated features in this analysis.
+Only those features that appeared in more than one case report were included to minimise the influence of isolated features in this analysis.
 
 Dimension reduction on this wide data set was performed using Multiple Correspondence Analysis (MCA) (an analogy to Principle Component Analysis (PCA) for categorical data) [@le2010multiple] in order to arrive at a set of meaningfully small dimensions that account for most of the variation in the data.
+This computation was carried out using the *FactoMineR* R package [@factominer].
 
@@
 
-This computation of the Multiple Correspondence Analysis was computed using the *FactoMineR* R package [@factominer].
-
 The *contribution* of each category was computed for each of the principal axis per @le2010multiple. This metric identifies the proportion of variance of that axis due to the point and is defined as:
@@
 
-The coordinates on the first few principal axes were inspected for any outliers or unusual data.
+The coordinates on the first few principal axes were inspected for any outliers or unusual data before being graphically interpreted and subject to further analysis.
 
@@
 These metrics were selected to represent each data point (clinical feature) with a number of MCA dimensions.
-A good cluster solution is one that is robust and meaningful across all of these dimensions [@fig-clvalid].
-The optimal method and cluster number was selected based on a trade off between the best performing method using the evaluation metrics and a sensible partitioning of the data for this analysis [@tbl-eig].
+A good cluster solution is one that is robust and meaningful across all of these dimensions ([@fig-clvalid]).
+The optimal method and cluster number was selected based on a trade off between the best performing method using the evaluation metrics and a sensible partitioning of the data for this analysis ([@tbl-eig]).
 
@@
 
-The initial results of the MCA were inspected with a bi-plot and identified two subjects (17, 18) from the data that were clear outliers and sat significantly beyond the 95% confidence ellipse.
+The initial results of the MCA were inspected with a bi-plot and identified two subjects (17, 18) from the data that were clear outliers and sat significantly beyond the 95% confidence ellipse (@fig-biplotprelim).
 These individuals represented two twins from @magini2019loss who were described with a much milder phenotype and were subject to a missense variation in *SMPD4*.
 Further analysis of this influence is conducted below.
-For now, these subjects were removed to allow for more meaningful analysis.
+For now, these subjects were removed to allow for more meaningful analysis of other features.
 
@@
 
-The *contribution* of the categories were ranked within each of the first four MCA dimensions (@fig-ctr).
-The first dimension is dominated by highly specific facial dysmorphisms such as epicanthal folds, short philtrum and low anterior hairline.
-
+The *contribution* of the categories were ranked within each of the first four principal dimensions in @fig-ctr.
+The first dimension is dominated by isolated cases with highly specific facial dysmorphisms such as epicanthal folds, short philtrum and low anterior hairline.
 The second MCA dimension is categorised by feeding and respiratory dysfunction with vocal cord palsy, tracheostomy, tube feeding and GERD.
-
-Examining the bi-plot of the first two principal axes @fig-biplot illustrates the contrasting sources of variation between conditions relating to airway and swallowing function and predominantly facial features and structural brain anomalies such as as cerebellar hypoplasia, development delay.
-
+Examining the bi-plot of the first two principal axes (@fig-biplot) illustrates the contrasting sources of variation between conditions relating to airway and swallowing function and predominantly facial features and structural brain anomalies such as as cerebellar hypoplasia, development delay.
 The third dimension is characterised by atypical, complex features such as patent ductus arteriosus, respiratory infections, small for gestational age and further facial features.
@@
 
-Cluster validation measures resulted in either 2 or 5 clusters as the optimal choice.
-
-In order to partition the data into meaningful groupings, k-means with 5 clusters was selected. This reflects the optimal choice for both AD and FOM cluster metrics.
-
-A scatter plot of the cluster results @fig-clust projected on just the first two dimensions highlights a core cluster of classic features of NEDMABA such as microcephaly, hypomyelination, arthrogryposis and feet deformity and structural brain anomalies.
-A separate cluster is formed of airway and feeding related conditions involving vocal cord paralysis, tracheostomy, tube feeding and enlarged ventricles.
-
+Cluster validation measures resulted in either 2 or 5 clusters as the optimal choice (@tbl-opticlust).
+In order to partition the data into meaningful groupings, k-means with 5 clusters was selected.
+This reflects the optimal choice for both AD and FOM cluster metrics.
+A scatter plot of the cluster results shown in @fig-clust is projected on the first two principal dimensions highlights a core cluster of classic features of NEDMABA such as microcephaly, hypomyelination, arthrogryposis and feet deformity and structural brain anomalies.
+A separate and adjacent cluster is formed of airway and feeding related conditions involving vocal cord paralysis, tracheostomy, tube feeding and enlarged ventricles.
 Three other smaller satellite clusters exists where highly distinctive features were associated with a single individual or family.
@@
 
-A comparative box plot of survival times by variant type shows significant outliers for missense+splice and missense variations.
-However, even within missense homozygous variants the survival time varies greatly.
+A comparative box plot of survival times by variant type shows significant outliers for compound heterozygous *missense + splice* and homozygous *missense* variants.
+However, even within homozygous missense variants the survival time varies greatly.
 Other missense variants, such as heterozygous missense with gene deletion do not show the same increased survival time as noted by @magini2019loss.
@@
 
-A log-rank test was carried out to compare the survival curves of the various gene variants and showed a significant difference in the stratified survival curves ($\chi^2 = 14.8, p = 0.04)$ however due to the small sample size, several variants had expected counts less than 5 which would create a violation in assumptions for the test.
+A log-rank test was carried out to compare the survival curves of patients stratified by the various gene variants and showed a significant difference in the stratified survival curves ($\chi^2 = 14.8, p = 0.04)$ however due to the small sample size and large number of possible variants, several had expected counts less than 5 which would create a violation in assumptions for the test.
 
@@
 
-The use of Multiple Correspondence Analysis confirmed the heterogeneous variation in the data with the first four principal axes being required to explain more than 50% of the variation in the data set.
-The examination of key contributing factors, as well as a visual inspection of the data projected into this lower dimensional space indicate the primary sources of variation in the clinical phenotype are explained by a few individuals with very different clinical features.
-This has confounded the typical or classical presentation.
+Often, data collected in studies is fit for descriptive purposes but not for in-depth statistical analysis.
+This meta-review of current studies was able to consolidate and make publicly available, data sets that allow analysis to be conducted on emerging conditions such as NEDMABA.
 
-The cluster analysis does not aim to group individuals that are alike, but rather the clinical features themselves and how these contrasting clusters account for variation in the data.
-This highlights common sources of variation in the clinical phenotype of the condition itself, rather than typical coincident features of any one child.
+The use of dimension reduction techniques were shown to be successful in uncovering relevant and novel substructure in the data.
+This was particularly evident with the detection of the accepted typical phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay.
 
-After excluding two outlying, remarkable cases this analysis clearly identifies a core cluster of the hallmark features of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay.
-Of significant note is the presence of an adjacent cluster of related features that are not as well described in The Human Phenotype Ontology [@kohler2021human] as shown in @tbl-hpo.
-We identify a prolific feature of 'respiratory distress' as an expression more associated with other features such as vocal cord palsy, tracheostomy, tube feeding and GERD which are present in the case reported data, but not well described in HPO.
-This represents a related but distinct grouping of severe features associated with this condition.
+After excluding two outlying remarkable cases, this analysis clearly identifies the presence of an adjacent cluster of related features that are not as well described in The Human Phenotype Ontology [@kohler2021human] as shown in @tbl-hpo.
+We identify the prolific feature of 'respiratory distress' as an expression more commonly associated with *other* features such as vocal cord palsy, tracheostomy, tube feeding and Gastroesophageal reflux disease (GERD).
+These features are present in the case reported data, but not well described in HPO and help account for the variation seen in the clinical phenotype in the first few principal dimensions.
 
@@
 Many of the outlying features are highly specific facial or anatomical dysmorphisms.
-While coarse facial features are a noted and core presentation, those with a milder course had clinical reports that articulated these dysmorphisms in much more detail.
-On a similar note, the reporting of 'talipes' as a condition which seems distinct as a source of variation, appearing as an isolated feature without the typical microcephaly and hypomyelination.
+While coarse facial features are a noted and core presentation, those with a milder course had clinical reports that articulated these dysmorphisms in much more detail, which skewed the interpretation.
+
+On a similar note, talipes appears as an isolated feature without the typical microcephaly and hypomyelination.
 However, there was significant heterogeneity in how feet contractures were reported in clinical case reports.
-In the data pre-processing, a separate feature was created for reported cases with rocker bottom feet or congenital vertical talus versus those with clubfeet/talipes.
-In practice its not clear how these diagnoses were arrived at, and given the non-specific nature of these features, care needs to be taken in distinguishing between similar manifesting features in the literature where possible.
-This is particularly clear as rocker bottom feet was found closely related in our analysis to other core features despite the prevalence in HPO being quite low (2/21), suggesting possible under-reporting or confounded reporting of this feature.
-Another feature that clusters within the core identified features is bilateral cleft lip.
+While distinctions were made during data pre-processing between rocker-bottom feed and talipes, this likely to make intpretation somewhat difficult.
+This is particularly clear as rocker-bottom feet was found closely related in our analysis to other core features despite the prevalence in HPO being quite low (2/21), suggesting possible under-reporting or confounded reporting of this feature.
+
+Another feature clusted within core identified features is bilateral cleft lip.
 This condition has a reported frequency of just 2/21 (@tbl-hpo) and may be an emerging feature for a more typical presentation given its close relationship with other classic features.
+
+The heterogeneous variation in the data was clearly evident however, with the first four principal axes being required to explain more than 50% of the variation in the data set.
+Visual inspection of the data projected into this lower dimensional space indicate the primary sources of variation in the clinical phenotype are explained by a few individuals with very different clinical features.
+This has confounded the typical or classical presentation.
 Not all features in the data were included in this analysis, particular if isolated to just one case.
@@
 This type of partition is useful when segregating outliers from the rest of the data and is valid in some applications.
-In this study, a larger number was selected to align the goals of the analysis which was to uncover distinct groups of features.
+In this study, a larger number (5) was selected to align the goals of the analysis which was to uncover distinct groups of features.
 
@@
 This work identifies a one year survival rate of 0.403 (95% CI: 0.252, 0.644).
-A naive estimate, using just the number of deceased patients at this time as proportion of the study population is 0.57.
+On the other hand, a naive estimate using just the number of deceased patients at this time as proportion of the study population is 0.57 (16/28).
 This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to erroneously overestimate the survival probability.
 The current literature suggests the type of gene variant has an impact on survival time, which seems to be supported in this work.
-Through a comparison of multiple survival curves, stratified by variation type, the limited sample size prevented other meaningful analysis such as log-rank tests or cox proportional hazard modelling which are likely under-powered when comparing multiple groups.
+Through a comparison of multiple survival curves, stratified by variation type, the limited sample size prevented definitive conclusions with many cells having expected counts less than 5.
 
@@
 This has isolated the classic NEDMABA phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay.
-Furthermore it has isolated additional distinct groupings of features that are exhibited in patients such as vocal cord paralysis with tracheostomy, swallowing dysfunction, tube feeding and GERD.
+Furthermore it has isolated additional distinct groupings of features that are exhibited in patients such as vocal cord paralysis, swallowing dysfunction, tube feeding and GERD.
 Furthermore, we were able to quantify the expected baseline survival probabilities for children with this condition as well as quantifying the significant variation in these estimates.

### 2022-09-09 paper/paper.tex
more edits

@@
 \hypersetup{
-  pdftitle={Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies},
+  pdftitle={Clinical Phenotype and Survival Analysis in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD4},
   pdfauthor={Dean Marchiori},
@@
 \begin{frontmatter}
-\title{Statistical Analysis of Clinical Phenotype and Patient Survival
-in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and
-Structural Brain Anomalies}
+\title{Clinical Phenotype and Survival Analysis in Neurodevelopmental
+Disorder with Microcephaly, Arthrogryposis, and Structural Brain
+Anomalies Due to Bi-allelic Loss of Function Variants in \emph{SMPD4}}
 \author[]{Dean Marchiori%
@@
 paucity of analysis-ready data on reported cases. This research aims to
-collate and standardise available case reports, to analyse and identify
-meaningful clusters in the clinical phenotype, and to quantify the
-survival probability for children with NEDMABA. To overcome the
-challenge of multidimensional data on very few subjects, we employ
-Multiple Correspondence Analysis (MCA) as a dimension reduction
+collate and standardise available case reports into a common dataset, to
+analyse and identify meaningful clusters in the clinical phenotype, and
+to quantify the survival probability for children with NEDMABA. To
+overcome the challenge of multidimensional data on very few subjects, we
+employ Multiple Correspondence Analysis (MCA) as a dimension reduction
 technique, which is then subject to cluster analysis and interpretation.
-To quantify more accurate survival probabilities, Kaplan-Meier
-estimation is formulated to account for censoring in the data. The
-analysis correctly identified the classic phenotype for this condition,
-as well as one other distinct feature cluster relating to findings of
-vocal cord paralysis and respiratory failure. Several additional
-outlying and isolated clinical features were also identified. The
-survival probability for those affected was found to decline sharply in
-early infancy with median survival of 150 days, but with some surviving
-as long as 12.5 years. This wide range of outcomes is provisionally
-associated with variant type however interpretation of these results are
-guarded based on very low sample sizes. This analysis represents the
-first of its kind to help describe associations and trajectories of
-individuals with this condition, despite challenges with sparse and
-inconsistent data. This analysis can provide clinicians and genetic
-counsellors with better information to aide in decision making and
-support for families with this rare condition.
+To account for censoring in the data Kaplan-Meier estimation is
+formulated calculate patient survival time. The analysis correctly
+detected the classic phenotype for this condition, as well as an
+additional distinct feature-cluster relating to findings of vocal cord
+paralysis, feeding dysfunction and respiratory failure. Several
+additional outlying and isolated clinical features were also identified.
+The survival probability for those affected was found to decline sharply
+in early infancy with median survival of 150 days, but with some
+surviving as long as 12.5 years. This wide range of outcomes is
+provisionally associated with different variant types however this
+conclusion could not be validated based on very low sample sizes. This
+analysis represents the first of its kind to help describe associations
+and trajectories of individuals with this newly reported condition,
+despite challenges with sparse and inconsistent data. This analysis can
+provide clinicians and genetic counsellors with better information to
+aide in decision making and support for families with this rare
+condition.
 \end{abstract}
@@
 \end{keyword}
-\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
+\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, enhanced, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 genotype-phenotype analysis. The authors argue patients with rare
-disorders (often in small samples) often present with varied symptoms
-that do not match exactly with the described phenotype. When considering
-this situation of low sample size data with many features, methods in
-the field of multivariate statistical analysis are commonly deployed.
-Here the aim is to find substructure in the data, or a simple
+disorders (often in small samples) frequently present with varied
+symptoms that do not match exactly with the described phenotype. When
+considering this situation of low sample size data with many features,
+methods in the field of multivariate statistical analysis are commonly
+deployed. Here the aim is to find substructure in the data, or a simple
 representation of the multi-dimensional space. Methods, such as Multiple
@@
 
-In the case where two variables were formed from features that are
-considered to be synonymous, these were merged into one indicator column
-to prevent duplication e.g.~\{bilateral\_cleft\_lips,
-bilateral\_cleft\_lip, cleft\_lip\_b\_l\} -\textgreater{}
-\{bilateral\_cleft\_lip\}.
+In the case where synonymous terms are present, these indicator columns
+are consolidated and merged into one indicator column to prevent
+duplication e.g.~\{bilateral\_cleft\_lips, bilateral\_cleft\_lip,
+cleft\_lip\_b\_l\} -\textgreater{} \{bilateral\_cleft\_lip\}.
 
@@
 
-This format was preserved and other case studies identified in the
-literature
-\citep{ravenscroft2021neurogenetic, monies2019lessons, bijarnia2022growth, ji2022case}
-were manually entered to conform to this template to allow the data to
-be combined for further analysis.
+This format was used as a template to align other written case reports
+identified in the literature
+\citep{ravenscroft2021neurogenetic, monies2019lessons, bijarnia2022growth, ji2022case}.
+These were manually entered to conform to this template to allow the
+data to be combined for further analysis.
 
@@
 represent the presence or absence of a given clinical feature. Only
-those features that appeared in more than one case were included to
-minimise the influence of isolated features in this analysis.
+those features that appeared in more than one case report were included
+to minimise the influence of isolated features in this analysis.
 
@@
 arrive at a set of meaningfully small dimensions that account for most
-of the variation in the data.
+of the variation in the data. This computation was carried out using the
+\emph{FactoMineR} R package \citep{factominer}.
 
@@
 
-This computation of the Multiple Correspondence Analysis was computed
-using the \emph{FactoMineR} R package \citep{factominer}.
-
 The \emph{contribution} of each category was computed for each of the
@@
 The coordinates on the first few principal axes were inspected for any
-outliers or unusual data.
+outliers or unusual data before being graphically interpreted and
+subject to further analysis.
 
@@
 cluster solution is one that is robust and meaningful across all of
-these dimensions Figure~\ref{fig-clvalid}. The optimal method and
+these dimensions (Figure~\ref{fig-clvalid}). The optimal method and
 cluster number was selected based on a trade off between the best
 performing method using the evaluation metrics and a sensible
-partitioning of the data for this analysis Table~\ref{tbl-eig}.
+partitioning of the data for this analysis (Table~\ref{tbl-eig}).
 
@@
 identified two subjects (17, 18) from the data that were clear outliers
-and sat significantly beyond the 95\% confidence ellipse. These
-individuals represented two twins from \citet{magini2019loss} who were
-described with a much milder phenotype and were subject to a missense
-variation in \emph{SMPD4}. Further analysis of this influence is
-conducted below. For now, these subjects were removed to allow for more
-meaningful analysis.
+and sat significantly beyond the 95\% confidence ellipse
+(Figure~\ref{fig-biplotprelim}). These individuals represented two twins
+from \citet{magini2019loss} who were described with a much milder
+phenotype and were subject to a missense variation in \emph{SMPD4}.
+Further analysis of this influence is conducted below. For now, these
+subjects were removed to allow for more meaningful analysis of other
+features.
 
@@
 The \emph{contribution} of the categories were ranked within each of the
-first four MCA dimensions (Figure~\ref{fig-ctr}). The first dimension is
-dominated by highly specific facial dysmorphisms such as epicanthal
-folds, short philtrum and low anterior hairline.
-
-The second MCA dimension is categorised by feeding and respiratory
-dysfunction with vocal cord palsy, tracheostomy, tube feeding and GERD.
-
-Examining the bi-plot of the first two principal axes
-Figure~\ref{fig-biplot} illustrates the contrasting sources of variation
-between conditions relating to airway and swallowing function and
-predominantly facial features and structural brain anomalies such as as
-cerebellar hypoplasia, development delay.
-
-The third dimension is characterised by atypical, complex features such
-as patent ductus arteriosus, respiratory infections, small for
-gestational age and further facial features. These are contrasted by the
-fourth principal dimension which is strongly categorised by respiratory
-distress including stridor and hypercapnia in the absence of typical
+first four principal dimensions in Figure~\ref{fig-ctr}. The first
+dimension is dominated by isolated cases with highly specific facial
+dysmorphisms such as epicanthal folds, short philtrum and low anterior
+hairline. The second MCA dimension is categorised by feeding and
+respiratory dysfunction with vocal cord palsy, tracheostomy, tube
+feeding and GERD. Examining the bi-plot of the first two principal axes
+(Figure~\ref{fig-biplot}) illustrates the contrasting sources of
+variation between conditions relating to airway and swallowing function
+and predominantly facial features and structural brain anomalies such as
+as cerebellar hypoplasia, development delay. The third dimension is
+characterised by atypical, complex features such as patent ductus
+arteriosus, respiratory infections, small for gestational age and
+further facial features. These are contrasted by the fourth principal
+dimension which is strongly categorised by respiratory distress
+including stridor and hypercapnia in the absence of typical
 presentations of simplified gyral pattern, intrauterine growth
@@
 Cluster validation measures resulted in either 2 or 5 clusters as the
-optimal choice.
-
-In order to partition the data into meaningful groupings, k-means with 5
-clusters was selected. This reflects the optimal choice for both AD and
-FOM cluster metrics.
-
-A scatter plot of the cluster results Figure~\ref{fig-clust} projected
-on just the first two dimensions highlights a core cluster of classic
-features of NEDMABA such as microcephaly, hypomyelination,
-arthrogryposis and feet deformity and structural brain anomalies. A
-separate cluster is formed of airway and feeding related conditions
-involving vocal cord paralysis, tracheostomy, tube feeding and enlarged
-ventricles.
-
-Three other smaller satellite clusters exists where highly distinctive
-features were associated with a single individual or family. A full
-articulation of feature to cluster assignment is given in
-Table~\ref{tbl-profile}.
+optimal choice (Table~\ref{tbl-opticlust}). In order to partition the
+data into meaningful groupings, k-means with 5 clusters was selected.
+This reflects the optimal choice for both AD and FOM cluster metrics. A
+scatter plot of the cluster results shown in Figure~\ref{fig-clust} is
+projected on the first two principal dimensions highlights a core
+cluster of classic features of NEDMABA such as microcephaly,
+hypomyelination, arthrogryposis and feet deformity and structural brain
+anomalies. A separate and adjacent cluster is formed of airway and
+feeding related conditions involving vocal cord paralysis, tracheostomy,
+tube feeding and enlarged ventricles. Three other smaller satellite
+clusters exists where highly distinctive features were associated with a
+single individual or family. A full articulation of feature to cluster
+assignment is given in Table~\ref{tbl-profile}.
 
@@
 A comparative box plot of survival times by variant type shows
-significant outliers for missense+splice and missense variations.
-However, even within missense homozygous variants the survival time
-varies greatly. Other missense variants, such as heterozygous missense
-with gene deletion do not show the same increased survival time as noted
-by \citet{magini2019loss}.
+significant outliers for compound heterozygous \emph{missense + splice}
+and homozygous \emph{missense} variants. However, even within homozygous
+missense variants the survival time varies greatly. Other missense
+variants, such as heterozygous missense with gene deletion do not show
+the same increased survival time as noted by \citet{magini2019loss}.
 
@@
 
-A log-rank test was carried out to compare the survival curves of the
-various gene variants and showed a significant difference in the
-stratified survival curves (\(\chi^2 = 14.8, p = 0.04)\) however due to
-the small sample size, several variants had expected counts less than 5
-which would create a violation in assumptions for the test.
+A log-rank test was carried out to compare the survival curves of
+patients stratified by the various gene variants and showed a
+significant difference in the stratified survival curves
+(\(\chi^2 = 14.8, p = 0.04)\) however due to the small sample size and
+large number of possible variants, several had expected counts less than
+5 which would create a violation in assumptions for the test.
 
@@
 
-The use of Multiple Correspondence Analysis confirmed the heterogeneous
-variation in the data with the first four principal axes being required
-to explain more than 50\% of the variation in the data set. The
-examination of key contributing factors, as well as a visual inspection
-of the data projected into this lower dimensional space indicate the
-primary sources of variation in the clinical phenotype are explained by
-a few individuals with very different clinical features. This has
-confounded the typical or classical presentation.
-
-The cluster analysis does not aim to group individuals that are alike,
-but rather the clinical features themselves and how these contrasting
-clusters account for variation in the data. This highlights common
-sources of variation in the clinical phenotype of the condition itself,
-rather than typical coincident features of any one child.
-
-After excluding two outlying, remarkable cases this analysis clearly
-identifies a core cluster of the hallmark features of microcephaly,
-hypomyelination, arthrogryposis, simplified gyral pattern and
-developmental delay. Of significant note is the presence of an adjacent
-cluster of related features that are not as well described in The Human
-Phenotype Ontology \citep{kohler2021human} as shown in
-Table~\ref{tbl-hpo}. We identify a prolific feature of `respiratory
-distress' as an expression more associated with other features such as
-vocal cord palsy, tracheostomy, tube feeding and GERD which are present
-in the case reported data, but not well described in HPO. This
-represents a related but distinct grouping of severe features associated
-with this condition.
+Often, data collected in studies is fit for descriptive purposes but not
+for in-depth statistical analysis. This meta-review of current studies
+was able to consolidate and make publicly available, data sets that
+allow analysis to be conducted on emerging conditions such as NEDMABA.
+
+The use of dimension reduction techniques were shown to be successful in
+uncovering relevant and novel substructure in the data. This was
+particularly evident with the detection of the accepted typical
+phenotype of microcephaly, hypomyelination, arthrogryposis, simplified
+gyral pattern and developmental delay.
+
+After excluding two outlying remarkable cases, this analysis clearly
+identifies the presence of an adjacent cluster of related features that
+are not as well described in The Human Phenotype Ontology
+\citep{kohler2021human} as shown in Table~\ref{tbl-hpo}. We identify the
+prolific feature of `respiratory distress' as an expression more
+commonly associated with \emph{other} features such as vocal cord palsy,
+tracheostomy, tube feeding and Gastroesophageal reflux disease (GERD).
+These features are present in the case reported data, but not well
+described in HPO and help account for the variation seen in the clinical
+phenotype in the first few principal dimensions.
 
@@
 had clinical reports that articulated these dysmorphisms in much more
-detail. On a similar note, the reporting of `talipes' as a condition
-which seems distinct as a source of variation, appearing as an isolated
-feature without the typical microcephaly and hypomyelination. However,
-there was significant heterogeneity in how feet contractures were
-reported in clinical case reports. In the data pre-processing, a
-separate feature was created for reported cases with rocker bottom feet
-or congenital vertical talus versus those with clubfeet/talipes. In
-practice its not clear how these diagnoses were arrived at, and given
-the non-specific nature of these features, care needs to be taken in
-distinguishing between similar manifesting features in the literature
-where possible. This is particularly clear as rocker bottom feet was
+detail, which skewed the interpretation.
+
+On a similar note, talipes appears as an isolated feature without the
+typical microcephaly and hypomyelination. However, there was significant
+heterogeneity in how feet contractures were reported in clinical case
+reports. While distinctions were made during data pre-processing between
+rocker-bottom feed and talipes, this likely to make intpretation
+somewhat difficult. This is particularly clear as rocker-bottom feet was
 found closely related in our analysis to other core features despite the
 prevalence in HPO being quite low (2/21), suggesting possible
-under-reporting or confounded reporting of this feature. Another feature
-that clusters within the core identified features is bilateral cleft
-lip. This condition has a reported frequency of just 2/21
+under-reporting or confounded reporting of this feature.
+
+Another feature clusted within core identified features is bilateral
+cleft lip. This condition has a reported frequency of just 2/21
 (Table~\ref{tbl-hpo}) and may be an emerging feature for a more typical
 presentation given its close relationship with other classic features.
-Not all features in the data were included in this analysis, particular
-if isolated to just one case. In this circumstance, these terms in
-Table~\ref{tbl-hpo} have been labelled as `Isolated'.
+
+The heterogeneous variation in the data was clearly evident however,
+with the first four principal axes being required to explain more than
+50\% of the variation in the data set. Visual inspection of the data
+projected into this lower dimensional space indicate the primary sources
+of variation in the clinical phenotype are explained by a few
+individuals with very different clinical features. This has confounded
+the typical or classical presentation. Not all features in the data were
+included in this analysis, particular if isolated to just one case. In
+this circumstance, these terms in Table~\ref{tbl-hpo} have been labelled
+as `Isolated'.
 
@@
 the data and is valid in some applications. In this study, a larger
-number was selected to align the goals of the analysis which was to
+number (5) was selected to align the goals of the analysis which was to
 uncover distinct groups of features.
@@
 childhood and in some cases into the second decade of life. This work
-identifies a one year survival rate of 0.403 (95\% CI: 0.252, 0.644). A
-naive estimate, using just the number of deceased patients at this time
-as proportion of the study population is 0.57. This highlights the
-importance of applying a survival analysis technique that incorporates
-censoring so as not to erroneously overestimate the survival
-probability. The current literature suggests the type of gene variant
-has an impact on survival time, which seems to be supported in this
-work. Through a comparison of multiple survival curves, stratified by
-variation type, the limited sample size prevented other meaningful
-analysis such as log-rank tests or cox proportional hazard modelling
-which are likely under-powered when comparing multiple groups.
+identifies a one year survival rate of 0.403 (95\% CI: 0.252, 0.644). On
+the other hand, a naive estimate using just the number of deceased
+patients at this time as proportion of the study population is 0.57
+(16/28). This highlights the importance of applying a survival analysis
+technique that incorporates censoring so as not to erroneously
+overestimate the survival probability. The current literature suggests
+the type of gene variant has an impact on survival time, which seems to
+be supported in this work. Through a comparison of multiple survival
+curves, stratified by variation type, the limited sample size prevented
+definitive conclusions with many cells having expected counts less than
+5.
 
@@
 groupings of features that are exhibited in patients such as vocal cord
-paralysis with tracheostomy, swallowing dysfunction, tube feeding and
-GERD. Furthermore, we were able to quantify the expected baseline
-survival probabilities for children with this condition as well as
-quantifying the significant variation in these estimates. Future work is
-required with the benefit of more cases, to better understand the
-differences in survival curves for other factors associated with
-genotype.
+paralysis, swallowing dysfunction, tube feeding and GERD. Furthermore,
+we were able to quantify the expected baseline survival probabilities
+for children with this condition as well as quantifying the significant
+variation in these estimates. Future work is required with the benefit
+of more cases, to better understand the differences in survival curves
+for other factors associated with genotype.
 

### 2022-09-13 paper/paper.qmd
edits

@@
 ---
-title: Clinical Phenotype and Survival Analysis in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in *SMPD4*
+title: Clinical Phenotype and Survival Analysis of Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in *SMPD4*
 author:
@@
 This is driven in part to the paucity of analysis-ready data on reported cases.
-This research aims to collate and standardise available case reports into a common dataset, to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA.
+This analysis aims to collate and standardise available case reports into a common dataset, to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA.
 To overcome the challenge of multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation.
-To account for censoring in the data Kaplan-Meier estimation is formulated calculate patient survival time.
-The analysis correctly detected the classic phenotype for this condition, as well as an additional distinct feature-cluster relating to findings of vocal cord paralysis, feeding dysfunction and respiratory failure.
-Several additional outlying and isolated clinical features were also identified.
+To account for censoring in the data, Kaplan-Meier estimation is formulated to calculate patient survival time.
+The analysis correctly detected the classic phenotype for this condition, as well as a new distinct feature-cluster relating to findings of vocal cord paralysis, feeding dysfunction and respiratory failure.
 The survival probability for those affected was found to decline sharply in early infancy with median survival of 150 days, but with some surviving as long as 12.5 years.
 This wide range of outcomes is provisionally associated with different variant types however this conclusion could not be validated based on very low sample sizes.
+An `R` package called `SMPD4` was developed to publish standardised analysis-ready datasets used in this study.
 This analysis represents the first of its kind to help describe associations and trajectories of individuals with this newly reported condition, despite challenges with sparse and inconsistent data.
@@
 
-@ravenscroft2021neurogenetic performed a study over 190 probands and identified a novel missense variation in SMPD4 which at the time of whole exome sequencing was not well described in the literature.
-The impacted family from Melbourne exhibited features involving arthrogryposis multiplex congenita, complex brain malformations, small for gestation age and hypoplasia of the corpus callosum.
-In two of the three related cases were additional features of microcephaly, congenital encephalopathy, cerebellar malformation and hypoplasia and hypomyelination.
-
 A detailed study by @magini2019loss involved 12 unrelated families with 32 individuals (21 with detailed clinical information).
@@
 Other patients with truncating variants are shown to have more severe presentations, while a range of additional significant clinical features were reported involving dysmorphic facial features, seizure, vocal cord paralysis and hearing impairment.
-
+@ravenscroft2021neurogenetic studied a family from Melbourne exhibiting features involving arthrogryposis multiplex congenita, complex brain malformations, small for gestation age and hypoplasia of the corpus callosum.
+In two of the three related cases were additional features of microcephaly, congenital encephalopathy, cerebellar malformation and hypoplasia and hypomyelination.
 A case study from China is described in @ji2022case involving a girl presenting in infancy with intrauterine growth restriction, microcephaly, postnatal developmental delay, arthrogryposis, hypertonicity, seizure, and hypomyelination on brain magnetic resonance imaging.
@@
 An individual reported in @monies2019lessons presented with distinct symptoms of brain atrophy and skeletal dysplasia whereas the case in @magini2019loss with the same variant exhibited more typical clinical features.
-
 A recent study by @bijarnia2022growth presents the case of a 22-month old girl presenting with the typical phenotype of neurodevelopmental delay, prenatal onset growth failure, arthrogryposis, microcephaly and brain anomalies including severe hypomyelination, simplified gyral pattern and hypoplasia of corpus callosum and brainstem.
@@
 Further work to collate and analyse data relating to NEDMABA is challenging.
-While @magini2019loss cataloged a detailed clinical phenotype data set as a supplementary data set to their study, these data are not suitable for statistical analysis as presented.
-Many of the clinical features are represented as free text descriptions and a variety of non-standard terminologies are applied between cases, making machine interpretation of the data difficult.
+While @magini2019loss included a supplementary clinical phenotype data set in their study, these data are not suitable for statistical analysis.
+Many of the clinical features are represented as text descriptions and a variety of non-standard terminologies are applied between cases, making machine interpretation of the data difficult.
 Other studies [@ji2022case; @bijarnia2022growth] present only written case reports with some tabulated summaries.
-While larger studies [@ravenscroft2021neurogenetic; @monies2019lessons] focus more on genetic data, with less detail included on the clinical presentation of individuals in the cohort.
-
+While larger studies [@ravenscroft2021neurogenetic; @monies2019lessons] focus more on genetic data, with less detail included on the clinical presentation of individuals.
 To overcome these data challenges, @JSSv059i10 proposes a data structure know as 'tidy' data which organises each observation in a row, with each feature as its own column and each value in just one cell.
@@
 The authors argue patients with rare disorders (often in small samples) frequently present with varied symptoms that do not match exactly with the described phenotype.
-When considering this situation of low sample size data with many features, methods in the field of multivariate statistical analysis are commonly deployed.
+When considering the situation of low sample size data with many features, methods in the field of multivariate statistical analysis are commonly deployed.
 Here the aim is to find substructure in the data, or a simple representation of the multi-dimensional space.
@@
 An open research question exists around the clinical pathway and survival time of children exhibiting bi-allelic loss-of-function variants in *SMPD4*.
-With current research highlighting a diverse and heterogeneous phenotype, the relationship and correlations between these diverse features is not well understood.
-Furthermore, the survival time of affected children has not been extensively analysed despite early-demise featuring as a severe outcome in many cases.
+With current research highlighting a diverse and heterogeneous phenotype, the relationship between these diverse features is not well understood.
+Furthermore, the survival time of affected children has not been extensively analysed despite early-demise featuring as a severe outcome.
 While a connection has been highlighted between some missense variants and a milder presentation with longer survival [@magini2019loss], this hypothesis has not been analysed further.
 
-This research has three key aims.
+This analysis has three key aims.
 First, to collate and transform the variety of early case reports and studies on clinical phenotype data for this novel variant into an analysis-ready *tidy* data set.
@@
 The data were transposed to ensure it could be presented as *tidy* formatted data [@JSSv059i10].
-This requires each variable (clinical phenotype) forms a column; each observation (individual) forms a row and each type of observational unit forms a table (every cell has just one item).
+This requires that each variable (clinical phenotype) forms a column; each observation (individual) forms a row and each type of observational unit forms a table (every cell has just one item).
 This resulted in a long-formatted data set of 21 rows (individuals) and 64 columns (clinical features).
@@
 A full list of this data set is available in @tbl-dict.
-These data sets were compiled into a publicly available `R` Package called `SMPD4` [@smpd4-data] in order to allow for reproducibility and sharing.
+These data sets were compiled into a publicly available `R` Package called `SMPD4` [@smpd4-data] to allow for reproducibility and sharing.
 
@@
 
-The data on subjects introduced above is subset to include only variables that can be transformed into binary indicator variables that represent the presence or absence of a given clinical feature.
-Only those features that appeared in more than one case report were included to minimise the influence of isolated features in this analysis.
+The data on subjects above is subset to include only variables that can be transformed into binary indicator variables representing the presence or absence of a given clinical feature.
+Only those features that appeared in more than one case report were included to minimise the influence of isolated features.
 
@@
   geom_col() +
-  scale_y_reordered()+
+  scale_y_reordered() +
   scale_fill_grey(start = 0.8, end = 0.2) +
@@
 This reflects the optimal choice for both AD and FOM cluster metrics.
-A scatter plot of the cluster results shown in @fig-clust is projected on the first two principal dimensions highlights a core cluster of classic features of NEDMABA such as microcephaly, hypomyelination, arthrogryposis and feet deformity and structural brain anomalies.
+The results are summarised in @tbl-clusttab and a scatter plot of the cluster results is shown in @fig-clust is projected on the first two principal dimensions.
+This highlights a core cluster of classic features of NEDMABA such as microcephaly, hypomyelination, arthrogryposis and feet deformity and structural brain anomalies.
 A separate and adjacent cluster is formed of airway and feeding related conditions involving vocal cord paralysis, tracheostomy, tube feeding and enlarged ventricles.
@@
 #| label: tbl-opticlust
-#| tbl-cap: "Optimal clusters"
+#| tbl-cap: "Optimal cluster numbers and methods based on a range of stability validation metrics."
 optimalScores(cluster_select) |> 
@@
   facet_wrap(~measure, scales= 'free')
+```
+
+```{r}
+#| label: tbl-clusttab
+#| tbl-cap: "A summary table to cluster membership showing the cluster number and number of features in each cluster and a sample of features that have membership in each."
+cluster_assignments |> 
+  as_tibble(rownames = 'feature') |> 
+  mutate(feature = str_remove(feature, "_ind_1|_1"),
+         feature = str_replace_all(feature, "_", " ")) |> 
+  mutate('Cluster No.' = value) |> 
+    mutate("Cluster Description" = fct_recode(as.character(value),
+    Core = "4",
+    Adjacent = "2",
+    Outlying = "3",
+    Outlying = "5",
+    Outlying = "1",
+    )) |> 
+  group_by(`Cluster No.`, `Cluster Description`) |> 
+  summarise('No. Features' = n(),
+            `Sample Features in Cluster` = paste(head(feature, 4), collapse = ", "), .groups = 'drop') |> 
+  select(-`Cluster Description`) |> 
+  gt() |> 
+  gt::tab_footnote(footnote = "A sample of 4 features shown to illustrate cluster membership", locations = cells_column_labels("Sample Features in Cluster")) 
 ```
@@
 At time 0 the survival probability was 0.82 (95% CI: 0.69, 0.98).
-Survival probability after one year at 0.403 (95% CI: 0.252, 0.644) and survival probability at close to ten years of 0.24 (95% CI: 0.11, 0.52).
+Survival probability after one year of 0.403 (95% CI: 0.252, 0.644) and survival probability at close to ten years of 0.24 (95% CI: 0.11, 0.52).
 
@@
 ```{r,fig.width=7, warning=FALSE}
-# KM by variant
+#| label: fig-box
+#| fig-cap: "A compariative box plot showing the range of survival times for subjects split by the type of variant.
+Green bars indicate homozygous variants, while red indicates compound heterozygous variants."
 
@@
 
-Often, data collected in studies is fit for descriptive purposes but not for in-depth statistical analysis.
-This meta-review of current studies was able to consolidate and make publicly available, data sets that allow analysis to be conducted on emerging conditions such as NEDMABA.
-
-The use of dimension reduction techniques were shown to be successful in uncovering relevant and novel substructure in the data.
+This meta-review of current research was able to consolidate data that allowed statistical analysis to be conducted on affected individuals.
+The use of dimension reduction techniques was shown to be successful in uncovering relevant and novel substructure in the data.
 This was particularly evident with the detection of the accepted typical phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay.
-
-After excluding two outlying remarkable cases, this analysis clearly identifies the presence of an adjacent cluster of related features that are not as well described in The Human Phenotype Ontology [@kohler2021human] as shown in @tbl-hpo.
-We identify the prolific feature of 'respiratory distress' as an expression more commonly associated with *other* features such as vocal cord palsy, tracheostomy, tube feeding and Gastroesophageal reflux disease (GERD).
+After excluding two outlying remarkable cases, we identified the presence of an adjacent cluster of related features that are not well described in The Human Phenotype Ontology [@kohler2021human] as shown in @tbl-hpo.
+It shows the prolific feature of 'respiratory distress' as more commonly associated with these adjacent features such as vocal cord palsy, tracheostomy, tube feeding and Gastroesophageal reflux disease (GERD).
 These features are present in the case reported data, but not well described in HPO and help account for the variation seen in the clinical phenotype in the first few principal dimensions.
 
-Other smaller clusters identify outlying features that are expressed in individuals or specific families.
-Many of the outlying features are highly specific facial or anatomical dysmorphisms.
-While coarse facial features are a noted and core presentation, those with a milder course had clinical reports that articulated these dysmorphisms in much more detail, which skewed the interpretation.
+The heterogeneous variation in the data was clearly evident with the first four principal axes being required to explain more than 50% of the variation in the data set.
+Visual inspection shows the primary sources of variation in the clinical phenotype are explained by a few individuals with very different clinical features.
+Many of these exhibit highly specific facial or anatomical dysmorphisms.
+While coarse facial features are a core presentation, those with a milder course had clinical reports that articulated these dysmorphisms in much more detail.
+Additionally, talipes appears as an isolated feature without the typical microcephaly and hypomyelination, however there was significant heterogeneity in how feet contractures were reported in clinical case reports.
+This is particularly evident in rocker-bottom feet which was found closely related in our analysis to core features despite the prevalence in HPO being quite low (2/21), suggesting possible under-reporting or confounded reporting of this feature.
+Another core identified feature is bilateral cleft lip.
+This condition has a reported frequency of just 2/21 (@tbl-hpo) and may be an emerging feature for a more typical presentation given its close relationship with other classic features.
+Features isolated to just one or few cases in our consolidated data have been labelled as 'Isolated' in @tbl-hpo.
+An optimal choice of both 2 and 5 clusters were recommended in the analysis.
+A partition of the data with just two clusters would likely be useful in separating outliers from other data.
+In this analysis a larger number (5) was selected to align the goals of the analysis which was to uncover distinct groups of features.
 
-On a similar note, talipes appears as an isolated feature without the typical microcephaly and hypomyelination.
-However, there was significant heterogeneity in how feet contractures were reported in clinical case reports.
-While distinctions were made during data pre-processing between rocker-bottom feed and talipes, this likely to make intpretation somewhat difficult.
-This is particularly clear as rocker-bottom feet was found closely related in our analysis to other core features despite the prevalence in HPO being quite low (2/21), suggesting possible under-reporting or confounded reporting of this feature.
-
-Another feature clusted within core identified features is bilateral cleft lip.
-This condition has a reported frequency of just 2/21 (@tbl-hpo) and may be an emerging feature for a more typical presentation given its close relationship with other classic features.
-
-The heterogeneous variation in the data was clearly evident however, with the first four principal axes being required to explain more than 50% of the variation in the data set.
-Visual inspection of the data projected into this lower dimensional space indicate the primary sources of variation in the clinical phenotype are explained by a few individuals with very different clinical features.
-This has confounded the typical or classical presentation.
-Not all features in the data were included in this analysis, particular if isolated to just one case.
-In this circumstance, these terms in @tbl-hpo have been labelled as 'Isolated'.
-
-While the number of clusters chosen all represent valid partition of the data, some metrics indicated an optimal selection of 2 clusters.
-This type of partition is useful when segregating outliers from the rest of the data and is valid in some applications.
-In this study, a larger number (5) was selected to align the goals of the analysis which was to uncover distinct groups of features.
-
-In terms of patient survival, the first few months show a sharp decline in survival probability confirming early demise in the infant period as a key factor.
-A small number of cases show much longer survival into childhood and in some cases into the second decade of life.
+In terms of patient survival, the first few months of life show a sharp decline in survival probability confirming early demise in the infant period as a key factor.
+A small number of cases show survival into childhood and in some cases into the second decade of life.
 This work identifies a one year survival rate of 0.403 (95% CI: 0.252, 0.644).
-On the other hand, a naive estimate using just the number of deceased patients at this time as proportion of the study population is 0.57 (16/28).
+On the other hand, a naive estimate using just the number of deceased patients at this time as proportion of the study population is 0.43 (1 - 16/28).
 This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to erroneously overestimate the survival probability.

### 2022-09-13 paper/paper.tex
edits

@@
 \hypersetup{
-  pdftitle={Clinical Phenotype and Survival Analysis in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD4},
+  pdftitle={Clinical Phenotype and Survival Analysis of Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD4},
   pdfauthor={Dean Marchiori},
@@
 \begin{frontmatter}
-\title{Clinical Phenotype and Survival Analysis in Neurodevelopmental
+\title{Clinical Phenotype and Survival Analysis of Neurodevelopmental
 Disorder with Microcephaly, Arthrogryposis, and Structural Brain
@@
 probabilities of affected individuals. This is driven in part to the
-paucity of analysis-ready data on reported cases. This research aims to
+paucity of analysis-ready data on reported cases. This analysis aims to
 collate and standardise available case reports into a common dataset, to
@@
 technique, which is then subject to cluster analysis and interpretation.
-To account for censoring in the data Kaplan-Meier estimation is
-formulated calculate patient survival time. The analysis correctly
-detected the classic phenotype for this condition, as well as an
-additional distinct feature-cluster relating to findings of vocal cord
-paralysis, feeding dysfunction and respiratory failure. Several
-additional outlying and isolated clinical features were also identified.
-The survival probability for those affected was found to decline sharply
-in early infancy with median survival of 150 days, but with some
-surviving as long as 12.5 years. This wide range of outcomes is
-provisionally associated with different variant types however this
-conclusion could not be validated based on very low sample sizes. This
-analysis represents the first of its kind to help describe associations
-and trajectories of individuals with this newly reported condition,
-despite challenges with sparse and inconsistent data. This analysis can
-provide clinicians and genetic counsellors with better information to
-aide in decision making and support for families with this rare
-condition.
+To account for censoring in the data, Kaplan-Meier estimation is
+formulated to calculate patient survival time. The analysis correctly
+detected the classic phenotype for this condition, as well as a new
+distinct feature-cluster relating to findings of vocal cord paralysis,
+feeding dysfunction and respiratory failure. The survival probability
+for those affected was found to decline sharply in early infancy with
+median survival of 150 days, but with some surviving as long as 12.5
+years. This wide range of outcomes is provisionally associated with
+different variant types however this conclusion could not be validated
+based on very low sample sizes. An \texttt{R} package called
+\texttt{SMPD4} was developed to publish standardised analysis-ready
+datasets used in this study. This analysis represents the first of its
+kind to help describe associations and trajectories of individuals with
+this newly reported condition, despite challenges with sparse and
+inconsistent data. This analysis can provide clinicians and genetic
+counsellors with better information to aide in decision making and
+support for families with this rare condition.
 \end{abstract}
@@
 \end{keyword}
-\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, enhanced, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, enhanced, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-\citet{ravenscroft2021neurogenetic} performed a study over 190 probands
-and identified a novel missense variation in SMPD4 which at the time of
-whole exome sequencing was not well described in the literature. The
-impacted family from Melbourne exhibited features involving
-arthrogryposis multiplex congenita, complex brain malformations, small
-for gestation age and hypoplasia of the corpus callosum. In two of the
-three related cases were additional features of microcephaly, congenital
-encephalopathy, cerebellar malformation and hypoplasia and
-hypomyelination.
-
 A detailed study by \citet{magini2019loss} involved 12 unrelated
@@
 dysmorphic facial features, seizure, vocal cord paralysis and hearing
-impairment.
-
-A case study from China is described in \citet{ji2022case} involving a
-girl presenting in infancy with intrauterine growth restriction,
+impairment. \citet{ravenscroft2021neurogenetic} studied a family from
+Melbourne exhibiting features involving arthrogryposis multiplex
+congenita, complex brain malformations, small for gestation age and
+hypoplasia of the corpus callosum. In two of the three related cases
+were additional features of microcephaly, congenital encephalopathy,
+cerebellar malformation and hypoplasia and hypomyelination. A case study
+from China is described in \citet{ji2022case} involving a girl
+presenting in infancy with intrauterine growth restriction,
 microcephaly, postnatal developmental delay, arthrogryposis,
@@
 \citet{magini2019loss} with the same variant exhibited more typical
-clinical features.
-
-A recent study by \citet{bijarnia2022growth} presents the case of a
-22-month old girl presenting with the typical phenotype of
+clinical features. A recent study by \citet{bijarnia2022growth} presents
+the case of a 22-month old girl presenting with the typical phenotype of
 neurodevelopmental delay, prenatal onset growth failure, arthrogryposis,
@@
 Further work to collate and analyse data relating to NEDMABA is
-challenging. While \citet{magini2019loss} cataloged a detailed clinical
-phenotype data set as a supplementary data set to their study, these
-data are not suitable for statistical analysis as presented. Many of the
-clinical features are represented as free text descriptions and a
-variety of non-standard terminologies are applied between cases, making
-machine interpretation of the data difficult. Other studies
-\citep{ji2022case, bijarnia2022growth} present only written case reports
-with some tabulated summaries. While larger studies
-\citep{ravenscroft2021neurogenetic, monies2019lessons} focus more on
-genetic data, with less detail included on the clinical presentation of
-individuals in the cohort.
-
-To overcome these data challenges, \citet{JSSv059i10} proposes a data
-structure know as `tidy' data which organises each observation in a row,
-with each feature as its own column and each value in just one cell. In
-this case detailed text-based data can be transformed to high
-dimensional binary indicators of key clinical features. This structure
-is conducive to effective statistical analysis and integrates
-deliberately with the \emph{tidyverse} \citep{tidyverse} collection of
-packages for data analysis in R \citep{rbase}.
+challenging. While \citet{magini2019loss} included a supplementary
+clinical phenotype data set in their study, these data are not suitable
+for statistical analysis. Many of the clinical features are represented
+as text descriptions and a variety of non-standard terminologies are
+applied between cases, making machine interpretation of the data
+difficult. Other studies \citep{ji2022case, bijarnia2022growth} present
+only written case reports with some tabulated summaries. While larger
+studies \citep{ravenscroft2021neurogenetic, monies2019lessons} focus
+more on genetic data, with less detail included on the clinical
+presentation of individuals. To overcome these data challenges,
+\citet{JSSv059i10} proposes a data structure know as `tidy' data which
+organises each observation in a row, with each feature as its own column
+and each value in just one cell. In this case detailed text-based data
+can be transformed to high dimensional binary indicators of key clinical
+features. This structure is conducive to effective statistical analysis
+and integrates deliberately with the \emph{tidyverse} \citep{tidyverse}
+collection of packages for data analysis in R \citep{rbase}.
 
@@
 symptoms that do not match exactly with the described phenotype. When
-considering this situation of low sample size data with many features,
+considering the situation of low sample size data with many features,
 methods in the field of multivariate statistical analysis are commonly
@@
 variants in \emph{SMPD4}. With current research highlighting a diverse
-and heterogeneous phenotype, the relationship and correlations between
-these diverse features is not well understood. Furthermore, the survival
-time of affected children has not been extensively analysed despite
-early-demise featuring as a severe outcome in many cases. While a
-connection has been highlighted between some missense variants and a
-milder presentation with longer survival \citep{magini2019loss}, this
-hypothesis has not been analysed further.
-
-This research has three key aims. First, to collate and transform the
+and heterogeneous phenotype, the relationship between these diverse
+features is not well understood. Furthermore, the survival time of
+affected children has not been extensively analysed despite early-demise
+featuring as a severe outcome. While a connection has been highlighted
+between some missense variants and a milder presentation with longer
+survival \citep{magini2019loss}, this hypothesis has not been analysed
+further.
+
+This analysis has three key aims. First, to collate and transform the
 variety of early case reports and studies on clinical phenotype data for
@@
 data were transposed to ensure it could be presented as \emph{tidy}
-formatted data \citep{JSSv059i10}. This requires each variable (clinical
-phenotype) forms a column; each observation (individual) forms a row and
-each type of observational unit forms a table (every cell has just one
-item). This resulted in a long-formatted data set of 21 rows
+formatted data \citep{JSSv059i10}. This requires that each variable
+(clinical phenotype) forms a column; each observation (individual) forms
+a row and each type of observational unit forms a table (every cell has
+just one item). This resulted in a long-formatted data set of 21 rows
 (individuals) and 64 columns (clinical features). In many cases, key
@@
 publicly available \texttt{R} Package called \texttt{SMPD4}
-\citep{smpd4-data} in order to allow for reproducibility and sharing.
+\citep{smpd4-data} to allow for reproducibility and sharing.
 
@@
 
-The data on subjects introduced above is subset to include only
-variables that can be transformed into binary indicator variables that
-represent the presence or absence of a given clinical feature. Only
-those features that appeared in more than one case report were included
-to minimise the influence of isolated features in this analysis.
+The data on subjects above is subset to include only variables that can
+be transformed into binary indicator variables representing the presence
+or absence of a given clinical feature. Only those features that
+appeared in more than one case report were included to minimise the
+influence of isolated features.
 
@@
 data into meaningful groupings, k-means with 5 clusters was selected.
-This reflects the optimal choice for both AD and FOM cluster metrics. A
-scatter plot of the cluster results shown in Figure~\ref{fig-clust} is
-projected on the first two principal dimensions highlights a core
+This reflects the optimal choice for both AD and FOM cluster metrics.
+The results are summarised in Table~\ref{tbl-clusttab} and a scatter
+plot of the cluster results is shown in Figure~\ref{fig-clust} is
+projected on the first two principal dimensions. This highlights a core
 cluster of classic features of NEDMABA such as microcephaly,
@@
 \begin{longtable}{lrlr}
-\caption{\label{tbl-opticlust}Optimal clusters }\tabularnewline
+\caption{\label{tbl-opticlust}Optimal cluster numbers and methods based on a range of stability
+validation metrics. }\tabularnewline
 
@@
 
+\hypertarget{tbl-clusttab}{}
+\setlength{\LTpost}{0mm}
+\begin{longtable}{rrl}
+\caption{\label{tbl-clusttab}A summary table to cluster membership showing the cluster number and
+number of features in each cluster and a sample of features that have
+membership in each. }\tabularnewline
+
+\toprule
+Cluster No. & No. Features & Sample Features in Cluster\textsuperscript{1} \\ 
+\midrule
+1 & 7 & polyhydramnios, short palpebral fissures, patent ductus arteriosus, thin lips \\ 
+2 & 11 & vocal cord palsy, respiratory distress, tracheostomy, posteriorly angulated ears \\ 
+3 & 7 & epicanthal folds, receding forehead, spastic tetraparesis, intubation \\ 
+4 & 21 & microcephaly, sgp, iugr, talipes \\ 
+5 & 6 & hypercapnia, inspiratory stridor, oligohydramnios, prominent ears \\ 
+\bottomrule
+\end{longtable}
+\begin{minipage}{\linewidth}
+\textsuperscript{1}A sample of 4 features shown to illustrate cluster membership\\
+\end{minipage}
+
 \begin{figure}
@@
 survival probability was 0.82 (95\% CI: 0.69, 0.98). Survival
-probability after one year at 0.403 (95\% CI: 0.252, 0.644) and survival
+probability after one year of 0.403 (95\% CI: 0.252, 0.644) and survival
 probability at close to ten years of 0.24 (95\% CI: 0.11, 0.52).
@@
 
-\includegraphics{paper_files/figure-pdf/unnamed-chunk-14-1.pdf}
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-pdf/fig-box-1.pdf}
+
+}
+
+\caption{\label{fig-box}A compariative box plot showing the range of
+survival times for subjects split by the type of variant. Green bars
+indicate homozygous variants, while red indicates compound heterozygous
+variants.}
+
+\end{figure}
 
@@
 
-Often, data collected in studies is fit for descriptive purposes but not
-for in-depth statistical analysis. This meta-review of current studies
-was able to consolidate and make publicly available, data sets that
-allow analysis to be conducted on emerging conditions such as NEDMABA.
-
-The use of dimension reduction techniques were shown to be successful in
+This meta-review of current research was able to consolidate data that
+allowed statistical analysis to be conducted on affected individuals.
+The use of dimension reduction techniques was shown to be successful in
 uncovering relevant and novel substructure in the data. This was
@@
 phenotype of microcephaly, hypomyelination, arthrogryposis, simplified
-gyral pattern and developmental delay.
-
-After excluding two outlying remarkable cases, this analysis clearly
-identifies the presence of an adjacent cluster of related features that
-are not as well described in The Human Phenotype Ontology
-\citep{kohler2021human} as shown in Table~\ref{tbl-hpo}. We identify the
-prolific feature of `respiratory distress' as an expression more
-commonly associated with \emph{other} features such as vocal cord palsy,
+gyral pattern and developmental delay. After excluding two outlying
+remarkable cases, we identified the presence of an adjacent cluster of
+related features that are not well described in The Human Phenotype
+Ontology \citep{kohler2021human} as shown in Table~\ref{tbl-hpo}. It
+shows the prolific feature of `respiratory distress' as more commonly
+associated with these adjacent features such as vocal cord palsy,
 tracheostomy, tube feeding and Gastroesophageal reflux disease (GERD).
@@
 
-Other smaller clusters identify outlying features that are expressed in
-individuals or specific families. Many of the outlying features are
+The heterogeneous variation in the data was clearly evident with the
+first four principal axes being required to explain more than 50\% of
+the variation in the data set. Visual inspection shows the primary
+sources of variation in the clinical phenotype are explained by a few
+individuals with very different clinical features. Many of these exhibit
 highly specific facial or anatomical dysmorphisms. While coarse facial
-features are a noted and core presentation, those with a milder course
-had clinical reports that articulated these dysmorphisms in much more
-detail, which skewed the interpretation.
-
-On a similar note, talipes appears as an isolated feature without the
-typical microcephaly and hypomyelination. However, there was significant
+features are a core presentation, those with a milder course had
+clinical reports that articulated these dysmorphisms in much more
+detail. Additionally, talipes appears as an isolated feature without the
+typical microcephaly and hypomyelination, however there was significant
 heterogeneity in how feet contractures were reported in clinical case
-reports. While distinctions were made during data pre-processing between
-rocker-bottom feed and talipes, this likely to make intpretation
-somewhat difficult. This is particularly clear as rocker-bottom feet was
-found closely related in our analysis to other core features despite the
+reports. This is particularly evident in rocker-bottom feet which was
+found closely related in our analysis to core features despite the
 prevalence in HPO being quite low (2/21), suggesting possible
-under-reporting or confounded reporting of this feature.
-
-Another feature clusted within core identified features is bilateral
-cleft lip. This condition has a reported frequency of just 2/21
-(Table~\ref{tbl-hpo}) and may be an emerging feature for a more typical
-presentation given its close relationship with other classic features.
-
-The heterogeneous variation in the data was clearly evident however,
-with the first four principal axes being required to explain more than
-50\% of the variation in the data set. Visual inspection of the data
-projected into this lower dimensional space indicate the primary sources
-of variation in the clinical phenotype are explained by a few
-individuals with very different clinical features. This has confounded
-the typical or classical presentation. Not all features in the data were
-included in this analysis, particular if isolated to just one case. In
-this circumstance, these terms in Table~\ref{tbl-hpo} have been labelled
-as `Isolated'.
-
-While the number of clusters chosen all represent valid partition of the
-data, some metrics indicated an optimal selection of 2 clusters. This
-type of partition is useful when segregating outliers from the rest of
-the data and is valid in some applications. In this study, a larger
-number (5) was selected to align the goals of the analysis which was to
-uncover distinct groups of features.
-
-In terms of patient survival, the first few months show a sharp decline
-in survival probability confirming early demise in the infant period as
-a key factor. A small number of cases show much longer survival into
+under-reporting or confounded reporting of this feature. Another core
+identified feature is bilateral cleft lip. This condition has a reported
+frequency of just 2/21 (Table~\ref{tbl-hpo}) and may be an emerging
+feature for a more typical presentation given its close relationship
+with other classic features. Features isolated to just one or few cases
+in our consolidated data have been labelled as `Isolated' in
+Table~\ref{tbl-hpo}. An optimal choice of both 2 and 5 clusters were
+recommended in the analysis. A partition of the data with just two
+clusters would likely be useful in separating outliers from other data.
+In this analysis a larger number (5) was selected to align the goals of
+the analysis which was to uncover distinct groups of features.
+
+In terms of patient survival, the first few months of life show a sharp
+decline in survival probability confirming early demise in the infant
+period as a key factor. A small number of cases show survival into
 childhood and in some cases into the second decade of life. This work
@@
 the other hand, a naive estimate using just the number of deceased
-patients at this time as proportion of the study population is 0.57
-(16/28). This highlights the importance of applying a survival analysis
+patients at this time as proportion of the study population is 0.43 (1 -
+16/28). This highlights the importance of applying a survival analysis
 technique that incorporates censoring so as not to erroneously

### 2022-09-15 paper/paper.tex
simplified

@@
 \documentclass[
-  authoryear,
-  preprint,
-  3p]{elsarticle}
+  letterpaper,
+  DIV=11,
+  numbers=noendperiod]{scrartcl}
 
@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 % Make \paragraph and \subparagraph free-standing
@@
 \makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \usepackage{longtable}
+\KOMAoption{captions}{tableheading}
 \makeatletter
@@
 \fi
-\usepackage[]{natbib}
-\bibliographystyle{elsarticle-harv}
 \IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
@@
 \hypersetup{
-  pdftitle={Clinical Phenotype and Survival Analysis of Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD4},
+  pdftitle={Meta-Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD},
   pdfauthor={Dean Marchiori},
-  pdfkeywords={NEDMABA, SMPD4},
   colorlinks=true,
@@
 
-\setlength{\parindent}{6pt}
+\title{Meta-Analysis of Clinical Phenotype and Patient Survival in
+Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and
+Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants
+in SMPD}
+\author{Dean Marchiori}
+\date{9/15/22}
+
 \begin{document}
-
-\begin{frontmatter}
-\title{Clinical Phenotype and Survival Analysis of Neurodevelopmental
-Disorder with Microcephaly, Arthrogryposis, and Structural Brain
-Anomalies Due to Bi-allelic Loss of Function Variants in \emph{SMPD4}}
-\author[]{Dean Marchiori%
-\corref{cor1}%
-}
- \ead{deanmarchiori@gmail.com}
-
-\cortext[cor1]{Corresponding author}
-
-\begin{abstract}
+\maketitle
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, boxrule=0pt, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+
+\hypertarget{abstract}{%
+\section{Abstract}\label{abstract}}
+
 A recently described, rare genetic condition known as Neurodevelopmental
@@
 support for families with this rare condition.
-\end{abstract}
-
-\begin{keyword}
-    NEDMABA \sep 
-    SMPD4
-\end{keyword}
-\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, enhanced, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 children with bi-allelic loss-of-function variants in \emph{SMPD4}
-\citep{magini2019loss}. Sphingomyelinases such as \emph{SMPD4} play an
+(Magini et al. 2019). Sphingomyelinases such as \emph{SMPD4} play an
 important cellular role by hydrolyzing sphingomyelin into ceramide and
@@
 
-A detailed study by \citet{magini2019loss} involved 12 unrelated
-families with 32 individuals (21 with detailed clinical information).
+A detailed study by Magini et al. (2019) involved 12 unrelated families
+with 32 individuals (21 with detailed clinical information).
 Presentations were of microcephaly, simplified gyral pattern of the
@@
 dysmorphic facial features, seizure, vocal cord paralysis and hearing
-impairment. \citet{ravenscroft2021neurogenetic} studied a family from
-Melbourne exhibiting features involving arthrogryposis multiplex
-congenita, complex brain malformations, small for gestation age and
-hypoplasia of the corpus callosum. In two of the three related cases
-were additional features of microcephaly, congenital encephalopathy,
-cerebellar malformation and hypoplasia and hypomyelination. A case study
-from China is described in \citet{ji2022case} involving a girl
-presenting in infancy with intrauterine growth restriction,
-microcephaly, postnatal developmental delay, arthrogryposis,
-hypertonicity, seizure, and hypomyelination on brain magnetic resonance
-imaging. The authors report parallels with two cases showing the same
-homozygous null variant. An individual reported in
-\citet{monies2019lessons} presented with distinct symptoms of brain
-atrophy and skeletal dysplasia whereas the case in
-\citet{magini2019loss} with the same variant exhibited more typical
-clinical features. A recent study by \citet{bijarnia2022growth} presents
-the case of a 22-month old girl presenting with the typical phenotype of
-neurodevelopmental delay, prenatal onset growth failure, arthrogryposis,
-microcephaly and brain anomalies including severe hypomyelination,
-simplified gyral pattern and hypoplasia of corpus callosum and
-brainstem. Notably, there was also additional non-typical clinical
-findings of nystagmus and visual impairment secondary to macular
-dystrophy and retinal pigment epithelial stippling at posterior pole.
+impairment. Ravenscroft et al. (2021) studied a family from Melbourne
+exhibiting features involving arthrogryposis multiplex congenita,
+complex brain malformations, small for gestation age and hypoplasia of
+the corpus callosum. In two of the three related cases were additional
+features of microcephaly, congenital encephalopathy, cerebellar
+malformation and hypoplasia and hypomyelination. A case study from China
+is described in Ji et al. (2022) involving a girl presenting in infancy
+with intrauterine growth restriction, microcephaly, postnatal
+developmental delay, arthrogryposis, hypertonicity, seizure, and
+hypomyelination on brain magnetic resonance imaging. The authors report
+parallels with two cases showing the same homozygous null variant. An
+individual reported in Monies et al. (2019) presented with distinct
+symptoms of brain atrophy and skeletal dysplasia whereas the case in
+Magini et al. (2019) with the same variant exhibited more typical
+clinical features. A recent study by Bijarnia-Mahay et al. (2022)
+presents the case of a 22-month old girl presenting with the typical
+phenotype of neurodevelopmental delay, prenatal onset growth failure,
+arthrogryposis, microcephaly and brain anomalies including severe
+hypomyelination, simplified gyral pattern and hypoplasia of corpus
+callosum and brainstem. Notably, there was also additional non-typical
+clinical findings of nystagmus and visual impairment secondary to
+macular dystrophy and retinal pigment epithelial stippling at posterior
+pole.
 
 Further work to collate and analyse data relating to NEDMABA is
-challenging. While \citet{magini2019loss} included a supplementary
+challenging. While Magini et al. (2019) included a supplementary
 clinical phenotype data set in their study, these data are not suitable
@@
 applied between cases, making machine interpretation of the data
-difficult. Other studies \citep{ji2022case, bijarnia2022growth} present
-only written case reports with some tabulated summaries. While larger
-studies \citep{ravenscroft2021neurogenetic, monies2019lessons} focus
-more on genetic data, with less detail included on the clinical
-presentation of individuals. To overcome these data challenges,
-\citet{JSSv059i10} proposes a data structure know as `tidy' data which
-organises each observation in a row, with each feature as its own column
-and each value in just one cell. In this case detailed text-based data
-can be transformed to high dimensional binary indicators of key clinical
+difficult. Other studies (Ji et al. 2022; Bijarnia-Mahay et al. 2022)
+present only written case reports with some tabulated summaries. While
+larger studies (Ravenscroft et al. 2021; Monies et al. 2019) focus more
+on genetic data, with less detail included on the clinical presentation
+of individuals. To overcome these data challenges, Wickham (2014)
+proposes a data structure know as `tidy' data which organises each
+observation in a row, with each feature as its own column and each value
+in just one cell. In this case detailed text-based data can be
+transformed to high dimensional binary indicators of key clinical
 features. This structure is conducive to effective statistical analysis
-and integrates deliberately with the \emph{tidyverse} \citep{tidyverse}
-collection of packages for data analysis in R \citep{rbase}.
+and integrates deliberately with the \emph{tidyverse} (Wickham et al.
+2019) collection of packages for data analysis in R (R Core Team 2022).
 
 Further analysis to better describe rare genetic conditions was explored
-in \citet{diaz2020phenotype} in the context of large scale
+in Dı́az-Santiago et al. (2020) in the context of large scale
 genotype-phenotype analysis. The authors argue patients with rare
@@
 representation of the multi-dimensional space. Methods, such as Multiple
-Correspondence Analysis \citep{le2010multiple} are appropriate for high
-dimensional data comprised of binary indicators. This technique is cited
-in many studies in the application of dimension reduction and clustering
-of comorbidities and phenotypes from disease
-\citep{han2018cluster, costa2013use}.
+Correspondence Analysis (Le Roux and Rouanet 2010) are appropriate for
+high dimensional data comprised of binary indicators. This technique is
+cited in many studies in the application of dimension reduction and
+clustering of comorbidities and phenotypes from disease (Han, Benseler,
+and Tyrrell 2018; Costa et al. 2013).
 
 In terms of patient survival, methods in statistical survival analysis
-are well suited and commonly used in the field of genomics
-\citep{chen2014survival}. These methods in particular provide a
-mechanism for dealing with right-censoring, a common feature of clinical
-studies where the clinical outcome of interest may not be known by the
-end of the study period.
+are well suited and commonly used in the field of genomics (Chen, Sun,
+and Hoshida 2014). These methods in particular provide a mechanism for
+dealing with right-censoring, a common feature of clinical studies where
+the clinical outcome of interest may not be known by the end of the
+study period.
 
@@
 between some missense variants and a milder presentation with longer
-survival \citep{magini2019loss}, this hypothesis has not been analysed
+survival (Magini et al. 2019), this hypothesis has not been analysed
 further.
@@
 
-The data contained in \citet{magini2019loss} ``Summary of SMPD4-Related
+The data contained in Magini et al. (2019) ``Summary of SMPD4-Related
 Clinical Phenotype'' is an excel spreadsheet tabulating each of the 21
@@
 
-The data were read into \texttt{R} \citep{rbase} without any changes, so
-as to preserve the reproducibility of the data transformation steps. The
-data were transposed to ensure it could be presented as \emph{tidy}
-formatted data \citep{JSSv059i10}. This requires that each variable
-(clinical phenotype) forms a column; each observation (individual) forms
-a row and each type of observational unit forms a table (every cell has
-just one item). This resulted in a long-formatted data set of 21 rows
-(individuals) and 64 columns (clinical features). In many cases, key
-clinical information was entered as free-text descriptions, which
-rendered any attempt of meaningful analysis impractical
-(Table~\ref{tbl-nontidy}). In these cases, the text was tokenised by
-separating the list of clinical observations at each comma and forming a
-binary indicator column noting its presence `1' or absence `0'
-(Table~\ref{tbl-tidy}).
+The data were read into \texttt{R} (R Core Team 2022) without any
+changes, so as to preserve the reproducibility of the data
+transformation steps. The data were transposed to ensure it could be
+presented as \emph{tidy} formatted data (Wickham 2014). This requires
+that each variable (clinical phenotype) forms a column; each observation
+(individual) forms a row and each type of observational unit forms a
+table (every cell has just one item). This resulted in a long-formatted
+data set of 21 rows (individuals) and 64 columns (clinical features). In
+many cases, key clinical information was entered as free-text
+descriptions, which rendered any attempt of meaningful analysis
+impractical (Table~\ref{tbl-nontidy}). In these cases, the text was
+tokenised by separating the list of clinical observations at each comma
+and forming a binary indicator column noting its presence `1' or absence
+`0' (Table~\ref{tbl-tidy}).
 
 \hypertarget{tbl-nontidy}{}
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2083}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5139}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2778}}@{}}
-\caption{\label{tbl-nontidy}Example of non-tidy free-text descriptions
-of features}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Family 1- Individual 1
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Family 1- Individual 4
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Family 1- Individual 1
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Family 1- Individual 4
-\end{minipage} \\
-\midrule()
-\endhead
-Facial dysmorphisms & short palpebral fissures, large ears, simple
-helices, smooth philtrum, thin lips, bilateral simian creases & short
-palpebral fissures, receding forehead, thin upper lip \\
-\bottomrule()
+\begin{longtable}{ll}
+\caption{\label{tbl-nontidy}Example of non-tidy free-text descriptions of features. }\tabularnewline
+
+\toprule
+Family 1- Individual 1 & Family 1- Individual 4  \\ 
+\midrule
+short palpebral fissures, large ears, simple helices & short palpebral fissures, receding forehead \\ 
+\bottomrule
 \end{longtable}
@@
 \hypertarget{tbl-tidy}{}
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1781}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}@{}}
-\caption{\label{tbl-tidy}Example of tidy formatted data where
-individuals are transposed into rows and text into binary
-indicators}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-id
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-short palpebral fissures
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-large ears
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-simple helices
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-receding forehead
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-\ldots{}
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-id
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-short palpebral fissures
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-large ears
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-simple helices
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-receding forehead
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-\ldots{}
-\end{minipage} \\
-\midrule()
-\endhead
-Family 1- Individual 1 & 1 & 1 & 1 & 0 & \ldots{} \\
-Family 1- Individual 4 & 1 & 0 & 0 & 1 & \ldots{} \\
-\bottomrule()
+\begin{longtable}{lrrrr}
+\caption{\label{tbl-tidy}Example of tidy formatted data where individuals are transposed into
+rows and text into binary indicators }\tabularnewline
+
+\toprule
+id & short palpebral fissures & large ears & simple helices & receding forehead \\ 
+\midrule
+Family 1- Individual 1 & 1 & 1 & 1 & 0 \\ 
+Family 1- Individual 4 & 1 & 0 & 0 & 1 \\ 
+\bottomrule
 \end{longtable}
@@
 This format was used as a template to align other written case reports
-identified in the literature
-\citep{ravenscroft2021neurogenetic, monies2019lessons, bijarnia2022growth, ji2022case}.
-These were manually entered to conform to this template to allow the
-data to be combined for further analysis.
+identified in the literature (Ravenscroft et al. 2021; Monies et al.
+2019; Bijarnia-Mahay et al. 2022; Ji et al. 2022). These were manually
+entered to conform to this template to allow the data to be combined for
+further analysis.
 
@@
 available in Table~\ref{tbl-dict}. These data sets were compiled into a
-publicly available \texttt{R} Package called \texttt{SMPD4}
-\citep{smpd4-data} to allow for reproducibility and sharing.
+publicly available \texttt{R} Package called \texttt{SMPD4} (Marchiori
+2022) to allow for reproducibility and sharing.
 
@@
 Correspondence Analysis (MCA) (an analogy to Principle Component
-Analysis (PCA) for categorical data) \citep{le2010multiple} in order to
-arrive at a set of meaningfully small dimensions that account for most
-of the variation in the data. This computation was carried out using the
-\emph{FactoMineR} R package \citep{factominer}.
+Analysis (PCA) for categorical data) (Le Roux and Rouanet 2010) in order
+to arrive at a set of meaningfully small dimensions that account for
+most of the variation in the data. This computation was carried out
+using the \emph{FactoMineR} R package (Lê, Josse, and Husson 2008).
 
@@
 The \emph{contribution} of each category was computed for each of the
-principal axis per \citet{le2010multiple}. This metric identifies the
-proportion of variance of that axis due to the point and is defined as:
+principal axis per Le Roux and Rouanet (2010). This metric identifies
+the proportion of variance of that axis due to the point and is defined
+as:
 
@@
 homogeneous and externally heterogeneous. This was computed in
-\texttt{R} using the \texttt{clValid} package \citep{clvalid}.
+\texttt{R} using the \texttt{clValid} package (Brock et al. 2008).
 
 A range of clustering algorithms were compared including AGNES
-hierarchical agglomerative \citep{kaufman2009finding}, k-means
-\citep{hartigan1979algorithm} and self-organising maps (SOM)
-\citep{kohonen2012self}. A sensible range of clusters from 2-5 were
-trialed for each clustering algorithm. The results for each method were
-evaluated using three measures of stability: Average proportion of
-non-overlap (APN), Average Distance (AD) and Average Distance Between
-Means (ADM) \citep{datta2003comparisons}. Stability measures compare the
-full clustering result with a result based on dropping each of the
-columns one at a time. These metrics were selected to represent each
-data point (clinical feature) with a number of MCA dimensions. A good
-cluster solution is one that is robust and meaningful across all of
-these dimensions (Figure~\ref{fig-clvalid}). The optimal method and
-cluster number was selected based on a trade off between the best
-performing method using the evaluation metrics and a sensible
-partitioning of the data for this analysis (Table~\ref{tbl-eig}).
+hierarchical agglomerative (Kaufman and Rousseeuw 2009), k-means
+(Hartigan and Wong 1979) and self-organising maps (SOM) (Kohonen 2012).
+A sensible range of clusters from 2-5 were trialed for each clustering
+algorithm. The results for each method were evaluated using three
+measures of stability: Average proportion of non-overlap (APN), Average
+Distance (AD) and Average Distance Between Means (ADM) (Datta and Datta
+2003). Stability measures compare the full clustering result with a
+result based on dropping each of the columns one at a time. These
+metrics were selected to represent each data point (clinical feature)
+with a number of MCA dimensions. A good cluster solution is one that is
+robust and meaningful across all of these dimensions
+(Figure~\ref{fig-clvalid}). The optimal method and cluster number was
+selected based on a trade off between the best performing method using
+the evaluation metrics and a sensible partitioning of the data for this
+analysis (Table~\ref{tbl-eig}).
 
@@
 To model survival probability for all subjects in the combined data, the
-data were subsetted to include \emph{survival\_time} which is either the
-number of days the individual survived for, or the age in days at last
-follow up, and \emph{deceased} a numeric event indicator which is equal
-to 1 if the subject is deceased and 0 if the subject was still alive at
-the time the respective study had concluded. The Kaplan-Meier estimator
-\citep{kaplan1958nonparametric} was used to estimate the survival
+data were subsetted to include \texttt{survival\_time} which is either
+the number of days the individual survived for, or the age in days at
+last follow up, and \texttt{deceased} a numeric event indicator which is
+equal to 1 if the subject is deceased and 0 if the subject was still
+alive at the time the respective study had concluded. The Kaplan-Meier
+estimator (Kaplan and Meier 1958) was used to estimate the survival
 function of the data.
@@
 The baseline estimator of all individuals was calculated using the
-\emph{survival} package in \texttt{R} \citep{survival-book}. Next the
-Kaplan-Meier estimator was stratified by the type of genetic variation
-reported in the literature. A log-rank test was performed to detect
-differences in the survival curves using methods from
-\citet{harrington1982class}, again implemented in the \emph{survival}
-\texttt{R} package.
+\texttt{survival} package in \texttt{R} (Terry M. Therneau and Patricia
+M. Grambsch 2000). Next the Kaplan-Meier estimator was stratified by the
+type of genetic variation reported in the literature. A log-rank test
+was performed to detect differences in the survival curves using methods
+from Harrington and Fleming (1982), again implemented in the
+\texttt{survival} \texttt{R} package.
 
@@
 (Figure~\ref{fig-biplotprelim}). These individuals represented two twins
-from \citet{magini2019loss} who were described with a much milder
+from Magini et al. (2019) who were described with a much milder
 phenotype and were subject to a missense variation in \emph{SMPD4}.
@@
 
-\caption{\label{fig-ctr}Top 10 feature categories from MCA analysis for
-the first four MCA dimensions. The red dashed line indicates the mean
-contribution value. Each feature may be present or absent as indicated
-by 1 or 0 respectively}
+\caption{\label{fig-ctr}Top 5 feature categories as measured by feature
+contribution from MCA analysis for the first four MCA dimensions. The
+red dashed line indicates the mean contribution value.}
 
@@
 probability after one year of 0.403 (95\% CI: 0.252, 0.644) and survival
-probability at close to ten years of 0.24 (95\% CI: 0.11, 0.52).
-
-\begin{longtable}[]{@{}rrrrrrrr@{}}
-\caption{Survival probabilities derived from Kaplan-Meier estimation for
-the range of recorded survival times, including 95\% Confidence
-Interval.}\tabularnewline
-\toprule()
-time & n.risk & n.event & n.censor & surv & std.err & upper & lower \\
-\midrule()
-\endfirsthead
-\toprule()
-time & n.risk & n.event & n.censor & surv & std.err & upper & lower \\
-\midrule()
-\endhead
-0.0 & 28 & 5 & 0 & 0.82 & 0.09 & 0.98 & 0.69 \\
-15.0 & 23 & 1 & 0 & 0.79 & 0.10 & 0.95 & 0.65 \\
-30.0 & 22 & 2 & 0 & 0.71 & 0.12 & 0.90 & 0.57 \\
-47.0 & 20 & 1 & 0 & 0.68 & 0.13 & 0.88 & 0.53 \\
-60.0 & 19 & 1 & 1 & 0.64 & 0.14 & 0.85 & 0.49 \\
-67.0 & 17 & 1 & 0 & 0.61 & 0.15 & 0.82 & 0.45 \\
-90.0 & 16 & 0 & 1 & 0.61 & 0.15 & 0.82 & 0.45 \\
-120.0 & 15 & 2 & 0 & 0.52 & 0.18 & 0.75 & 0.37 \\
-150.0 & 13 & 1 & 0 & 0.48 & 0.20 & 0.72 & 0.33 \\
-180.0 & 12 & 1 & 0 & 0.44 & 0.22 & 0.68 & 0.29 \\
-240.0 & 11 & 1 & 0 & 0.40 & 0.24 & 0.64 & 0.25 \\
-270.0 & 10 & 0 & 1 & 0.40 & 0.24 & 0.64 & 0.25 \\
-365.0 & 9 & 0 & 1 & 0.40 & 0.24 & 0.64 & 0.25 \\
-750.0 & 8 & 1 & 0 & 0.35 & 0.27 & 0.60 & 0.21 \\
-1095.0 & 7 & 1 & 0 & 0.30 & 0.31 & 0.56 & 0.16 \\
-1825.0 & 6 & 0 & 1 & 0.30 & 0.31 & 0.56 & 0.16 \\
-2190.0 & 5 & 1 & 0 & 0.24 & 0.39 & 0.52 & 0.11 \\
-3285.0 & 4 & 0 & 2 & 0.24 & 0.39 & 0.52 & 0.11 \\
-4380.0 & 2 & 1 & 0 & 0.12 & 0.81 & 0.59 & 0.02 \\
-4562.5 & 1 & 1 & 0 & 0.00 & Inf & NA & NA \\
-\bottomrule()
-\end{longtable}
+probability at close to ten years of 0.24 (95\% CI: 0.11, 0.52) See
+Table~\ref{tbl-km}.
 
@@
 
-A comparative box plot of survival times by variant type shows
-significant outliers for compound heterozygous \emph{missense + splice}
-and homozygous \emph{missense} variants. However, even within homozygous
-missense variants the survival time varies greatly. Other missense
-variants, such as heterozygous missense with gene deletion do not show
-the same increased survival time as noted by \citet{magini2019loss}.
+A comparative box plot of survival times (Figure~\ref{fig-box}) by
+variant type shows significant outliers for compound heterozygous
+\emph{missense + splice} and homozygous \emph{missense} variants.
+However, even within homozygous missense variants the survival time
+varies greatly. Other missense variants, such as heterozygous missense
+with gene deletion do not show the same increased survival time as noted
+by Magini et al. (2019).
 
@@
 
-This meta-review of current research was able to consolidate data that
+This meta-analysis of current research was able to consolidate data that
 allowed statistical analysis to be conducted on affected individuals.
@@
 related features that are not well described in The Human Phenotype
-Ontology \citep{kohler2021human} as shown in Table~\ref{tbl-hpo}. It
-shows the prolific feature of `respiratory distress' as more commonly
+Ontology (Köhler et al. 2021) as shown in Table~\ref{tbl-hpo}. It shows
+the prolific feature of `respiratory distress' as more commonly
 associated with these adjacent features such as vocal cord palsy,
@@
 \bottomrule()
+\end{longtable}
+
+\hypertarget{tbl-km}{}
+\begin{longtable}{rrrrrrrr}
+\caption{\label{tbl-km}Survival probabilities derived from Kaplan-Meier estimation for the
+range of recorded survival times, including 95\% Confidence Interval. }\tabularnewline
+
+\toprule
+time & n.risk & n.event & n.censor & surv & std.err & upper & lower \\ 
+\midrule
+0.0 & 28 & 5 & 0 & $82.14\%$ & $0.09$ & $97.63\%$ & $69.11\%$ \\ 
+15.0 & 23 & 1 & 0 & $78.57\%$ & $0.10$ & $95.34\%$ & $64.75\%$ \\ 
+30.0 & 22 & 2 & 0 & $71.43\%$ & $0.12$ & $90.28\%$ & $56.51\%$ \\ 
+47.0 & 20 & 1 & 0 & $67.86\%$ & $0.13$ & $87.56\%$ & $52.59\%$ \\ 
+60.0 & 19 & 1 & 1 & $64.29\%$ & $0.14$ & $84.73\%$ & $48.78\%$ \\ 
+67.0 & 17 & 1 & 0 & $60.50\%$ & $0.15$ & $81.72\%$ & $44.80\%$ \\ 
+90.0 & 16 & 0 & 1 & $60.50\%$ & $0.15$ & $81.72\%$ & $44.80\%$ \\ 
+120.0 & 15 & 2 & 0 & $52.44\%$ & $0.18$ & $75.17\%$ & $36.58\%$ \\ 
+150.0 & 13 & 1 & 0 & $48.40\%$ & $0.20$ & $71.70\%$ & $32.68\%$ \\ 
+180.0 & 12 & 1 & 0 & $44.37\%$ & $0.22$ & $68.09\%$ & $28.91\%$ \\ 
+240.0 & 11 & 1 & 0 & $40.34\%$ & $0.24$ & $64.37\%$ & $25.28\%$ \\ 
+270.0 & 10 & 0 & 1 & $40.34\%$ & $0.24$ & $64.37\%$ & $25.28\%$ \\ 
+365.0 & 9 & 0 & 1 & $40.34\%$ & $0.24$ & $64.37\%$ & $25.28\%$ \\ 
+750.0 & 8 & 1 & 0 & $35.29\%$ & $0.27$ & $60.31\%$ & $20.66\%$ \\ 
+1095.0 & 7 & 1 & 0 & $30.25\%$ & $0.31$ & $55.97\%$ & $16.35\%$ \\ 
+1825.0 & 6 & 0 & 1 & $30.25\%$ & $0.31$ & $55.97\%$ & $16.35\%$ \\ 
+2190.0 & 5 & 1 & 0 & $24.20\%$ & $0.39$ & $51.51\%$ & $11.37\%$ \\ 
+3285.0 & 4 & 0 & 2 & $24.20\%$ & $0.39$ & $51.51\%$ & $11.37\%$ \\ 
+4380.0 & 2 & 1 & 0 & $12.10\%$ & $0.81$ & $58.65\%$ & $2.50\%$ \\ 
+4562.5 & 1 & 1 & 0 & $0.00\%$ & $Inf$ & NA & NA \\ 
+\bottomrule
 \end{longtable}
@@
 
-\renewcommand\refname{References}
-  \bibliography{bibliography.bib}
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-bijarnia2022growth}{}}%
+Bijarnia-Mahay, Sunita, Puneeth H Somashekar, Parneet Kaur, Samarth
+Kulshrestha, Vedam L Ramprasad, Sakthivel Murugan, Seema Sud, and Anju
+Shukla. 2022. {``Growth and Neurodevelopmental Disorder with
+Arthrogryposis, Microcephaly and Structural Brain Anomalies Caused by
+Bi-Allelic Partial Deletion of Smpd4 Gene.''} \emph{Journal of Human
+Genetics} 67 (3): 133--36.
+
+\leavevmode\vadjust pre{\hypertarget{ref-clvalid}{}}%
+Brock, Guy, Vasyl Pihur, Susmita Datta, and Somnath Datta. 2008.
+{``{clValid}: An {R} Package for Cluster Validation.''} \emph{Journal of
+Statistical Software} 25 (4): 1--22.
+\url{https://www.jstatsoft.org/v25/i04/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-chen2014survival}{}}%
+Chen, Xintong, Xiaochen Sun, and Yujin Hoshida. 2014. {``Survival
+Analysis Tools in Genomics Research.''} \emph{Human Genomics} 8 (1):
+1--5.
+
+\leavevmode\vadjust pre{\hypertarget{ref-costa2013use}{}}%
+Costa, Patrı́cio Soares, Nadine Correia Santos, Pedro Cunha, Jorge
+Cotter, and Nuno Sousa. 2013. {``The Use of Multiple Correspondence
+Analysis to Explore Associations Between Categories of Qualitative
+Variables in Healthy Ageing.''} \emph{Journal of Aging Research} 2013.
+
+\leavevmode\vadjust pre{\hypertarget{ref-datta2003comparisons}{}}%
+Datta, Susmita, and Somnath Datta. 2003. {``Comparisons and Validation
+of Statistical Clustering Techniques for Microarray Gene Expression
+Data.''} \emph{Bioinformatics} 19 (4): 459--66.
+
+\leavevmode\vadjust pre{\hypertarget{ref-diaz2020phenotype}{}}%
+Dı́az-Santiago, Elena, Fernando M Jabato, Elena Rojano, Pedro Seoane,
+Florencio Pazos, James R Perkins, and Juan AG Ranea. 2020.
+{``Phenotype-Genotype Comorbidity Analysis of Patients with Rare
+Disorders Provides Insight into Their Pathological and Molecular
+Bases.''} \emph{PLoS Genetics} 16 (10): e1009054.
+
+\leavevmode\vadjust pre{\hypertarget{ref-han2018cluster}{}}%
+Han, Lu, Susanne M Benseler, and Pascal N Tyrrell. 2018. {``Cluster and
+Multiple Correspondence Analyses in Rheumatology: Paths to Uncovering
+Relationships in a Sea of Data.''} \emph{Rheumatic Disease Clinics} 44
+(2): 349--60.
+
+\leavevmode\vadjust pre{\hypertarget{ref-harrington1982class}{}}%
+Harrington, David P, and Thomas R Fleming. 1982. {``A Class of Rank Test
+Procedures for Censored Survival Data.''} \emph{Biometrika} 69 (3):
+553--66.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hartigan1979algorithm}{}}%
+Hartigan, John A, and Manchek A Wong. 1979. {``Algorithm AS 136: A
+k-Means Clustering Algorithm.''} \emph{Journal of the Royal Statistical
+Society. Series c (Applied Statistics)} 28 (1): 100--108.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ji2022case}{}}%
+Ji, Weigang, Xiangtian Kong, Honggang Yin, Jian Xu, and Xueqian Wang.
+2022. {``Case Report: Novel Biallelic Null Variants of Smpd4 Confirm Its
+Involvement in Neurodevelopmental Disorder with Microcephaly,
+Arthrogryposis, and Structural Brain Anomalies.''} \emph{Frontiers in
+Genetics} 13.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kaplan1958nonparametric}{}}%
+Kaplan, Edward L, and Paul Meier. 1958. {``Nonparametric Estimation from
+Incomplete Observations.''} \emph{Journal of the American Statistical
+Association} 53 (282): 457--81.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kaufman2009finding}{}}%
+Kaufman, Leonard, and Peter J Rousseeuw. 2009. \emph{Finding Groups in
+Data: An Introduction to Cluster Analysis}. John Wiley \& Sons.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kohler2021human}{}}%
+Köhler, Sebastian, Michael Gargano, Nicolas Matentzoglu, Leigh C
+Carmody, David Lewis-Smith, Nicole A Vasilevsky, Daniel Danis, et al.
+2021. {``The Human Phenotype Ontology in 2021.''} \emph{Nucleic Acids
+Research} 49 (D1): D1207--17.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kohonen2012self}{}}%
+Kohonen, Teuvo. 2012. \emph{Self-Organizing Maps}. Vol. 30. Springer
+Science \& Business Media.
+
+\leavevmode\vadjust pre{\hypertarget{ref-le2010multiple}{}}%
+Le Roux, Brigitte, and Henry Rouanet. 2010. \emph{Multiple
+Correspondence Analysis}. Vol. 163. Sage.
+
+\leavevmode\vadjust pre{\hypertarget{ref-factominer}{}}%
+Lê, Sébastien, Julie Josse, and François Husson. 2008. {``{FactoMineR}:
+A Package for Multivariate Analysis.''} \emph{Journal of Statistical
+Software} 25 (1): 1--18. \url{https://doi.org/10.18637/jss.v025.i01}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-magini2019loss}{}}%
+Magini, Pamela, Daphne J Smits, Laura Vandervore, Rachel Schot, Marta
+Columbaro, Esmee Kasteleijn, Mees van der Ent, et al. 2019. {``Loss of
+Smpd4 Causes a Developmental Disorder Characterized by Microcephaly and
+Congenital Arthrogryposis.''} \emph{The American Journal of Human
+Genetics} 105 (4): 689--705.
+
+\leavevmode\vadjust pre{\hypertarget{ref-smpd4-data}{}}%
+Marchiori, Dean. 2022. {``Deanmarchiori/Smpd4: Release V2.1.0.''}
+Zenodo. \url{https://doi.org/10.5281/zenodo.7071092}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-monies2019lessons}{}}%
+Monies, Dorota, Mohammed Abouelhoda, Mirna Assoum, Nabil Moghrabi,
+Rafiullah Rafiullah, Naif Almontashiri, Mohammed Alowain, et al. 2019.
+{``Lessons Learned from Large-Scale, First-Tier Clinical Exome
+Sequencing in a Highly Consanguineous Population.''} \emph{The American
+Journal of Human Genetics} 104 (6): 1182--1201.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rbase}{}}%
+R Core Team. 2022. \emph{R: A Language and Environment for Statistical
+Computing}. Vienna, Austria: R Foundation for Statistical Computing.
+\url{https://www.R-project.org/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ravenscroft2021neurogenetic}{}}%
+Ravenscroft, Gina, Joshua S Clayton, Fathimath Faiz, Padma Sivadorai, Di
+Milnes, Rob Cincotta, Phillip Moon, et al. 2021. {``Neurogenetic Fetal
+Akinesia and Arthrogryposis: Genetics, Expanding Genotype-Phenotypes and
+Functional Genomics.''} \emph{Journal of Medical Genetics} 58 (9):
+609--18.
+
+\leavevmode\vadjust pre{\hypertarget{ref-survival-book}{}}%
+Terry M. Therneau, and Patricia M. Grambsch. 2000. \emph{Modeling
+Survival Data: Extending the {C}ox Model}. New York: Springer.
+
+\leavevmode\vadjust pre{\hypertarget{ref-JSSv059i10}{}}%
+Wickham, Hadley. 2014. {``Tidy Data.''} \emph{Journal of Statistical
+Software} 59 (10): 1--23. \url{https://doi.org/10.18637/jss.v059.i10}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
+Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
+D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019.
+{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software}
+4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.
+
+\end{CSLReferences}
 

### 2022-09-15 paper/paper.qmd

@@
 ---
-title: Clinical Phenotype and Survival Analysis of Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in *SMPD4*
-author:
-  - name: Dean Marchiori
-    email: deanmarchiori@gmail.com
-    attributes:
-        corresponding: true
-abstract: |
+title: "Meta-Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD"
+author: Dean Marchiori
+format:
+  pdf:
+    keep-tex: true
+editor: visual
+bibliography: bibliography.bib
+date: last-modified
+execute: 
+  echo: false 
+---
+
+```{r echo=FALSE, message=FALSE}
+library(targets)
+source(here::here('_targets_packages.R'))
+source(here::here('R', 'utils.R'))
+theme_set(theme_bw())
+```
+
+# Abstract
+
 A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*.
@@
 This analysis can provide clinicians and genetic counsellors with better information to aide in decision making and support for families with this rare condition.
-keywords: 
-  - NEDMABA
-  - SMPD4  
-#journal: "Preprint"    
-date: last-modified
-execute: 
-  echo: false 
-bibliography: bibliography.bib
-format:
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      formatting: preprint
-      model: 3p
-      cite-style: authoryear
----
-
-```{r echo=FALSE, message=FALSE}
-library(targets)
-source(here::here('_targets_packages.R'))
-source(here::here('R', 'utils.R'))
-theme_set(theme_bw())
-```
 
@@
 
-|                     | Family 1- Individual 1                                                                                     | Family 1- Individual 4                                      |
-|---------------|-------------------------------------|--------------------|
-| Facial dysmorphisms | short palpebral fissures, large ears, simple helices, smooth philtrum, thin lips, bilateral simian creases | short palpebral fissures, receding forehead, thin upper lip |
+```{r}
+#| label: tbl-nontidy
+#| tbl-cap: Example of non-tidy free-text descriptions of features.
+tibble(
+  `Family 1- Individual 1` = c("short palpebral fissures, large ears, simple helices"),
+  `Family 1- Individual 4 ` = c("short palpebral fissures, receding forehead")
+  ) |> 
+  gt()
+```
 
-: Example of non-tidy free-text descriptions of features {#tbl-nontidy}
-
-| id                     | short palpebral fissures | large ears | simple helices | receding forehead | ... |
-|------------|-------------|------------|------------|------------|------------|
-| Family 1- Individual 1 | 1                        | 1          | 1              | 0                 | ... |
-| Family 1- Individual 4 | 1                        | 0          | 0              | 1                 | ... |
-
-: Example of tidy formatted data where individuals are transposed into rows and text into binary indicators {#tbl-tidy}
+```{r}
+#| label: tbl-tidy
+#| tbl-cap: Example of tidy formatted data where individuals are transposed into rows and text into binary indicators
+tribble(
+  ~"id", ~"short palpebral fissures", ~"large ears", ~"simple helices", ~"receding forehead", 
+  "Family 1- Individual 1", 1, 1, 1, 0,
+  "Family 1- Individual 4", 1, 0, 0, 1, 
+  ) |> 
+  gt()
+```
 
@@
 
-To model survival probability for all subjects in the combined data, the data were subsetted to include *survival_time* which is either the number of days the individual survived for, or the age in days at last follow up, and *deceased* a numeric event indicator which is equal to 1 if the subject is deceased and 0 if the subject was still alive at the time the respective study had concluded.
+To model survival probability for all subjects in the combined data, the data were subsetted to include `survival_time` which is either the number of days the individual survived for, or the age in days at last follow up, and `deceased` a numeric event indicator which is equal to 1 if the subject is deceased and 0 if the subject was still alive at the time the respective study had concluded.
 The Kaplan-Meier estimator [@kaplan1958nonparametric] was used to estimate the survival function of the data.
@@
 
-The baseline estimator of all individuals was calculated using the *survival* package in `R` [@survival-book].
+The baseline estimator of all individuals was calculated using the `survival` package in `R` [@survival-book].
 Next the Kaplan-Meier estimator was stratified by the type of genetic variation reported in the literature.
-A log-rank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* `R` package.
+A log-rank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the `survival` `R` package.
 
@@
 
-```{r fig.height=5, fig.width=5,  warning=FALSE}
+```{r  warning=FALSE}
 #| label: fig-biplotprelim
@@
 
-```{r ctr-plot, fig.height=6, fig.width = 8, warning=FALSE}
+```{r ctr-plot, warning=FALSE}
 #| label: fig-ctr
-#| fig-cap: "Top 10 feature categories from MCA analysis for the first four MCA dimensions.
-The red dashed line indicates the mean contribution value.
-Each feature may be present or absent as indicated by 1 or 0 respectively"
+#| fig-cap: "Top 5 feature categories as measured by feature contribution from MCA analysis for the first four MCA dimensions. The red dashed line indicates the mean contribution value."
 mca$var$contrib |> 
@@
   group_by(dim) |> 
-  top_n(10, wt = contribution) |> 
+  slice_max(n = 5, order_by = contribution, with_ties = F) |> 
   mutate(vars = fct_reorder(vars, contribution)) |> 
@@
   mutate(vars = str_replace_all(vars, "_", " ")) |> 
-  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim), fill = col)) +
+  mutate(vars = str_wrap(vars, width = 25)) |> 
+  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim))) +
   geom_col() +
@@
   geom_vline(xintercept = 100/122, lty = 2, col = 'red') +
-  guides(fill=guide_legend(title="Condition")) +
-  theme(legend.position = 'bottom') +
-  labs(y = '')
+  theme(legend.position = 'bottom', axis.text = element_text(size = 8)) +
+  labs(y = '', x = '')
 ```
 
-```{r fig.height=5, fig.width=5,  warning=FALSE}
+```{r , warning=FALSE}
 #| label: fig-biplot
@@
 
-```{r fig.height=5, fig.width=5,  warning=FALSE}
+```{r ,  warning=FALSE}
 #| label: fig-biplot2
@@
 
-```{r, fig.height=6, fig.width=6,  warning=FALSE}
+```{r,warning=FALSE}
 #| label: fig-clust
@@
   geom_hline(yintercept = 0, lty = 2, colour = "grey50") +
-  geom_label_repel(data = d |> group_by(clust) |> slice_sample(n=3), aes(label = var), size = 3, show.legend = F) +
+  geom_label_repel(data = d |> group_by(clust) |> slice_sample(n=3), aes(label = var), size = 2, show.legend = F) +
   geom_point(aes(colour = clust), position = "jitter", show.legend = F) +
@@
        fill = 'cluster') +
-    guides(color='none')
+    guides(color='none') +
+    theme(legend.position = 'none')
 ```
@@
 tar_load(kapm)
+
+x <- survminer::surv_summary(kapm_baseline, data = survival_data)
 ```
@@
 At time 0 the survival probability was 0.82 (95% CI: 0.69, 0.98).
-Survival probability after one year of 0.403 (95% CI: 0.252, 0.644) and survival probability at close to ten years of 0.24 (95% CI: 0.11, 0.52).
-
-```{r}
-# KM Curve
-x <- survminer::surv_summary(kapm_baseline, data = survival_data)
-kable(x, digits = 2, caption = "Survival probabilities derived from Kaplan-Meier estimation for the range of recorded survival times, including 95% Confidence Interval.")
-```
+Survival probability after one year of 0.403 (95% CI: 0.252, 0.644) and survival probability at close to ten years of 0.24 (95% CI: 0.11, 0.52) See @tbl-km.
 
@@
 
-A comparative box plot of survival times by variant type shows significant outliers for compound heterozygous *missense + splice* and homozygous *missense* variants.
+A comparative box plot of survival times (@fig-box) by variant type shows significant outliers for compound heterozygous *missense + splice* and homozygous *missense* variants.
 However, even within homozygous missense variants the survival time varies greatly.
@@
 
-```{r,fig.width=7, warning=FALSE}
+```{r, warning=FALSE}
 #| label: fig-box
@@
   theme(legend.position = 'bottom') +
+  scale_fill_grey(start = 0.4, end = 0.9) +
   labs(fill = 'Variant Type',
@@
 
-This meta-review of current research was able to consolidate data that allowed statistical analysis to be conducted on affected individuals.
+This meta-analysis of current research was able to consolidate data that allowed statistical analysis to be conducted on affected individuals.
 The use of dimension reduction techniques was shown to be successful in uncovering relevant and novel substructure in the data.
@@
 ```{r}
+#| label: tbl-km
+#| tbl-cap: "Survival probabilities derived from Kaplan-Meier estimation for the range of recorded survival times, including 95% Confidence Interval."
+x |> 
+  gt() |> 
+  gt::fmt_percent(columns = c(surv, upper, lower)) |> 
+  gt::fmt_number(columns = c(std.err), decimals = 2)
+```
+
+```{r}
 #| label: tbl-dict
@@
 
-# References {-}
+# References {.unnumbered}

### 2022-09-15 paper/paper.qmd
fixed typo

@@
 ---
-title: "Meta-Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD"
+title: "Meta-Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD4"
 author: Dean Marchiori

### 2022-09-15 paper/paper.tex
fixed typo

@@
 \hypersetup{
-  pdftitle={Meta-Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD},
+  pdftitle={Meta-Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD4},
   pdfauthor={Dean Marchiori},
@@
 Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants
-in SMPD}
+in SMPD4}
 \author{Dean Marchiori}
@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, boxrule=0pt, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, frame hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 
